### 2013-07-02 docs/paper/paper.tex

\documentclass{bioinfo}

% Bioinformatics definitions.

\copyrightyear{2013}
\pubyear{2013}

\begin{document}
\firstpage{1}

\title[Wormtable]{Wormtable: A data structure for genome scale tabular data}
\author[Kelleher \textit{et~al}]{Jerome Kelleher, Daniel L. Halligan 
    and Robert W. Ness }
\address{
University of Edinburgh,
King's Buildings,
West Mains Road,
EH9 3JT,
UK
}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\maketitle

\begin{abstract}
\section{Summary:}
Biological research generates vast quantities of tabular data, which must 
be processed to answer biological questions. These tables are produced by 
programs and exported to plain text files, which are often tens of gigabytes
long. Researchers must process these files 
line-by-line, parsing values encoded as plain text into native types so that 
calculations may be done. This is an enormously inefficient process.
Furthermore, there is no simple means of indexing these files so that 
arbitrary data values within rows can be quickly found. We introduce a new 
data format and software library called Wormtable, which provides efficient 
access to tabular data using languages such as Python. Wormtable avoids the 
overhead of repeatedly parsing text, random access to rows, 
and also provides efficient indexing of arbitrary columns within these tables.
\section{Availability:}
Package available at 
\href{http://pypi.python.org/pypi/wormtable}{http://pypi.python.org/pypi/wormtable}

\section{Contact:} \href{jerome.kelleher@ed.ac.uk}{jerome.kelleher@ed.ac.uk}
\end{abstract}

\section{Introduction}
The large volumes of data that present day bioinformatics generates 
is well documented. The improvement in sequencing technologies has
lead to an exponential increase in the volume of data that is available 
to biologists. Perfectly standard studies require the storage and processing 
of hundreds of gigabytes of data, and this volume of data is only set to 
increase as sequencing technologies continue to improve. [\emph{This 
intro is intended to set the stage: there is shit loads of data out there,
and this is only set to increase}].

Despite the increasing importance of data processing, however, the methods 
used in Bioinformatics are crude and inefficient being almost entirely 
based around the use of text files. Plain text files have many advantages:
they are the most portable format available, and can be processed on 
any platform with the minimum of library dependencies. Text files also 
have many disadvantages however; the principle difficulties for our 
purposes here are, they are slow to parse and difficult to index.
% text files are also easy to read/comprehend and don't require specialist
% knowledge to process.
% I think learning a database syntax is a barrier to many researchers that are
% not trained bioinformaticians

A popular file format for Bioinformatics use is the Variant Call Format~\citep{da+11},
or VCF. In VCF,  information about variant sites in a genome is encoded 
as tab-delimited rows in a text file. To access information about a 
particular site, one must proceed through the file line-by-line until the 
desired record is found. There are methods available to index this 
file~\citep{li11}, but these have no semantic 
understanding of the structure of the file and so are limited in the 
fields that can be indexed. One cannot, for example, build an index 
to quickly obtain all records in which the quality field is greater
than some value. Such queries can take a considerable time on large VCF 
files, which discourages testing and checking for the effects of 
arbitrary thresholds in analyses.

One may, of course, load VCF data into a relational database, and 
use SQL to retrieve the records of interest. This approach has several 
disadvantages, however. Firstly, one must first examine the VCF 
header and construct an appropriate SQL schema to hold the columns 
of interest. It is feasible to do this for more recent versions of VCF, 
but is certainly not a trivial process, particularly if the schema 
must be portable across different relational database platforms.
Once the schema has been constructed, the data must be inserted into 
this database line-by-line by constructing SQL code, which is a 
time-consuming and error prone process. Even assuming that the data 
has been inserted into a table with an appropriate schema, there 
are still difficulties to be addressed. Maintaining a database 
server is a difficult task and when the volumes of data become
large, a specialised administrator is usually required to 
ensure the server works correctly.

Much of the time, a database server is far more complicated than
we require for large scale biological data. In the case of VCF,
the file is written once by an application such as [gatk? references?]
%Also samtools
and is not subsequently expected to change. Thus, storing this information 
in a relational database with its sophisticated concurrency control
and guarantees of consistency in the face of asynchronous updates
is entirely unnecessary. Even the client-server architecture of 
a typical relational database is wasteful in this case: protocol 
commands must be read and parsed, and responses must be encoded,
written to a socket and subsequently converted back to native types,
all of which constitutes a significant overhead when streaming large 
quantities of data. The limitations of relational database servers 
have been acknowledged for some time now, and has led to a 
variety of `NoSQL' database systems~\citep{m12}.

Databases do, however, have significant advantages over 
plain text files. Data is typically encoded in an efficient binary 
format, saving substantial space and giving better precision when 
dealing with floating point formats. Records are stored in a
B-tree type structure, leading to fast access times even 
over very large datasets. Indexing over columns within rows is 
also very powerful, allowing us to quickly retrieve 
rows with particular features. Ideally, we would like a data format
that combines these useful features of databases without the 
drawbacks of a relational database server outlined above.

Wormtable 
%... [Write Once Read Only Table] ...
is a new data format and software library designed specifically
to deal with the challenges of data processing in bioinformatics.
It provides a portable, compact, read-only store for tabular
data with high-performance access to rows via indexes. Tables can
be either written directly via the software library, or converted
from existing formats via command line tools. Tables can be read 
concurrently by any number of processes, and are accessed by a 
simple API in Python.

% Also provided are scripts to import two common bioinformatic file standards - gff and vcf
% along with scripts to apply filters to these files?

\section{Wormtable}
To take advantage of the data processing capabilities of a database
without the overheads and disadvantages of a relational database 
server, Wormtable is built on top of Berkeley DB~\citep{obs99}.
Berkeley DB is an open source embedded database toolkit, that provides a
scalable key-value store. It is a mature and stable platform, 
and is currently the most widely deployed database toolkit in the 
world~\citep{sb12}.

Berkeley DB (or simply DB) 
differs from relational database servers in several ways. 
Firstly, it is a software library that a program links against, 
and so there is no need for a separate database server process.
Secondly, DB provides only a key-value store, and so has 
no knowledge of the structure of the records that it stores,
treating them simply as an array of bytes. While there 
are sophisticated 
mechanisms within DB for locking, transactions and so on, 
these are all optional, allowing a very lightweight 
read-only store to be built using the library.

This is the approach taken in Wormtable. We assume that tabular data 
is written once, by some program generating the data, and subsequently 
read many times. Under these assumptions, the primary database 
holding the rows can be opened in a read-only state, allowing any 
number of processes to read it concurrently in a lock free manner.

Since DB is agnostic to the structure of data stored in rows, we 
are free to design our own formats. In Wormtable, the row format is
a compact and flexible binary format holding integer, floating point 
or character values of various sizes. Each column in a row may contain 
either a fixed or variable number of elements (each element being of a
fixed type and size). This row format enables two important properties:
firstly, values from any column can be accessed in constant 
time; and secondly, values in columns can be converted directly into 
native machine values, ready for calculations.

Indexing of columns is achieved via DB also. An index is simply a
database mapping values in a column (or list of columns) to rows 
in the primary database. Indexes allow us to find rows with particular 
properties very quickly, and are a very useful method of excluding 
rows with certain properties from our analyses.

\section{Examples}
Wormtable is designed to make working with genome scale data more 
efficient, and in this section we describe some examples to 
illustrate the improvements over existing methods. To do this, 
we converted a VCF file with XXX rows from the YYY project 
into wormtable format. The (uncompressed) VCF file was XX
gigabytes, and the converted wormtable YY gigabytes, which required 
X hours.

Values are stored in wormtable in a binary format, so that no parsing is
required when reading in rows. Methods using the VCF directly must parse 
each row before it can be used, and this parsing is by far the most time 
consuming part of processing VCF data. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset, using wormtable and using PyVCF[citation?]. Both of these
methods proceeded row-by-row, checking the ALT and REF columns. The 
PyVCF method required X hours to complete, while the wormtable 
based script needed only Y seconds. This stark difference is not a 
criticism of the PyVCF library, which provides an efficient parser 
for VCF files; it is rather a measure of the inherent slowness of
parsing text in large volumes.

It is well known, of course, that plain text cannot be retrieved 
efficiently at large scales. The BVCF format is an binary version 
of VCF which is intended to avoid this overhead [citation?]. 
This binary format does alleviate the problem of parsing overhead,
but does not solve the problem of indexing \emph{within} rows.

To illustrate the use of indexes in wormtable, 
we created an index on the ALT and REF columns, 
and another on the IN, UT, SY and NS columns. These indexes required 
X and Y MB of disc space and X and Y minutes to build, respectively. 
The ALT+REF index provides a much more efficient means of counting 
the transitions and transversions, and another script to calculate 
these required x seconds. Using the IN+UT+SY+NS index, we 
calculated the mean nucleotide diversity for SNPs (not indels) 
at synonymous, nonsynonymous, intronic and UTR sites. This required 
X seconds. Computing the same values with PyVCF required Y hours.

The examples in this section are intended to illustrate the 
magnitude of performance gains that wormtable can provide 
when working with large datasets. The gains are many: we 
do not have to parse the information before it is used;
we can quickly seek to arbitrary rows within a table; and 
we can use indexes to intelligently sort and aggregate our
data to make it more useful.

\section{Conclusions}
Text files are the oldest and most reliable means of exchanging 
information between computers. Plain text, however, does not provide an efficient 
means of storing and processing large volumes of data. Current practice 
in bioinformatics is to encode data in more and more complex plain 
text data formats. While these formats are excellent from the perspective 
of minimising software dependencies, they are very inefficient in terms of 
storage space, the time required to parse them, and the linear
seek times to find particular rows.

Wormtable is not intended to replace text files as the universal 
interchange format. It is intended to provide a persistent data structure 
that efficiently holds data in a form that can be efficiently processed 
and searched. 
% Furthermore, by providing command line programs to import and index common
% formats, we provide a very intuitive and easy to use method to deal
% with standard bioinformatic formats....
Using this data structure, researchers with no knowledge of 
database systems can take full advantage of sophisticated 
data management techniques, and write simple code that processes this 
data in a very high performance manner. In our example above, 
we [blah blah blah]. A systematic application of these techniques 
can substantially increase the researcher's productivity and ability
to explore their data.

Wormtable is an open and collaborative project seeking feedback and 
contributors. The project is in its early stages, currently 
providing only a Python interface. A C library, Perl and Java interfaces
are also planned for Wormtable, which should make the format available
to the majority of bioinformatics projects. With such broad support,
it is hoped that other software tools may use these libraries 
to support direct output of data in Wormtable format.

\section*{Acknowledgement}
\paragraph{Funding\textcolon}  
JK is supported by EPSRC grant EP/I013091/1; DLH is supported by XXXXX; 
and RWN is supported by XXXX.

\bibliographystyle{plainnat}
\bibliography{paper}

\end{document}

### 2013-07-02 docs/paper/paper.tex
Fixed author order and started reshuffle.

@@
 
-\title[Wormtable]{Wormtable: A data structure for genome scale tabular data}
-\author[Kelleher \textit{et~al}]{Jerome Kelleher, Daniel L. Halligan 
-    and Robert W. Ness }
+\title[Wormtable]{Wormtable: A data structure for genome scale data}
+\author[Kelleher \textit{et~al}]{Jerome Kelleher, Robert W. Ness 
+and Daniel L. Halligan}
 \address{
@@
 Biological research generates vast quantities of tabular data, which must 
-be processed to answer biological questions. These tables are produced by 
-programs and exported to plain text files, which are often tens of gigabytes
-long. Researchers must process these files 
-line-by-line, parsing values encoded as plain text into native types so that 
-calculations may be done. This is an enormously inefficient process.
+be processed to test hypotheses. These tables are produced by 
+programs and exported to plain text files, which are often very large. 
+Researchers must process these files 
+line-by-line, parsing values encoded as text into native types so that 
+calculations may be done. This is an extremely inefficient process.
 Furthermore, there is no simple means of indexing these files so that 
 arbitrary data values within rows can be quickly found. We introduce a new 
-data format and software library called Wormtable, which provides efficient 
-access to tabular data using languages such as Python. Wormtable avoids the 
-overhead of repeatedly parsing text, random access to rows, 
-and also provides efficient indexing of arbitrary columns within these tables.
+data format and software library called wormtable, which provides efficient 
+access to tabular data in Python. Wormtable stores data in a compact
+binary format, provides random access to rows, and enables sophisticated
+indexing on columns within these tables.
 \section{Availability:}
@@
 \section{Introduction}
-The large volumes of data that present day bioinformatics generates 
-is well documented. The improvement in sequencing technologies has
-lead to an exponential increase in the volume of data that is available 
-to biologists. Perfectly standard studies require the storage and processing 
+% Lots of data
+The improvement in sequencing technologies has
+lead to an exponential increase in the volume of data being managed 
+by biologists. Standard studies require the storage and processing 
 of hundreds of gigabytes of data, and this volume of data is only set to 
-increase as sequencing technologies continue to improve. [\emph{This 
-intro is intended to set the stage: there is shit loads of data out there,
-and this is only set to increase}].
+increase as sequencing technologies continue to improve.
 
+% but we use text files
 Despite the increasing importance of data processing, however, the methods 
-used in Bioinformatics are crude and inefficient being almost entirely 
+used in Bioinformatics are almost entirely 
 based around the use of text files. Plain text files have many advantages:
-they are the most portable format available, and can be processed on 
+they are extremely portable and can be processed on 
 any platform with the minimum of library dependencies. Text files also 
@@
 purposes here are, they are slow to parse and difficult to index.
+
 % text files are also easy to read/comprehend and don't require specialist
@@
 
-A popular file format for Bioinformatics use is the Variant Call Format~\citep{da+11},
-or VCF. In VCF,  information about variant sites in a genome is encoded 
-as tab-delimited rows in a text file. To access information about a 
-particular site, one must proceed through the file line-by-line until the 
-desired record is found. There are methods available to index this 
-file~\citep{li11}, but these have no semantic 
-understanding of the structure of the file and so are limited in the 
-fields that can be indexed. One cannot, for example, build an index 
-to quickly obtain all records in which the quality field is greater
-than some value. Such queries can take a considerable time on large VCF 
-files, which discourages testing and checking for the effects of 
-arbitrary thresholds in analyses.
+% Databases solve these problems
+The classical approach to working with large volumes of 
+structured data is to use a database, which solve these 
+problems very effectively.
+Data is typically encoded in an efficient binary 
+format, saving substantial space and giving better precision when 
+dealing with floating point formats. Records are stored in a
+B-tree type structure, leading to fast access times even 
+over very large datasets. Indexing over columns within rows is 
+also very powerful, allowing us to quickly retrieve 
+rows with particular features.
 
-One may, of course, load VCF data into a relational database, and 
+% Databases but introduce several others. 
+One may, of course, load textual data into a relational database, and 
 use SQL to retrieve the records of interest. This approach has several 
-disadvantages, however. Firstly, one must first examine the VCF 
-header and construct an appropriate SQL schema to hold the columns 
-of interest. It is feasible to do this for more recent versions of VCF, 
-but is certainly not a trivial process, particularly if the schema 
+disadvantages, however. Firstly, one must first examine the data 
+construct an appropriate SQL schema to hold the columns 
+of interest. It is sometimes feasible to do this automatically 
+for data files, but is certainly not a trivial process, 
+particularly if the schema 
 must be portable across different relational database platforms.
@@
 
+% We don't need concurrency, etc, as it's always read-only.
 Much of the time, a database server is far more complicated than
-we require for large scale biological data. In the case of VCF,
-the file is written once by an application such as [gatk? references?]
-%Also samtools
+we require for large scale biological data. Data files 
+are usually written once by an application 
 and is not subsequently expected to change. Thus, storing this information 
@@
 
-Databases do, however, have significant advantages over 
-plain text files. Data is typically encoded in an efficient binary 
-format, saving substantial space and giving better precision when 
-dealing with floating point formats. Records are stored in a
-B-tree type structure, leading to fast access times even 
-over very large datasets. Indexing over columns within rows is 
-also very powerful, allowing us to quickly retrieve 
-rows with particular features. Ideally, we would like a data format
-that combines these useful features of databases without the 
-drawbacks of a relational database server outlined above.
+\section{Wormtable}
 
-Wormtable 
-%... [Write Once Read Only Table] ...
+Wormtable (write-once read-many table) 
 is a new data format and software library designed specifically
@@
 
-% Also provided are scripts to import two common bioinformatic file standards - gff and vcf
-% along with scripts to apply filters to these files?
-
-\section{Wormtable}
 To take advantage of the data processing capabilities of a database
@@
 \section{Examples}
+
+A popular file format for Bioinformatics use is the Variant Call Format~\citep{da+11},
+or VCF. In VCF,  information about variant sites in a genome is encoded 
+as tab-delimited rows in a text file. To access information about a 
+particular site, one must proceed through the file line-by-line until the 
+desired record is found. There are methods available to index this 
+file~\citep{li11}, but these have no semantic 
+understanding of the structure of the file and so are limited in the 
+fields that can be indexed. One cannot, for example, build an index 
+to quickly obtain all records in which the quality field is greater
+than some value. Such queries can take a considerable time on large VCF 
+files, which discourages testing and checking for the effects of 
+arbitrary thresholds in analyses.
+
 Wormtable is designed to make working with genome scale data more 

### 2013-07-03 _wormtablemodule.c
Added initial support for configurable page size.

@@ -82,6 +82,7 @@ typedef struct {
     PyObject *filename;
     Column **columns;
     Py_ssize_t cache_size;
+    u_int32_t page_size;
     u_int32_t num_columns;
     u_int32_t fixed_region_size;
     void *row_buffer;
@@ -1782,7 +1783,8 @@ static int
 Table_init(Table *self, PyObject *args, PyObject *kwds)
 {
     int ret = -1;
-    static char *kwlist[] = {"filename", "columns", "cache_size", NULL}; 
+    static char *kwlist[] = {"filename", "columns", "cache_size", "page_size",
+            NULL}; 
     Column *col;
     PyObject *filename = NULL;
     PyObject *columns = NULL;
@@ -1792,10 +1794,14 @@ Table_init(Table *self, PyObject *args, PyObject *kwds)
     self->columns = NULL;
     self->filename = NULL;
     self->cache_size = 0;
-    if (!PyArg_ParseTupleAndKeywords(args, kwds, "O!O!n", kwlist, 
+    /* TODO: take page size out of this initialiser and make it 
+     * an optional setter. There should be a proper getter method 
+     * also which queries DB */
+    self->page_size = 64 * 1024;
+    if (!PyArg_ParseTupleAndKeywords(args, kwds, "O!O!n|n", kwlist, 
             &PyBytes_Type, &filename, 
             &PyList_Type,  &columns, 
-            &self->cache_size)) {
+            &self->cache_size, &self->page_size)) {
         goto out;
     }
     self->filename = filename;
@@ -1838,6 +1844,7 @@ out:
 static PyMemberDef Table_members[] = {
     {"filename", T_OBJECT_EX, offsetof(Table, filename), READONLY, "filename"},
     {"cache_size", T_PYSSIZET, offsetof(Table, cache_size), READONLY, "cache_size"},
+    {"page_size", T_PYSSIZET, offsetof(Table, page_size), READONLY, "page_size"},
     {"fixed_region_size", T_INT, offsetof(Table, fixed_region_size), READONLY, 
             "fixed_region_size"},
     {NULL}  /* Sentinel */
@@ -1986,13 +1993,13 @@ Table_open(Table* self, PyObject *args)
         handle_bdb_error(db_ret);
         goto out;    
     }
-    /* Disable DB error messages */
-    self->db->set_errcall(self->db, NULL);
-    db_ret = self->db->set_pagesize(self->db, 64 * 1024);
[…]

### 2013-07-03 wormtable.py

@@ -465,11 +465,23 @@ class Table(Database):
     """
     DB_NAME = "table"
     PRIMARY_KEY_NAME = "row_id"
+    DEFAULT_PAGE_SIZE = 64 * 1024
+
     def __init__(self, homedir):
         Database.__init__(self, homedir, self.DB_NAME)
         self.__columns = []
         self.__column_name_map = {}
         self.__num_rows = 0
+        self.__page_size = self.DEFAULT_PAGE_SIZE
+
+    def set_page_size(self, page_size):
+        """
+        Sets the Berkeley DB page size to the specified value.
+
+        This only has an effect when a table is first created; afterwards,
+        the page size for a table is fixed.
+        """
+        self.__page_size = page_size
 
     def _create_ll_object(self, path):
         """
@@ -478,7 +490,8 @@ class Table(Database):
         """ 
         filename = path.encode() 
         ll_cols = [c.get_ll_object() for c in self.__columns]
-        t = _wormtable.Table(filename, ll_cols, self.get_cache_size())
+        t = _wormtable.Table(filename, ll_cols, self.get_cache_size(), 
+                self.__page_size)
         return t
 
     def get_fixed_region_size(self):

### 2013-07-04 docs/paper/paper.tex

@@
 Data is typically encoded in an efficient binary 
-format, saving substantial space and giving better precision when 
-dealing with floating point formats. Records are stored in a
-B-tree type structure, leading to fast access times even 
+format, saving substantial space and avoiding the need
+for repeated parsing. Records are usually stored in a
+B-tree, leading to fast access times even 
 over very large datasets. Indexing over columns within rows is 
@@
 disadvantages, however. Firstly, one must first examine the data 
-construct an appropriate SQL schema to hold the columns 
-of interest. It is sometimes feasible to do this automatically 
+and construct an appropriate SQL schema to define the appropriate columns. 
+It is sometimes feasible to do this automatically 
 for data files, but is certainly not a trivial process, 
@@
 must be portable across different relational database platforms.
-Once the schema has been constructed, the data must be inserted into 
-this database line-by-line by constructing SQL code, which is a 
-time-consuming and error prone process. Even assuming that the data 
+Even assuming that the data 
 has been inserted into a table with an appropriate schema, there 
@@
 Much of the time, a database server is far more complicated than
-we require for large scale biological data. Data files 
+we require for biological data. Data files 
 are usually written once by an application 
-and is not subsequently expected to change. Thus, storing this information 
+and do not subsequently change. Thus, storing this information 
 in a relational database with its sophisticated concurrency control
 and guarantees of consistency in the face of asynchronous updates
-is entirely unnecessary. Even the client-server architecture of 
-a typical relational database is wasteful in this case: protocol 
-commands must be read and parsed, and responses must be encoded,
-written to a socket and subsequently converted back to native types,
-all of which constitutes a significant overhead when streaming large 
-quantities of data. The limitations of relational database servers 
-have been acknowledged for some time now, and has led to a 
-variety of `NoSQL' database systems~\citep{m12}.
+is entirely unnecessary. Interacting with a database is also a
+substantial hurdle, as another lanuage must be learned in order 
+to gain access to data. Relational databases, although providing 
+powerful data management technologies are far more complex than 
+is required, and are therefore little used for the types of 
+data that we are interested in.
 
 \section{Wormtable}
-
+% what is wormtable?
 Wormtable (write-once read-many table) 
@@
 It provides a portable, compact, read-only store for tabular
-data with high-performance access to rows via indexes. Tables can
-be either written directly via the software library, or converted
+data of essentially unlimited size. Tables can
+be either written directly via the library, or converted
 from existing formats via command line tools. Tables can be read 
@@
 
-To take advantage of the data processing capabilities of a database
-without the overheads and disadvantages of a relational database 
-server, Wormtable is built on top of Berkeley DB~\citep{obs99}.
+% why do we use Berkeley DB?
+To take advantage of the data processing capabilities of database
+technologies without the overheads and disadvantages of a relational database 
+server, wormtable uses Berkeley DB~\citep{obs99} for row storage and indexing.
 Berkeley DB is an open source embedded database toolkit, that provides a
-scalable key-value store. It is a mature and stable platform, 
+scalable key-value store~\citep{m12}. It is a mature and stable platform, 
 and is currently the most widely deployed database toolkit in the 
-world~\citep{sb12}.
+world~\citep{sb12}. DB is a highly modular library, providing the 
+means to define a very lightweight read-only database, without the 
+overheads of a server process
 
-Berkeley DB (or simply DB) 
-differs from relational database servers in several ways. 
-Firstly, it is a software library that a program links against, 
-and so there is no need for a separate database server process.
-Secondly, DB provides only a key-value store, and so has 
-no knowledge of the structure of the records that it stores,
-treating them simply as an array of bytes. While there 
-are sophisticated 
-mechanisms within DB for locking, transactions and so on, 
-these are all optional, allowing a very lightweight 
-read-only store to be built using the library.
+% Wormtable's binary row format 
+On this solid foundation of row storage in Berkeley DB, wormtable uses a
+compact binary row format designed specifically for biological applications.
+Each row contains data from a fixed number of columns, and each column 
+holds either a fixed or variable number of elements of a fixed type and size.
+Supported types are signed and unsigned integers, 
+floating point and character data. To ensure that values are 
+stored as compactly as possible, wormtable supports integer values from 
+$1$ to $8$ bytes, and the standard single and double precision 
+floating point values.
 
-This is the approach taken in Wormtable. We assume that tabular data 
-is written once, by some program generating the data, and subsequently 
-read many times. Under these assumptions, the primary database 
-holding the rows can be opened in a read-only state, allowing any 
-number of processes to read it concurrently in a lock free manner.
-
-Since DB is agnostic to the structure of data stored in rows, we 
-are free to design our own formats. In Wormtable, the row format is
-a compact and flexible binary format holding integer, floating point 
-or character values of various sizes. Each column in a row may contain 
-either a fixed or variable number of elements (each element being of a
-fixed type and size). This row format enables two important properties:
-firstly, values from any column can be accessed in constant 
-time; and secondly, values in columns can be converted directly into 
-native machine values, ready for calculations.
-
-Indexing of columns is achieved via DB also. An index is simply a
-database mapping values in a column (or list of columns) to rows 
-in the primary database. Indexes allow us to find rows with particular 
-properties very quickly, and are a very useful method of excluding 
-rows with certain properties from our analyses.
+% Wormtable's indexes 
+An essential element of database technology is the ability to index tables
+to rows with certain properties to be quickly retrieved. Wormtable 
+supports indexing over columns or groups of columns within a table, 
+making many operations extremely efficient.
 
@@
 
+% What is VCF?
 A popular file format for Bioinformatics use is the Variant Call Format~\citep{da+11},
@@
 we converted a VCF file with XXX rows from the YYY project 
-into wormtable format. The (uncompressed) VCF file was XX
+into wormtable format using the included \texttt{vcf2wt} 
+conversion program.  The (uncompressed) VCF file was XX
 gigabytes, and the converted wormtable YY gigabytes, which required 
@@
 we created an index on the ALT and REF columns, 
-and another on the IN, UT, SY and NS columns. These indexes required 
-X and Y MB of disc space and X and Y minutes to build, respectively. 
+This index required X MB of disc space Y minutes to build, respectively. 
 The ALT+REF index provides a much more efficient means of counting 
 the transitions and transversions, and another script to calculate 
-these required x seconds. Using the IN+UT+SY+NS index, we 
-calculated the mean nucleotide diversity for SNPs (not indels) 
-at synonymous, nonsynonymous, intronic and UTR sites. This required 
-X seconds. Computing the same values with PyVCF required Y hours.
+these required x seconds.
 
@@
 \section{Conclusions}
+% text files are good for archiving, but terrible for processing
 Text files are the oldest and most reliable means of exchanging 
@@
 
+% We're not trying to define a universal bioinformatics file format,
+% just to help people be more productive.
 Wormtable is not intended to replace text files as the universal 
@@
 that efficiently holds data in a form that can be efficiently processed 
-and searched. 
-% Furthermore, by providing command line programs to import and index common
-% formats, we provide a very intuitive and easy to use method to deal
-% with standard bioinformatic formats....
-Using this data structure, researchers with no knowledge of 
+and searched. Using this data structure, researchers with no knowledge of 
 database systems can take full advantage of sophisticated 
 data management techniques, and write simple code that processes this 
-data in a very high performance manner. In our example above, 
-we [blah blah blah]. A systematic application of these techniques 
-can substantially increase the researcher's productivity and ability
-to explore their data.
+data in a very high performance manner. 
+By providing conversion programs to convert 
+common formats to wormtable, we also provide an easy to 
+use common platform for 
+bioinformatics analysis. This reduces the need for third party libraries 
+to parse complex files and considerably simplifies the code required 
+to process data.
+Together, these advantages of increased performance and reduced 
+code complexity can substantially increase a researcher's 
+productivity and ability to explore their data.
 
+% If we get some help, it could be pretty awesome...
 Wormtable is an open and collaborative project seeking feedback and 
 contributors. The project is in its early stages, currently 
-providing only a Python interface. A C library, Perl and Java interfaces
-are also planned for Wormtable, which should make the format available
-to the majority of bioinformatics projects. With such broad support,
-it is hoped that other software tools may use these libraries 
-to support direct output of data in Wormtable format.
+providing a conversion tool for the VCF format only. With community
+participation it may grow to support many other formats and become
+a powerful common platform for bioinformatics data analysis.
 

### 2013-07-04 examples/count-keys.py
Minor edits

@@ -15,7 +15,7 @@ def count(homedir, index):
     with wt.open_table(homedir) as t, t.open_index(index) as i:
         table = [[k,v] for k,v in i.counter().items()]
         assert(len(t) == sum(r[1] for r in table))
-        return(table)
+    return(table)
 
 
 def main():

### 2013-07-04 docs/paper/paper.tex
Suggestions + edits to examples section

@@
 based around the use of text files. Plain text files have many advantages:
-they are extremely portable and can be processed on 
+they are extremely portable, easy to view and interpret and can be processed on 
 any platform with the minimum of library dependencies. Text files also 
@@
 The classical approach to working with large volumes of 
-structured data is to use a database, which solve these 
+structured data is to use a database, which can solve these 
 problems very effectively.
@@
 rows with particular features.
+
+% there are a range of standard (well defined) tabular data formats: vcf, gtf
+% lends itself to a database solution geared towards specific file types
 
@@
 
+% are other database solutions less flexible when it comes to where the 
+% data is stored?
+
 \section{Wormtable}
@@
 and is currently the most widely deployed database toolkit in the 
-world~\citep{sb12}. DB is a highly modular library, providing the 
+world~\citep{sb12}. DB is a highly modular library, providing the  % do you mean Berkeley DB?
 means to define a very lightweight read-only database, without the 
@@
 A popular file format for Bioinformatics use is the Variant Call Format~\citep{da+11},
-or VCF. In VCF,  information about variant sites in a genome is encoded 
+or VCF. In a VCF file, information about variant sites in a genome is encoded 
 as tab-delimited rows in a text file. To access information about a 
-particular site, one must proceed through the file line-by-line until the 
+particular site naively, one must proceed through the file line-by-line until the 
 desired record is found. There are methods available to index this 
-file~\citep{li11}, but these have no semantic 
+file~\citep{li11} by genomic position, but these have no semantic 
 understanding of the structure of the file and so are limited in the 
@@
 Wormtable is designed to make working with genome scale data more 
-efficient, and in this section we describe some examples to 
-illustrate the improvements over existing methods. To do this, 
-we converted a VCF file with XXX rows from the YYY project 
-into wormtable format using the included \texttt{vcf2wt} 
-conversion program.  The (uncompressed) VCF file was XX
-gigabytes, and the converted wormtable YY gigabytes, which required 
-X hours.
+efficient, and in this section we describe some examples of wormtable
+that illustrate the improvements over existing methods. We applied these 
+examples to a large, publicly available VCF file produced as part of the
+"Drosophila Genetic Reference Panel". This VCF ("freeze2.vcf.gz, available 
+here: ftp://ftp.hgsc.bcm.edu/DGRP/freeze2_Feb_2013/) contains variant calls 
+from a population consisting 192 inbred Drosophila lines and contains data 
+from the whole genome over a total of 6,146,611 rows). This VCF was converted to 
+wormtable format using the included \texttt{vcf2wt} conversion program.  
+The (uncompressed) VCF file was 15Gb, and the converted wormtable was 12Gb,
+which required X hours.
 
 Values are stored in wormtable in a binary format, so that no parsing is
-required when reading in rows. Methods using the VCF directly must parse 
-each row before it can be used, and this parsing is by far the most time 
+required when reading in rows whereas methods that use the VCF directly 
+must parse each row before it can be used, and is by far the most time 
 consuming part of processing VCF data. To illustrate this advantage, we
 wrote a script to count the number of transitions and transversions
-in the dataset, using wormtable and using PyVCF[citation?]. Both of these
-methods proceeded row-by-row, checking the ALT and REF columns. The 
+in the dataset, using wormtable and using PyVCF[citation?], a module for
+python to parse VCF files. In both of these examples we proceed 
+row-by-row, checking the REF and ALT columns for the type of variant and
+summing transitions and transversions (ignoring indels). The 
 PyVCF method required X hours to complete, while the wormtable 
@@
 It is well known, of course, that plain text cannot be retrieved 
-efficiently at large scales. The BVCF format is an binary version 
+efficiently at large scales. The BVCF format is a binary version 
 of VCF which is intended to avoid this overhead [citation?]. 
@@
 
-To illustrate the use of indexes in wormtable, 
-we created an index on the ALT and REF columns, 
-This index required X MB of disc space Y minutes to build, respectively. 
-The ALT+REF index provides a much more efficient means of counting 
-the transitions and transversions, and another script to calculate 
-these required x seconds.
+To illustrate the use of indexes in wormtable, we created an index 
+on the REF and ALT columns. This index required X MB of disc space 
+Y minutes to build, respectively. The REF+ALT index provides a much more 
+efficient means of counting the number of transitions and transversions 
+(we can just count the number of rows matching keys in the index 
+corresponding the SNPs of each type). A script to calculate these 
+counts by this method required x seconds.
 

### 2013-07-04 docs/paper/paper.tex
Added some timings

@@
 wormtable format using the included \texttt{vcf2wt} conversion program.  
-The (uncompressed) VCF file was 15Gb, and the converted wormtable was 12Gb,
-which required X hours.
+The (uncompressed) VCF file was 15GB, and the converted wormtable was 12GB 
+and took 30 minutes, 55 seconds to create.
 
@@
 wrote a script to count the number of transitions and transversions
-in the dataset, using wormtable and using PyVCF[citation?], a module for
+in the dataset, using wormtable and using PyVCF[http://pyvcf.readthedocs.org/?], a module for
 python to parse VCF files. In both of these examples we proceed 
@@
 To illustrate the use of indexes in wormtable, we created an index 
-on the REF and ALT columns. This index required X MB of disc space 
-Y minutes to build, respectively. The REF+ALT index provides a much more 
+on the REF and ALT columns. This index required 66MB of disc space 
+3 minutes 50 seconds to build, respectively. The REF+ALT index provides a much more 
 efficient means of counting the number of transitions and transversions 

### 2013-07-05 examples/vcf-utils/snp-filter.py
added snp-filter.py script to perform arbitrary filters on a vcf wormtable

@@ -1,0 +1,105 @@
+#!/usr/bin/env python
+
+"""
+Filter variant calls and print selected columns
+"""
+
+import re
+import sys
+import math
+import wormtable as wt
+import argparse
+import operator
+
+ops = {
+    ">": operator.gt,
+    "<": operator.lt,
+    ">=": operator.ge,
+    "<=": operator.le,
+    "==": operator.eq,
+    "!=": operator.ne
+}
+
+class dictcursor(object):
+    def __init__(self, table, cols, index):
+        self.__cursor = table.cursor(cols, index)
+        self.__cols = cols
+        
+    def __iter__(self):
+        for row in self.__cursor:
+            yield {self.__cols[i]:row[i] for i in range(len(row))}
+
+    def set_min(self, *args):
+        self.__cursor.set_min(*args)
+
+    def set_max(self, *args):
+        self.__cursor.set_max(*args)
+
+
+def make_filter(s):
+    if s[1] in ['>','<','>=','<=']:   # make all of these comparisons numeric
+        return lambda x: ops[s[1]](x[s[0]], int(s[2]))
+    else:
+        return lambda x: ops[s[1]](str(x[s[0]]), s[2])
+
+
+def parse_cf(cf):
+    fns = []
+    cols = []
+    for filter in cf.split(';'):
+        s = re.split('(>=|<=|>|<|==)', filter)
+        fns.append(make_filter(s))
+        cols.append(s[0])
+    return fns, cols
+
+
+def filter(t, i, args):
+    pcols = args['cols'].split(',')
+    
+    allcols = pcols
+    if args['imode'] in ['e','f']:  
+        allcols = allcols+['REF','ALT']
+    
+    # parse functions
+    fns, fcols = parse_cf(args['cf'])
+    allcols = allcols+fcols
+    cols = list(set(allcols))
+    dc = dictcursor(t, allcols, i)
+    
+    # set region if defined
+    chr,start,end = re.split('\W', args['r'])
+    dc.set_min(chr, int(start))
+    dc.set_max(chr, int(end)+1)
+    
+    for row in dc:
+        if all([fn(row) for fn in fns]):
+            yield [str(row[c]) for c in pcols]
+
+
+def main():
+    parser = argparse.ArgumentParser(description=globals()['__doc__'])
+    
+    parser.add_argument('cols', default="CHROM,POS",
+        help='comma separated column names to print')
+
+    parser.add_argument('--imode', default='i', choices=['i','e','f'],
[…]

### 2013-07-06 examples/sliding-window.py
Restructued examples - different methods to count ts,tv are now in one script where you can select the method to use

@@ -1,195 +1,0 @@
-#!/usr/bin/env python
-
-from __future__ import print_function
-from __future__ import division 
-
-import re
-import sys
-import math
-import numpy as np
-import matplotlib.pyplot as plt
-import wormtable as wt
-
-def isindel(ref,alt): 
-    """
-    Returns True if the specified ref and alt VCF entries represent 
-    and indel. This is True if either contain two consecutive word characters
-    """
-    return re.search('\w\w', ref) or re.search('\w\w', alt)
-
-class SlidingWindow(object):
-    """
-    A class representing a sliding window over statistics in a VCF 
-    wormtable.
-    """
-    def __init__(self, homedir, window=500):
-        self.__table = wt.open_table(homedir, cache_size="256M")
-        self.__index = self.__table.open_index("CHROM+POS", cache_size="256M")
-        self.__window = 500
-
-    def run(self, chrom):
-        """
-        Runs the sliding window over the specified chromosome and writes
-        out plots.
-        """
-        cols = ["POS", "REF", "ALT", "INFO_AF"]
-        cursor = self.__table.cursor(cols, self.__index)
-        start = self.__index.get_min(chrom)[1]
-        end = self.__index.get_max(chrom)[1]
-        n = 1 + math.ceil((end - start) / self.__window)
-        position = np.zeros(n, dtype=np.int)
-        count = np.zeros(n, dtype=np.int)
-        indels = np.zeros(n, dtype=np.int)
-        allele_frequency = np.zeros(n)
-        cursor.set_min(chrom, start)
-        cursor.set_max(chrom, end + 1)
-        position[0] = start
-        position[-1] = end 
-        j = 0
-        for pos, ref, alt, af in cursor:
-            count[j] += 1
-            allele_frequency[j] += af[0]
-            if isindel(ref, alt):
-                indels[j] += 1
-            if pos >= start + j * self.__window:
-                position[j] = pos
-                j += 1
-        # Get the mean allele frequency
-        count_div_safe = np.array(count)
-        count_div_safe[count==0] = 1
-        allele_frequency /= count_div_safe 
-        # Now plot these data 
-        plt.plot(position, count)
-        plt.xlabel("position")
-        plt.ylabel("count")
[…]

### 2013-07-07 examples/count-ts-tv-methods.py
Tidied up ts-tv example.

@@ -14,43 +14,37 @@ import vcf
 from itertools import permutations
 import wormtable as wt
 
-bases = {'A':'purine', 'G':'purine', 'C':'pyrimidine', 'T':'pyrimidine'}
+bases = {b'A':b'purine', b'G':b'purine', b'C':b'pyrimidine', b'T':b'pyrimidine'}
 
 def count_Ts_Tv_pyvcf(vcf_file):
-	""" 
-	Count number of transitions and transversions using pyVCF
-	"""
-	vcf_reader = vcf.Reader(filename=vcf_file)
-	Ts, Tv = 0, 0
-	for record in vcf_reader:
-		if str(record.REF) in bases.keys() and \
-		record.ALT[0] in bases.keys() \
-		and len(record.ALT) == 1:
-			if bases[record.REF] == bases[str(record.ALT[0])]:
-				Ts +=1
-			else:
-				Tv +=1
-	return Ts, Tv
+    """ 
+    Count number of transitions and transversions using pyVCF
+    """
+    vcf_reader = vcf.Reader(filename=vcf_file)
+    Ts, Tv = 0, 0
+    for r in vcf_reader:
+        if r.REF in bases and r.ALT[0] in bases and len(r.ALT) == 1:
+            if bases[r.REF] == bases[r.ALT[0]]:
+                Ts +=1
+            else:
+                Tv +=1
+    return Ts, Tv
 
 
 def count_Ts_Tv_wtcursor(homedir):
-	""" 
-	Count number of transitions and transversions using wormtable and an 
-	index on CHROM+POS, counting Ts and Tv row by row
-	"""
-	t = wt.open_table(homedir)
-	i = t.open_index("CHROM+POS")
-	c = t.cursor(["REF","ALT"], i)
-	Ts, Tv = 0, 0
-	for row in c:
-		if str(row[0]) in bases.keys() and str(row[1]) in bases.keys():
-			if bases[str(row[0])] == bases[str(row[1])]:
-				Ts +=1
-			else:
-				Tv +=1
-	i.close()
-	t.close()
-	return Ts, Tv
+    """ 
+    Count number of transitions and transversions using wormtable and an 
+    index on CHROM+POS, counting Ts and Tv row by row
+    """
+    with wt.open_table(homedir) as t:
+        Ts, Tv = 0, 0
+        for ref, alt in t.cursor(["REF", "ALT"]):
+            if ref in bases and alt in bases:
[…]

### 2013-07-07 examples/count-ts-tv-methods.py
Made counts consistent by insisting alt != ref.

@@ -20,11 +20,12 @@ def count_Ts_Tv_pyvcf(vcf_file):
     """ 
     Count number of transitions and transversions using pyVCF
     """
-    vcf_reader = vcf.Reader(filename=vcf_file)
     Ts, Tv = 0, 0
-    for r in vcf_reader:
-        if r.REF in bases and r.ALT[0] in bases and len(r.ALT) == 1:
-            if bases[r.REF] == bases[r.ALT[0]]:
+    for r in vcf.Reader(filename=vcf_file):
+        alt = r.ALT
+        ret = r.REF
+        if ref != alt and ref in bases and alt in bases:
+            if bases[ref] == bases[alt]:
                 Ts +=1
             else:
                 Tv +=1
@@ -39,7 +40,7 @@ def count_Ts_Tv_wtcursor(homedir):
     with wt.open_table(homedir) as t:
         Ts, Tv = 0, 0
         for ref, alt in t.cursor(["REF", "ALT"]):
-            if ref in bases and alt in bases:
+            if ref != alt and ref in bases and alt in bases:
                 if bases[ref] == bases[alt]:
                     Ts +=1
                 else:
@@ -52,12 +53,10 @@ def count_Ts_Tv_wtindex(homedir):
     Count number of of transitions and transversions using wormtable and 
     an index on REF+ALT
     """
-    subs = [p for p in permutations(bases.keys(), 2)]
-    print(subs)
     with wt.open_table(homedir) as t, t.open_index("REF+ALT") as i:
         Ts, Tv = 0, 0
         c = i.counter()
-        for s in subs:
+        for s in permutations(bases.keys(), 2):
             if bases[s[0]] == bases[s[1]]: 
                 Ts += c[s] 
             else: 

### 2013-07-07 wormtable.py
Minor cleanup + progress bar fix.

@@ -44,7 +44,6 @@ INDEX_METADATA_VERSION = "0.1"
 DEFAULT_CACHE_SIZE = 16 * 2**20 # 16M 
 DEFAULT_CACHE_SIZE_STR = "16M" 
 
-
 WT_INT = _wormtable.WT_INT
 WT_UINT = _wormtable.WT_UINT
 WT_FLOAT = _wormtable.WT_FLOAT
@@ -1079,7 +1078,7 @@ class ProgressMonitor(object):
         self.__bar_index = (self.__bar_index + 1) % len(self.__bars)
         elapsed = max(1, time.clock() - self.__start_time)
         rate = processed / elapsed
-        s = '\r[{0}{1}] {2:2.2f}% @{3:4.2G} {4}/s {5}'.format('#' * filled, 
+        s = '\r[{0}{1}] {2:5.1f}% @{3:8.1E} {4}/s {5}'.format('#' * filled, 
             ' ' * spaces, complete * 100, rate, self.__units, bar)
         sys.stdout.write(s)
         sys.stdout.flush()

### 2013-07-07 _wormtablemodule.c
Removed printfs and old TODOs.

@@ -42,16 +42,6 @@
 
 static PyObject *WormtableError;
 
-/* TODO:
- * 1) We need much better error reporting for the parsers. These should have a
- * top level variadic function that takes arguments and set the Python exception
- * appropriately.
- * 3) The numerical types need a little cleaning up and thought put into. We
- * now have single and double floating point sizes and should export constants
- * to tell what they are. There should also be some constants telling range
- * of integers and so on.
- */
-
 typedef struct Column_t {
     PyObject_HEAD
     PyObject *name;
@@ -489,12 +479,6 @@ Column_unpack_elements_float(Column *self, void *source)
 static int 
 Column_unpack_elements_char(Column *self, void *source)
 {
-    /*
-    char v[1024];
-    memcpy(v, source, self->num_buffered_elements); 
-    v[self->num_buffered_elements] = 0;
-    printf("unpacked: '%s': %d\n", v, self->num_buffered_elements);
-    */
     memcpy(self->element_buffer, source, self->num_buffered_elements); 
     return  0; 
 }
@@ -543,7 +527,6 @@ Column_pack_elements_int(Column *self, void *dest)
     int64_t *elements = (int64_t *) self->element_buffer;
     int64_t u;
     for (j = 0; j < self->num_buffered_elements; j++) {
-        //printf("\npacking :%ld\n", elements[j]); 
         u = elements[j];
         /* flip the sign bit */
         u ^= 1LL << (self->element_size * 8 - 1);
@@ -587,12 +570,6 @@ static int
 Column_pack_elements_char(Column *self, void *dest)
 {
     int ret = -1;
-    /*
-    char v[1024];
-    memcpy(v, self->element_buffer, self->num_buffered_elements); 
-    v[self->num_buffered_elements] = 0;
-    printf("packed: '%s': %d\n", v, self->num_buffered_elements);
-    */
     memcpy(dest, self->element_buffer, self->num_buffered_elements); 
     ret = 0;
     return ret; 
@@ -750,7 +727,6 @@ Column_truncate_elements_float(Column *self, double bin_width)
         if (u != missing_value) {
             elements[j] = u - fmod(u, bin_width); 
         }
-        //printf("truncating :%f by %f = %f\n", u, bin_width, elements[j]);    
     }
     ret = 0;
 out:
@@ -1303,7 +1279,6 @@ Column_update_row(Column *self, void *row, uint32_t row_size)
                 num_elements) < 0) {
             goto out;
         }
[…]

### 2013-07-07 examples/count-ts-tv-methods.py
Fixed pyvcf reversion.

@@ -22,13 +22,13 @@ def count_Ts_Tv_pyvcf(vcf_file):
     """
     Ts, Tv = 0, 0
     for r in vcf.Reader(filename=vcf_file):
-        alt = r.ALT
-        ret = r.REF
-        if ref != alt and ref in bases and alt in bases:
-            if bases[ref] == bases[alt]:
+        if r.REF in bases and r.ALT[0] in bases and len(r.ALT) == 1 \
+                and r.ALT[0] != r.REF:
+            if bases[r.REF] == bases[r.ALT[0]]:
                 Ts +=1
             else:
                 Tv +=1
+
     return Ts, Tv
 
 

### 2013-07-07 wormtable.py
Added examples to tarball and bumped version.

@@ -37,7 +37,7 @@ from xml.etree import ElementTree
 
 import _wormtable
 
-__version__ = '0.1.0a1'
+__version__ = '0.1.0b1'
 SCHEMA_VERSION = "0.1"
 INDEX_METADATA_VERSION = "0.1"
 

### 2013-07-08 docs/paper/paper.tex
Put in numbers for DGRP discussion.

--- 2013-07-04 docs/paper/paper.tex
@@
 examples to a large, publicly available VCF file produced as part of the
-"Drosophila Genetic Reference Panel". This VCF ("freeze2.vcf.gz, available 
-here: ftp://ftp.hgsc.bcm.edu/DGRP/freeze2_Feb_2013/) contains variant calls 
+Drosophila genetic reference panel~\citep{mr+12}.
+This VCF\footnote{\href%
+{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}
+{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}}
+contains variant calls 
 from a population consisting 192 inbred Drosophila lines and contains data 
-from the whole genome over a total of 6,146,611 rows). This VCF was converted to 
-wormtable format using the included \texttt{vcf2wt} conversion program.  
-The (uncompressed) VCF file was 15Gb, and the converted wormtable was 12Gb,
-which required X hours.
+from the whole genome over a total of 6,146,611 rows and consists
+of 15GB of uncompressed text. The VCF was converted to 
+wormtable format using the included \texttt{vcf2wt} program
+on a standard workstation with a single hard disk. The
+conversion required approximately 70 minutes, and the size of the resulting  
+wormtable file was 10.1GB (using the smallest type required
+to represent the data in each column).
 
 Values are stored in wormtable in a binary format, so that no parsing is
-required when reading in rows whereas methods that use the VCF directly 
-must parse each row before it can be used, and is by far the most time 
-consuming part of processing VCF data. To illustrate this advantage, we
+required when reading in rows. To illustrate this advantage, we
 wrote a script to count the number of transitions and transversions
-in the dataset, using wormtable and using PyVCF[citation?], a module for
-python to parse VCF files. In both of these examples we proceed 
+in the dataset, using wormtable, standard Unix tools and PyVCF\footnote{%
+\href{http://pyvcf.rtfd.org/}{http://pyvcf.rtfd.org/}}, 
+a Python VCF parser.
+In each examples we proceed 
 row-by-row, checking the REF and ALT columns for the type of variant and
-summing transitions and transversions (ignoring indels). The 
-PyVCF method required X hours to complete, while the wormtable 
-based script needed only Y seconds. This stark difference is not a 
-criticism of the PyVCF library, which provides an efficient parser 
-for VCF files; it is rather a measure of the inherent slowness of
-parsing text in large volumes.
+counting the transitions and transversions we encounter. 
+Using PyVCF this required approximately 130 minutes. The Unix tools 
+(\texttt{cut}, \texttt{grep} and \texttt{perl}) method required 
+1m17s, and wormtable required 55 seconds. Both the Unix tools 
+and wormtable were limited by the speed of the hard disk in the
+test machine, and therefore could be much faster using 
+modern solid state 
+storage. It should be noted, however, that this simple task represents 
+something of a best case scenario for the Unix tools pipeline.
 
@@
 To illustrate the use of indexes in wormtable, we created an index 
-on the REF and ALT columns. This index required X MB of disc space 
-Y minutes to build, respectively. The REF+ALT index provides a much more 
-efficient means of counting the number of transitions and transversions 
-(we can just count the number of rows matching keys in the index 
-corresponding the SNPs of each type). A script to calculate these 
-counts by this method required x seconds.
+on the REF and ALT columns which required 3m42s to build and 
+consumed 66MB of storage space. Using this index we repeated the 
+example of counting transitions and transversions, which required 
+less than a second to complete.
 

### 2013-07-13 docs/paper/paper.tex
Good draft.

@@
 \section{Summary:}
-Biological research generates vast quantities of tabular data, which must 
-be processed to test hypotheses. These tables are produced by 
+Biological research generates large amounts of tabular data.
+These tables are produced by 
 programs and exported to plain text files, which are often very large. 
@@
 line-by-line, parsing values encoded as text into native types so that 
-calculations may be done. This is an extremely inefficient process.
+calculations may be done. This is a time consuming process.
 Furthermore, there is no simple means of indexing these files so that 
-arbitrary data values within rows can be quickly found. We introduce a new 
+rows containing particular values can be quickly found. We introduce a new 
 data format and software library called wormtable, which provides efficient 
@@
 \section{Introduction}
-% Lots of data
-The improvement in sequencing technologies has
-lead to an exponential increase in the volume of data being managed 
-by biologists. Standard studies require the storage and processing 
-of hundreds of gigabytes of data, and this volume of data is only set to 
-increase as sequencing technologies continue to improve.
+Despite the ever increasing volumes of data being processed in bioinformatics,
+the methods used are almost entirely based on plain text files. Data is 
+usually encoded in rows, with each row consisting of a series of tab-delimited 
+values.
+These files are easy to view and interpret and can be processed on 
+any platform with the minimum of library dependencies. Using compression,
+text files can be quite compact, and Tabix~\citep{li11} provides  
+fast retrieval of rows from arbitrary locations in the table.
 
-% but we use text files
-Despite the increasing importance of data processing, however, the methods 
-used in Bioinformatics are almost entirely 
-based around the use of text files. Plain text files have many advantages:
-they are extremely portable, easy to view and interpret and can be processed on 
-any platform with the minimum of library dependencies. Text files also 
-have many disadvantages however; the principle difficulties for our 
-purposes here are, they are slow to parse and difficult to index.
+It is not sufficient, however, to simply store and retrieve data. To use data
+it must be \emph{processed}. This is the major flaw in 
+using text files as a data format: before we can perform calculations, we 
+must first parse the encoded information into a native 
+machine value. Compression adds another substantial overhead. The result 
+of this is that simple calculations over a large dataset may take 
+many hours to complete.
 
-% text files are also easy to read/comprehend and don't require specialist
-% knowledge to process.
-% I think learning a database syntax is a barrier to many researchers that are
-% not trained bioinformaticians
+Another problem with tabular text files is that it is difficult 
+to index the information within partiular columns. 
+This means that most operations on the table require a complete scan 
+through the file. As a result, the only viable means of 
+working with a subset of the data
+is to create another file consisting of the subset of interest. 
+This is inflexible and error prone, and multiplies already significant  
+storage requirements.
 
-% Databases solve these problems
-The classical approach to working with large volumes of 
-structured data is to use a database, which can solve these 
-problems very effectively.
-Data is typically encoded in an efficient binary 
-format, saving substantial space and avoiding the need
-for repeated parsing. Records are usually stored in a
-B-tree, leading to fast access times even 
-over very large datasets. Indexing over columns within rows is 
-also very powerful, allowing us to quickly retrieve 
-rows with particular features.
+The obvious solution to these problems is to load tabular data into a
+relational database. Databases store values in binary form so that parsing is 
+not required, and support efficient retrieval and indexing. There 
+are many problems with this approach, however. Relational databases are 
+extremely complex systems, each supporting different features and SQL 
+dialects. It is not a straightforward task to design a schema for a 
+particular data set, particularly not if portability accross different 
+databases is required. Similiarly, accessing the data requires a knowledge
+of SQL. In the most common case, a database server must also be maintained
+and so storage and user permissons must be carefully managed. All of 
+these aspects require significant expertise.
 
-% there are a range of standard (well defined) tabular data formats: vcf, gtf
-% lends itself to a database solution geared towards specific file types
-
-% Databases but introduce several others. 
-One may, of course, load textual data into a relational database, and 
-use SQL to retrieve the records of interest. This approach has several 
-disadvantages, however. Firstly, one must first examine the data 
-and construct an appropriate SQL schema to define the appropriate columns. 
-It is sometimes feasible to do this automatically 
-for data files, but is certainly not a trivial process, 
-particularly if the schema 
-must be portable across different relational database platforms.
-Even assuming that the data 
-has been inserted into a table with an appropriate schema, there 
-are still difficulties to be addressed. Maintaining a database 
-server is a difficult task and when the volumes of data become
-large, a specialised administrator is usually required to 
-ensure the server works correctly.
-
-% We don't need concurrency, etc, as it's always read-only.
-Much of the time, a database server is far more complicated than
-we require for biological data. Data files 
-are usually written once by an application 
-and do not subsequently change. Thus, storing this information 
+A relational database server, however, is far more than we require in the 
+majority of cases. Data files are usually written once and 
+not subsequently expected to change.
+Thus, storing this information 
 in a relational database with its sophisticated concurrency control
-and guarantees of consistency in the face of asynchronous updates
-is entirely unnecessary. Interacting with a database is also a
-substantial hurdle, as another lanuage must be learned in order 
-to gain access to data. Relational databases, although providing 
-powerful data management technologies are far more complex than 
-is required, and are therefore little used for the types of 
-data that we are interested in.
-
-% are other database solutions less flexible when it comes to where the 
-% data is stored?
+is entirely unnecessary. Centralised storage of data sets
+creates an unnecessary administration burden, as does an extra
+layer of user management. Therefore, although relational databases
+provide powerful data management technologies, they are complex
+to use and maintain and are therefore rarely used for 
+static datasets.
 
@@
 be either written directly via the library, or converted
-from existing formats via command line tools. Tables can be read 
-concurrently by any number of processes, and are accessed by a 
-simple API in Python.
+from existing formats via command line tools. 
+Wormtable provides a simple Python API to access rows
+in the table, facilitating efficient and convenient access to 
+data.
+
+% Wormtable's binary row format 
+Data is stored using a  
+compact binary format designed specifically for biological applications.
+Each row consists of a fixed number of columns, and each column 
+holds either a fixed or variable number of elements of a fixed type and size.
+Wormtable currently supports signed and unsigned integers, 
+floating point and character data. To ensure that values are 
+stored as compactly as possible, integer sizes can range from  
+$1$ to $8$ bytes. The standard single and double precision 
+floating point values are also supported, along with fixed and 
+variable length strings.
 
@@
 To take advantage of the data processing capabilities of database
-technologies without the overheads and disadvantages of a relational database 
-server, wormtable uses Berkeley DB~\citep{obs99} for row storage and indexing.
-Berkeley DB is an open source embedded database toolkit, that provides a
-scalable key-value store~\citep{m12}. It is a mature and stable platform, 
+technologies without the overheads and disadvantages of relational databases,
+wormtable uses Berkeley DB~\citep{obs99} for row storage and indexing.
+Berkeley DB is an open source embedded database toolkit that provides a
+scalable key-value store. It is a mature and stable platform, 
 and is currently the most widely deployed database toolkit in the 
-world~\citep{sb12}. DB is a highly modular library, providing the  % do you mean Berkeley DB?
+world~\citep{sb12}. Berkeley DB is a highly modular library, providing the 
 means to define a very lightweight read-only database, without the 
-overheads of a server process
+overheads and complexity of a relational database system.
 
-% Wormtable's binary row format 
-On this solid foundation of row storage in Berkeley DB, wormtable uses a
-compact binary row format designed specifically for biological applications.
-Each row contains data from a fixed number of columns, and each column 
-holds either a fixed or variable number of elements of a fixed type and size.
-Supported types are signed and unsigned integers, 
-floating point and character data. To ensure that values are 
-stored as compactly as possible, wormtable supports integer values from 
-$1$ to $8$ bytes, and the standard single and double precision 
-floating point values.
-
-% Wormtable's indexes 
-An essential element of database technology is the ability to index tables
-to rows with certain properties to be quickly retrieved. Wormtable 
-supports indexing over columns or groups of columns within a table, 
-making many operations extremely efficient.
+Combining wormtable's custom row format with Berkeley DB gives us a 
+powerful platform for data storage and processing.
+An important advantage
+of this combination is the ability to index data. Wormtable supports
+indexes over arbitary combinations of columns, along with a flexibile
+system of `binning' of values in an index.
+Indexes can often be used to avoid the cost of a full table scan, and 
+can therefore make many operations much faster.
 
 \section{Examples}
-
-% What is VCF?
-A popular file format for Bioinformatics use is the Variant Call Format~\citep{da+11},
-or VCF. In a VCF file, information about variant sites in a genome is encoded 
-as tab-delimited rows in a text file. To access information about a 
-particular site naively, one must proceed through the file line-by-line until the 
-desired record is found. There are methods available to index this 
-file~\citep{li11} by genomic position, but these have no semantic 
-understanding of the structure of the file and so are limited in the 
-fields that can be indexed. One cannot, for example, build an index 
-to quickly obtain all records in which the quality field is greater
-than some value. Such queries can take a considerable time on large VCF 
-files, which discourages testing and checking for the effects of 
-arbitrary thresholds in analyses.
-
-Wormtable is designed to make working with genome scale data more 
-efficient, and in this section we describe some examples of wormtable
-that illustrate the improvements over existing methods. We applied these 
-examples to a large, publicly available VCF file produced as part of the
-Drosophila genetic reference panel~\citep{mr+12}.
+The Variant Call Format (VCF) encodes information about variant sites in a genome 
+as tab-delimited rows in a text file~\citep{da+11}.  
+To illustrate the advantages of wormtable, we
+converted a large 
+publicly available VCF file, produced as part of the
+Drosophila genetic reference panel~\citep{mr+12}, to wormtable format.
 This VCF\footnote{\href%
@@
 {ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}}
-contains variant calls 
-from a population consisting 192 inbred Drosophila lines and contains data 
-from the whole genome over a total of 6,146,611 rows and consists
-of 15GB of uncompressed text. The VCF was converted to 
-wormtable format using the included \texttt{vcf2wt} program
+consists of 15GB of uncompressed text and contains data from the 
+whole genome over 6,146,611 rows.
+Using the included \texttt{vcf2wt} program
+the VCF was converted to wormtable format 
 on a standard workstation with a single hard disk. The
-conversion required approximately 70 minutes, and the size of the resulting  
+conversion required approximately 68 minutes, and the size of the resulting  
 wormtable file was 10.1GB (using the smallest type required
@@
 
-Values are stored in wormtable in a binary format, so that no parsing is
+Values are stored in wormtable in a portable binary format, so that no parsing is
 required when reading in rows. To illustrate this advantage, we
 wrote a script to count the number of transitions and transversions
-in the dataset, using wormtable, standard Unix tools and PyVCF\footnote{%
+in the dataset using wormtable and PyVCF\footnote{%
 \href{http://pyvcf.rtfd.org/}{http://pyvcf.rtfd.org/}}, 
 a Python VCF parser.
-In each examples we proceed 
-row-by-row, checking the REF and ALT columns for the type of variant and
+In this example we proceed 
+row-by-row, examining the REF and ALT columns and
 counting the transitions and transversions we encounter. 
-Using PyVCF this required approximately 130 minutes. The Unix tools 
-(\texttt{cut}, \texttt{grep} and \texttt{perl}) method required 
-1m17s, and wormtable required 55 seconds. Both the Unix tools 
-and wormtable were limited by the speed of the hard disk in the
-test machine, and therefore could be much faster using 
-modern solid state 
-storage. It should be noted, however, that this simple task represents 
+Using PyVCF this required approximately 130 minutes;
+the wormtable version required
+55 seconds. To compare against methods that are known to be extremely
+efficient, we repeated the same 
+example using the Unix tools \texttt{cut}, \texttt{grep} 
+and \texttt{awk},  which required 77 seconds. Both the
+Unix pipeline and wormtable were limited by the sequential read
+bandwidth of the hard drive, and would therefore be much 
+faster using modern solid state storage.
+It should be noted, however, that this simple task represents 
 something of a best case scenario for the Unix tools pipeline.
 
-It is well known, of course, that plain text cannot be retrieved 
-efficiently at large scales. The BVCF format is a binary version 
-of VCF which is intended to avoid this overhead [citation?]. 
-This binary format does alleviate the problem of parsing overhead,
-but does not solve the problem of indexing \emph{within} rows.
-
-To illustrate the use of indexes in wormtable, we created an index 
-on the REF and ALT columns which required 3m42s to build and 
-consumed 66MB of storage space. Using this index we repeated the 
-example of counting transitions and transversions, which required 
-less than a second to complete.
-
-The examples in this section are intended to illustrate the 
-magnitude of performance gains that wormtable can provide 
-when working with large datasets. The gains are many: we 
-do not have to parse the information before it is used;
-we can quickly seek to arbitrary rows within a table; and 
-we can use indexes to intelligently sort and aggregate our
-data to make it more useful.
+Another major advantage of wormtable is the ability to create 
+an index over arbitary combinations of columns within a table.
+Indexes make many operations much more efficient, as they provide 
+a way to quickly find particular rows and avoid the 
+heavy cost of a full table scan.
+To illustrate this advantage, we repeated the example 
+of counting transitions and transversions using an index 
+on the REF and ALT columns.
+The index required 3m42s to build and 
+consumed 66MB of storage space. Counting the number of 
+transitions and transversions using this index required 
+less than a second.
 
 \section{Conclusions}
-% text files are good for archiving, but terrible for processing
-Text files are the oldest and most reliable means of exchanging 
-information between computers. Plain text, however, does not provide an efficient 
-means of storing and processing large volumes of data. Current practice 
-in bioinformatics is to encode data in more and more complex plain 
-text data formats. While these formats are excellent from the perspective 
-of minimising software dependencies, they are very inefficient in terms of 
-storage space, the time required to parse them, and the linear
-seek times to find particular rows.
+It is well known that parsing text files is a very inefficient process, 
+and binary file formats are preferable for this reason. The BCF format,
+for example, is a binary version of the VCF format discussed above which 
+is designed to avoid just this problem. There are significant difficulties,
+however, with having many different binary file formats for bioinformatics 
+data. Each binary format requires a library and set of tools to view 
+and process it, as it is not reasonable to expect users to decode binary
+files. Bindings for several different languages must also be provided, if 
+the file format is to be widely used.
+Maintaining these libraries, tools and language bindings across 
+different processor architectures and operating systems
+is a complex software engineering task. Maintaining this ecosystem 
+separately for several different file formats is surely 
+unsustainable.
 
-% We're not trying to define a universal bioinformatics file format,
-% just to help people be more productive.
+Wormtable alleviates the need for these different formats and libraries,
+as it is flexible enough to store many different types of data. Python 
+is currently the only programming language supported, but a C library
+and bindings for other languages are planned. Wormtable is very portable,
+and has been tested on big- and little-endian platforms with 32 and 64
+bit word sizes, along with many operating system combinations. To 
+take advantage of the advanced data processing features of wormtable
+all that is required is a conversion program, a considerably simpler
+task than designing and supporting a custom binary file format.
+
 Wormtable is not intended to replace text files as the universal 
-interchange format. It is intended to provide a persistent data structure 
-that efficiently holds data in a form that can be efficiently processed 
+interchange format for biological data. 
+It is intended to provide a persistent data structure 
+that can be efficiently processed 
 and searched. Using this data structure, researchers with no knowledge of 
@@
 data in a very high performance manner. 
-By providing conversion programs to convert 
-common formats to wormtable, we also provide an easy to 
-use common platform for 
-bioinformatics analysis. This reduces the need for third party libraries 
-to parse complex files and considerably simplifies the code required 
+Different file formats can be handled consistently 
+in wormtable, reducing the need for third party libraries 
+to parse complex files and simplifying the code required 
 to process data.

### 2013-07-13 docs/paper/paper.tex
Final revision submitted for comments.

@@
 the methods used are almost entirely based on plain text files. Data is 
-usually encoded in rows, with each row consisting of a series of tab-delimited 
+usually encoded in lines of text, with each row consisting of a series of tab-delimited 
 values.
@@
 using text files as a data format: before we can perform calculations, we 
-must first parse the encoded information into a native 
-machine value. Compression adds another substantial overhead. The result 
+must first parse the encoded information into native 
+machine values. This is an expensive process, and 
+compression (if it is used) adds another substantial overhead. The result 
 of this is that simple calculations over a large dataset may take 
@@
 
-Another problem with tabular text files is that it is difficult 
-to index the information within partiular columns. 
+Another problem with tables stored text files is that it is difficult 
+to index the information within particular columns. 
 This means that most operations on the table require a complete scan 
@@
 dialects. It is not a straightforward task to design a schema for a 
-particular data set, particularly not if portability accross different 
-databases is required. Similiarly, accessing the data requires a knowledge
+particular data set, particularly not if portability across different 
+databases is required. Similarly, accessing the data requires a knowledge
 of SQL. In the most common case, a database server must also be maintained
-and so storage and user permissons must be carefully managed. All of 
+and so storage and user permissions must be carefully managed. All of 
 these aspects require significant expertise.
@@
 
-Combining wormtable's custom row format with Berkeley DB gives us a 
-powerful platform for data storage and processing.
-An important advantage
-of this combination is the ability to index data. Wormtable supports
-indexes over arbitary combinations of columns, along with a flexibile
-system of `binning' of values in an index.
-Indexes can often be used to avoid the cost of a full table scan, and 
-can therefore make many operations much faster.
+In wormtable, each table corresponds to a directory in the file system
+which is used to store the table, indexes and metadata.
+Indexes over arbitrary combinations of columns are supported, 
+and these are managed using the \texttt{wtadmin} program, which 
+provides several other administration utilities.
+Wormtable files are portable across operating systems 
+and processor architectures.
 
@@
 Another major advantage of wormtable is the ability to create 
-an index over arbitary combinations of columns within a table.
+an index over arbitrary combinations of columns within a table.
 Indexes make many operations much more efficient, as they provide 
@@
 is currently the only programming language supported, but a C library
-and bindings for other languages are planned. Wormtable is very portable,
+and bindings for other languages are planned. Wormtable is portable,
 and has been tested on big- and little-endian platforms with 32 and 64
@@
 database systems can take full advantage of sophisticated 
-data management techniques, and write simple code that processes this 
-data in a very high performance manner. 
+data management techniques, and write simple code to process 
+data in a high performance manner. 
 Different file formats can be handled consistently 

### 2013-07-23 _wormtablemodule.c
Improved floating point tests and docs.

@@ -157,6 +157,14 @@ byteswap_copy(void* dest, void *source, size_t n)
 }
 #endif
 
+/*
+ * Floating point packing and unpacking. Floating point values are stored as a
+ * slight perturbation to the IEEE standards so that they sort correctly. This 
+ * is done by a simple rule: if the number is positive, flip the sign bit; if
+ * it is negative, flip all bits. This procedure is reversed when we unpack 
+ * to recover the original values exactly.
+ */
+
 static void 
 pack_half(double value, void *dest)
 {
@@ -352,19 +360,21 @@ min_uint(u_int32_t k)
 static uint64_t 
 missing_float(u_int32_t k) 
 {
-    uint64_t ret = 0;
+    uint64_t zero = 0uLL;
+    union { double value; uint64_t bits; } conv;
+    conv.bits = 0;
     switch (k) {
         case 2:
-            ret = 0xfffffc0000000000uLL;
+            conv.value = unpack_half(&zero);
             break;
         case 4:
-            ret = 0xffffffffe0000000uLL;
+            conv.value = unpack_float(&zero);
             break;
         case 8:
-            ret = 0xffffffffffffffffuLL;
+            conv.value = unpack_double(&zero);
             break;
     }
-    return ret;
+    return conv.bits;
 }
 
 

### 2013-07-23 test/highlevel.py

@@ -487,25 +487,33 @@ class FloatTest(WormtableTest):
     Tests the limits of the floating point types to see if they are correct
     under IEEE rules.
     """
+    # Maximum representable numbers
     max_half = (2 - 2**-10) * 2**15
     max_float = (2 - 2**-23) * 2**127
     max_double = (1 + (1 - 2**-52)) * 2**1023
-    
-    min_half_normal = 2**-24 
-    min_half_denormal = 2**-14
+    # Minimum values 
+    min_half_normal = 2**-14 
+    min_half_denormal = 2**-24
     min_float_normal = 2**-126
     min_float_denormal = 2**-149
     min_double_normal = 2**-1022
     min_double_denormal = 2**-1074
+    # The number of decimal digits of precision
+    max_half_digits = 3
+    max_float_digits = 6
+    max_double_digits = 15
 
-
-    def assert_tables_equal(self, in_values, out_values):
+    def create_table(self):
         t = wt.Table(self._homedir)
         t.add_id_column()
         t.add_float_column("half", size=2)
         t.add_float_column("single", size=4)
         t.add_float_column("double", size=8)
         t.open("w")
+        return t 
+
+    def assert_tables_equal(self, in_values, out_values):
+        t = self.create_table() 
         for r in in_values:
             t.append([None] + list(r)) 
         t.close()
@@ -523,10 +531,9 @@ class FloatTest(WormtableTest):
         results = [[None, None, None] for v in values]
         self.assert_tables_equal(values, results)
 
-
     def test_exact_values(self):
         """
-        Checks that exact values are added stored correctly.
+        Checks that exact values are stored correctly.
         """
         # Get a bunch of small integers
         in_values = []
@@ -558,7 +565,6 @@ class FloatTest(WormtableTest):
                     self.min_double_denormal]]
         in_values.extend(v) 
         self.assert_tables_equal(in_values, in_values)
-      
     
     def test_infinity(self):
         """
@@ -577,7 +583,6 @@ class FloatTest(WormtableTest):
         neg = [[-1 * v for v in x] for x in values]
         neg_inf = [[-inf for v in x] for x in values]
         self.assert_tables_equal(neg, neg_inf)
-
     
     def test_half_infinity(self):
         inf = 1e1000
@@ -599,6 +604,64 @@ class FloatTest(WormtableTest):
         y = [(-v, None, None) for v in values]
         self.assert_tables_equal(y, y)
 
+    def test_underflow(self):
+        """
[…]

### 2013-07-23 test/lowlevel.py

@@ -615,28 +615,36 @@ class TestDatabaseFloat(TestDatabase):
         cols = self._columns
         self.num_rows = num_rows
         self.rows = []
+        # either generate random ints or exactly representable fractions
+        # to avoid rounding issues with 2 and 4 byte floats
+        def generate_int():
+            min_v, max_v = -20, 20 
+            return random.randint(min_v, max_v) 
+        def generate_fraction():
+            denoms = [2**j for j in range(2, 10)]
+            denoms += [-v for v in denoms]
+            return 1.0 / denoms[random.randint(0, len(denoms) - 1)]
         for j in range(self.num_rows):
             row = [0 for c in cols]
             for k in range(1, self.num_columns): 
                 c = cols[k]
-                min_v, max_v = -10, 10 
+                f = generate_int 
+                if random.random() < 0.5:
+                    f = generate_fraction 
                 if c.num_elements == 1:
-                    row[k] = random.uniform(min_v, max_v) 
+                    row[k] = f() 
                 else:
                     n = c.num_elements
                     if n == _wormtable.WT_VAR_1:
                         n = random.randint(1, _wormtable.MAX_NUM_ELEMENTS)
-                    v = tuple([random.uniform(min_v, max_v) for l in range(n)])
-                    row[k] = v
+                    row[k] = tuple([f() for l in range(n)])
                 if j % 2 == 0:
                     rb.insert_elements(k, row[k]) 
                 else:
-                    # Use the exact hex encoding for floats to avoid loss of
-                    # precision by converting to base 10.
                     if c.num_elements == 1:
-                        s = row[k].hex()
+                        s = str(row[k])
                     else:
-                        s = ",".join(f.hex() for f in row[k])
+                        s = ",".join(str(x) for x in row[k])
                     rb.insert_encoded_elements(k, s.encode())
             rb.commit_row()
             self.rows.append(tuple(row))
@@ -838,9 +846,7 @@ class TestIndexIntegrity(object):
             # get the list from the original rows
             l3 = [row[j] for row in self.rows]
             l3.sort()
-            # these tests don't make sense for inexact types. 
-            if not (col.element_type == _wormtable.WT_FLOAT and col.element_size != 8):
-                self.assertEqual(l, l3)
+            self.assertEqual(l, l3)
             index.close()
             del row_iter
          self.destroy_indexes()
[…]

### 2013-07-23 docs/paper/paper.tex

@@
 \title[Wormtable]{Wormtable: A data structure for genome scale data}
-\author[Kelleher \textit{et~al}]{Jerome Kelleher, Robert W. Ness 
+\author[Kelleher \textit{et~al}]{Jerome Kelleher, Rob W. Ness 
 and Daniel L. Halligan}
@@
 wormtable uses Berkeley DB~\citep{obs99} for row storage and indexing.
-Berkeley DB is an open source embedded database toolkit that provides a
+Berkeley DB is a free and open-source embedded database toolkit that provides a
 scalable key-value store. It is a mature and stable platform, 
@@
 counting the transitions and transversions we encounter. 
-Using PyVCF this required approximately 130 minutes;
+Using PyVCF this required approximately 126 minutes;
 the wormtable version required
@@
 \section*{Acknowledgement}
+We would like to thank the Ashworth Codemonkeys for valuable feedback on earlier
+versions of wormtable.
 \paragraph{Funding\textcolon}  
-JK is supported by EPSRC grant EP/I013091/1; DLH is supported by XXXXX; 
-and RWN is supported by XXXX.
+JK is supported by EPSRC grant EP/I013091/1.
 

### 2013-07-24 dev.py
Removed dev.py file and bumped to beta 2.

@@ -1,79 +1,0 @@
-"""
-This is an example file get started with building a wormtable for 
-a VCF file, dumping out the values, and reading indexes.
-"""
-from __future__ import print_function
-from __future__ import division 
-
-import sys
-
-import numpy as np
-import wormtable as wt
-
-def read_position_index(homedir, chromosome, position):
-    """
-    Read the index on CHROM and POS, which is assumed to already exist.
-    """
-    with wt.open_table(homedir) as t, t.open_index("CHROM+POS") as i:
-        cursor = t.cursor(["CHROM", "POS", "FILTER"], i)
-        cursor.set_min(chromosome, position)
-        for row in cursor:
-            print(row)
-
-def read_filter_index(homedir):
-    """
-    Example demonstrating the use of the counter method on indexes
-    """
-    with wt.open_table(homedir) as t, t.open_index("FILTER") as i:
-        c = i.counter()
-        total = 0
-        for k, v in c.items():
-            print(v, "\t", k)
-            total += v
-        assert(len(t) == total)
-
-def min_max_example(homedir):
-    """
-    A straightforward example of using index.get_min and get_max.
-    """
-    name = "CHROM+POS+QUAL[10]"
-    with wt.open_table(homedir) as t, t.open_index(name) as i:
-    
-        print("whole index: ")
-        print("\t", i.get_min(), "\t", i.get_max())
-        # Now, let's get the min and max of the two chromosomes
-        print("Individual chromosomes:")
-        print("\t", i.get_min('1'), "\t", i.get_max('1'))
-        print("\t", i.get_min('Y'), "\t", i.get_max('Y'))
-        # This can be extended indefinitely, matching as many columns 
-        # as are in the index
-        print("\t", i.get_min('1', 3000055), "\t", i.get_max('1', 3000555))
-
-
-def dump_table(homedir):
-    """
-    Dumps the wormtable in the specified directory to stdout row-by-row.
-    
-    This uses the get_row(j) function, which returns the jth row of the 
-    table as a dictionary mapping column name to values.
-    """
-    with wt.open_table(homedir) as t:
-        r = 0
-        for row in t: 
-            r += 1
-            print(row)
-        print("num rows = ", len(t), r)
-    
-def main():
-    homedir = sys.argv[1]
-    # Uncomment this is you want to just dump out the whole table in 
-    # the native order.
-    # dump_table(homedir)
-    
[…]

### 2013-07-24 wormtable.py

@@ -37,7 +37,7 @@ from xml.etree import ElementTree
 
 import _wormtable
 
-__version__ = '0.1.0b1'
+__version__ = '0.1.0b2'
 SCHEMA_VERSION = "0.1"
 INDEX_METADATA_VERSION = "0.1"
 

### 2013-07-24 _wormtablemodule.c
Initial work to support table stats and change XML.
This adds a non-backward compatible change to the XML
metadata format for tables. This is unfortunate but necessary
as the existing setup was not extensible, as we didn't have a
table root element.

Also changed the types in C api to be consistent with the Python
C api.

@@ -66,20 +66,29 @@ typedef struct Column_t {
     PyObject *(*native_to_python)(struct Column_t *, int);
 } Column;
 
+/*
+ * We're using a mixture of fixed and non-fixed size integer types because 
+ * of the conflicting needs of being exact about the sizes and interfacing 
+ * with Python. Ideally, all of the types would be fixed size for simplicity
+ */
 
 typedef struct {
     PyObject_HEAD
     DB *db;
     PyObject *filename;
     Column **columns;
-    Py_ssize_t cache_size;
-    u_int32_t page_size;
-    u_int32_t num_columns;
-    u_int32_t fixed_region_size;
+    unsigned long long cache_size;
+    unsigned int page_size;
+    unsigned int fixed_region_size;
+    unsigned int num_columns;
     void *row_buffer;
     u_int32_t row_buffer_size;     /* max size */
     u_int32_t current_row_size;    /* current size */
-    u_int64_t current_row_id;      
+    unsigned long long num_rows;     
+    /* row stats */
+    unsigned long long total_row_size;
+    unsigned int min_row_size;
+    unsigned int max_row_size;
 } Table;
 
 
@@ -88,7 +97,7 @@ typedef struct {
     Table *table;
     DB *db;
     PyObject *filename;
-    Py_ssize_t cache_size;
+    unsigned long long cache_size;
     u_int32_t *columns;
     u_int32_t num_columns;
     void *key_buffer;
@@ -1912,7 +1921,10 @@ Table_init(Table *self, PyObject *args, PyObject *kwds)
         }
     }
     self->current_row_size = self->fixed_region_size;
-    self->current_row_id = 0; 
+    self->num_rows = 0; 
+    self->max_row_size = 0;
+    self->min_row_size = MAX_ROW_SIZE;
+    self->total_row_size = 0;
     ret = 0;
 out:
     return ret;
@@ -1920,9 +1932,13 @@ out:
 
 static PyMemberDef Table_members[] = {
     {"filename", T_OBJECT_EX, offsetof(Table, filename), READONLY, "filename"},
-    {"cache_size", T_PYSSIZET, offsetof(Table, cache_size), READONLY, "cache_size"},
-    {"page_size", T_PYSSIZET, offsetof(Table, page_size), READONLY, "page_size"},
[…]

### 2013-07-24 wormtable.py

@@ -38,7 +38,7 @@ from xml.etree import ElementTree
 import _wormtable
 
 __version__ = '0.1.0b2'
-SCHEMA_VERSION = "0.1"
+TABLE_METADATA_VERSION = "0.1"
 INDEX_METADATA_VERSION = "0.1"
 
 DEFAULT_CACHE_SIZE = 16 * 2**20 # 16M 
@@ -470,8 +470,29 @@ class Table(Database):
         Database.__init__(self, homedir, self.DB_NAME)
         self.__columns = []
         self.__column_name_map = {}
-        self.__num_rows = 0
         self.__page_size = self.DEFAULT_PAGE_SIZE
+        self.__num_rows = 0
+        self.__total_row_size = 0
+        self.__min_row_size = 0 
+        self.__max_row_size = 0
+        
+    def get_total_row_size(self):
+        """
+        Returns the total number of bytes stored within rows in this table.
+        """
+        return self.__total_row_size
+
+    def get_min_row_size(self):
+        """
+        Returns the size of the smallest row in this table in bytes.
+        """
+        return self.__min_row_size
+
+    def get_max_row_size(self):
+        """
+        Returns the size of the largest row in this table in bytes.
+        """
+        return self.__max_row_size
 
     def set_page_size(self, page_size):
         """
@@ -588,38 +609,101 @@ class Table(Database):
             raise TypeError("column ids must be strings or integers")
         return ret
     
+    
+    def __generate_schema_xml(self):
+        """
+        Generates the XML representing the schema for this table.
+        """
+        schema = ElementTree.Element("schema")
+        columns = ElementTree.Element("columns")
+        schema.append(columns)
+        for c in self.__columns:
+            columns.append(c.get_xml())
+        return schema
+   
+    def __generate_stats_xml(self):
+        """
+        Generates the XML representing the statistics for this table.
+        """
+        stats = ElementTree.Element("stats")
+        l = [ 
+            ("num_rows", self.__num_rows), 
+            ("max_row_size", self.__max_row_size),
+            ("min_row_size", self.__min_row_size),
+            ("total_row_size", self.__total_row_size)
+        ] 
+        for name, value in l:
+            d = {"name":name, "value":str(value)}
+            s = ElementTree.Element("stat", d)
+            stats.append(s)
+        return stats
+
     def get_metadata(self):
         """
         Returns an ElementTree instance describing the metadata for this 
         Table.
         """
[…]

### 2013-07-24 _wormtablemodule.c
Fixed bug in reading integer types.

@@ -1884,7 +1884,7 @@ Table_init(Table *self, PyObject *args, PyObject *kwds)
      * an optional setter. There should be a proper getter method 
      * also which queries DB */
     self->page_size = 64 * 1024;
-    if (!PyArg_ParseTupleAndKeywords(args, kwds, "O!O!n|n", kwlist, 
+    if (!PyArg_ParseTupleAndKeywords(args, kwds, "O!O!K|I", kwlist, 
             &PyBytes_Type, &filename, 
             &PyList_Type,  &columns, 
             &self->cache_size, &self->page_size)) {
@@ -2469,7 +2469,7 @@ Index_init(Index *self, PyObject *args, PyObject *kwds)
     self->key_buffer = NULL; 
     self->row_buffer = NULL; 
     self->columns = NULL;
-    if (!PyArg_ParseTupleAndKeywords(args, kwds, "O!O!O!n", kwlist, 
+    if (!PyArg_ParseTupleAndKeywords(args, kwds, "O!O!O!K", kwlist, 
             &TableType, &table, 
             &PyBytes_Type, &filename, 
             &PyList_Type,  &columns, 

### 2013-07-25 _wormtablemodule.c
Improved XML schema and metadata handling.

@@ -57,7 +57,7 @@ typedef struct Column_t {
     void **input_elements; /* pointer to each elements in input format */
     void *element_buffer; /* parsed input elements in native CPU format */
     int num_buffered_elements;
-    int (*string_to_native)(struct Column_t*, char *);  /* DEPRECATED */
+    int (*string_to_native)(struct Column_t*, char *);  
     int (*python_to_native)(struct Column_t*, PyObject *);
     int (*verify_elements)(struct Column_t*);
     int (*truncate_elements)(struct Column_t*, double);
@@ -263,6 +263,10 @@ unpack_double(void *src)
     return conv.value;
 }
 
+
+/* Integer packing and unpacking.
+ * TODO document the format and create pack functions.
+ */
 
 static u_int64_t
 unpack_uint(void *src, u_int8_t size) 

### 2013-07-25 wormtable.py

@@ -356,19 +356,22 @@ class Database(object):
         """
         tree = self.get_metadata()
         root = tree.getroot()
+        s = "Do not edit this file!"
+        comment = ElementTree.Comment(s)
+        root.insert(0, comment)
         raw_xml = ElementTree.tostring(root, 'utf-8')
         reparsed = minidom.parseString(raw_xml)
         pretty = reparsed.toprettyxml(indent="  ")
         with open(filename, "w") as f:
             f.write(pretty)
    
-    def read_metadata(self, filename):
+    def read_metadata(self):
         """
-        Reads metadata for this database from the specified filename,
+        Reads metadata for this database from the metadata file 
         and calls set_metadata with the result.
         """
-        tree = ElementTree.parse(filename)
-        self.set_metadata(tree)
+        tree = ElementTree.parse(self.get_metadata_path())
+        self.set_metadata(tree) 
 
     def finalise_build(self):
         """
@@ -410,7 +413,7 @@ class Database(object):
         if m == WT_WRITE:
             path = self.get_db_build_path()
         else:
-            self.read_metadata(self.get_metadata_path())
+            self.read_metadata()
         llo = self._create_ll_object(path)
         llo.open(m)
         self.__ll_object = llo
@@ -609,8 +612,39 @@ class Table(Database):
             raise TypeError("column ids must be strings or integers")
         return ret
     
+    def read_schema(self, filename):
+        """
+        Reads the schema from the specified file and sets up the columns 
+        in this table accordingly.
+        """
+        tree = ElementTree.parse(filename)
+        root = tree.getroot()
+        if root.tag != "schema":
+            raise ValueError("root element must be <schema>")
+        version = root.get("version")
+        if version is None:
+            raise ValueError("invalid xml: schema version missing")
+        supported_versions = [TABLE_METADATA_VERSION]
+        if version not in supported_versions: 
+            raise ValueError("Unsupported schema version.")
+        self._parse_schema_xml(root)
     
-    def __generate_schema_xml(self):
+    def write_schema(self, filename):
+        """
+        Writes the schema for this table to the specified file in XML format.
+        """
+        root = self._generate_schema_xml()
+        s = "Edit this candidate schema to suit your needs." 
+        comment = ElementTree.Comment(s)
+        root.insert(0, comment)
+        root.set("version", TABLE_METADATA_VERSION)
+        raw_xml = ElementTree.tostring(root, 'utf-8')
[…]

### 2013-07-28 _wormtablemodule.c
Added buffering.

@@ -2142,8 +2142,11 @@ Table_open(Table* self, PyObject *args)
         handle_io_error();
         goto out;
     }
-
-
+    if (setvbuf(self->data_file, NULL, _IOFBF, 1024 * 1024) != 0) {
+        handle_io_error();
+        goto out;
+    }
+    
     Py_INCREF(Py_None);
     ret = Py_None;
 out:

### 2013-07-28 _wormtablemodule.c
Tidied up data file handling and buffers.
Decided against attempting a proper implementation at this point. It
works very well as it is, and there are other elements of the design
that need reworking too. Not currently a priority.

@@ -38,6 +38,8 @@
 #define MAX_NUM_ELEMENTS 255
 #define MAX_ROW_SIZE 65536
 
+#define OFFSET_LEN_RECORD_SIZE 10 
+
 #define MODULE_DOC \
 "Low level Berkeley DB interface for wormtable"
 
@@ -104,11 +106,6 @@ typedef struct {
     u_int32_t num_columns;
     void *key_buffer;
     u_int32_t key_buffer_size;
-    /* this buffer isn't really necessary - we should just use the table's
-     * row buffer.
-     */
-    void *row_buffer;
-    u_int32_t row_buffer_size;
     double *bin_widths;
 } Index;
 
@@ -2274,10 +2271,15 @@ Table_update_row_stats(Table *self, u_int32_t row_size)
  * DB has filled in the offset and length of the row into the start of
  * the row buffer and now we must read this row from the data file.
  *
- * TODO this is an ugly quick hack and should be fixed! 
+ * TODO this is not a very nice way of doing things, as there is all 
+ * sorts of coupling between the different objects. A much cleaner 
+ * approach would be to have a seperate row_buffer object that 
+ * is responsible for all of this that is explicity shared by all 
+ * of the objects and that interact with it through a fixed 
+ * interface. 
  */
 static int    
-Table_retrieve_row(Table *self, void *buffer) 
+Table_retrieve_row(Table *self) 
 {
     int ret = -1;
     void *v;
@@ -2285,18 +2287,16 @@ Table_retrieve_row(Table *self, void *buffer)
     u_int32_t key_size = id_col->element_size; 
     u_int64_t offset = 0;
     u_int16_t len = 0;
-    
-    v = buffer + key_size;
+    v = self->row_buffer + key_size;
     offset = unpack_uint(v, sizeof(offset));
     v += sizeof(offset);
     len = unpack_uint(v, sizeof(len));
-    //printf("read: %lu\t%u\n", offset, len);
     /* Now read this record from the file and put it in the row buffer */
     if (fseeko(self->data_file, (off_t) offset, SEEK_SET) != 0) {
         handle_io_error();
         goto out;
     }
-    v = buffer + key_size;
+    v = self->row_buffer + key_size;
     if (fread(v, len, 1, self->data_file) != 1) {
         handle_io_error();
         goto out;
@@ -2313,10 +2313,10 @@ Table_commit_row(Table* self)
[…]

### 2013-07-28 _wormtablemodule.c
Removed page size as it's no longer an issue.

@@ -82,7 +82,6 @@ typedef struct {
     PyObject *data_filename;
     Column **columns;
     unsigned long long cache_size;
-    unsigned int page_size;
     unsigned int fixed_region_size;
     unsigned int num_columns;
     void *row_buffer;
@@ -1894,7 +1893,7 @@ Table_init(Table *self, PyObject *args, PyObject *kwds)
 {
     int ret = -1;
     static char *kwlist[] = {"db_filename", "data_filename", "columns", 
-            "cache_size", "page_size", NULL}; 
+            "cache_size", NULL}; 
     Column *col;
     PyObject *db_filename = NULL;
     PyObject *data_filename = NULL;
@@ -1905,15 +1904,11 @@ Table_init(Table *self, PyObject *args, PyObject *kwds)
     self->columns = NULL;
     self->db_filename = NULL;
     self->cache_size = 0;
-    /* TODO: take page size out of this initialiser and make it 
-     * an optional setter. There should be a proper getter method 
-     * also which queries DB */
-    self->page_size = 64 * 1024;
-    if (!PyArg_ParseTupleAndKeywords(args, kwds, "O!O!O!K|I", kwlist, 
+    if (!PyArg_ParseTupleAndKeywords(args, kwds, "O!O!O!K", kwlist, 
             &PyBytes_Type, &db_filename, 
             &PyBytes_Type, &data_filename, 
             &PyList_Type,  &columns, 
-            &self->cache_size, &self->page_size)) {
+            &self->cache_size)) {
         goto out;
     }
     self->db_filename = db_filename;
@@ -1961,7 +1956,6 @@ out:
 static PyMemberDef Table_members[] = {
     {"db_filename", T_OBJECT_EX, offsetof(Table, db_filename), READONLY, "db_filename"},
     {"cache_size", T_ULONGLONG, offsetof(Table, cache_size), READONLY, "cache_size"},
-    {"page_size", T_UINT, offsetof(Table, page_size), READONLY, "page_size"},
     {"num_rows", T_ULONGLONG, offsetof(Table, num_rows), READONLY, "num_rows"},
     {"total_row_size", T_ULONGLONG, offsetof(Table, total_row_size), READONLY, "total_row_size"},
     {"min_row_size", T_UINT, offsetof(Table, min_row_size), READONLY, "min_row_size"},
[…]

### 2013-07-28 wormtable.py

@@ -467,13 +467,11 @@ class Table(Database):
     """
     DB_NAME = "table"
     PRIMARY_KEY_NAME = "row_id"
-    DEFAULT_PAGE_SIZE = 64 * 1024
 
     def __init__(self, homedir):
         Database.__init__(self, homedir, self.DB_NAME)
         self.__columns = []
         self.__column_name_map = {}
-        self.__page_size = self.DEFAULT_PAGE_SIZE
         self.__num_rows = 0
         self.__total_row_size = 0
         self.__min_row_size = 0 
@@ -504,15 +502,6 @@ class Table(Database):
         """
         return self.__max_row_size
 
-    def set_page_size(self, page_size):
-        """
-        Sets the Berkeley DB page size to the specified value.
-
-        This only has an effect when a table is first created; afterwards,
-        the page size for a table is fixed.
-        """
-        self.__page_size = page_size
-
     def _create_ll_object(self, path):
         """
         Returns a new instance of _wormtable.Table using the specified 
@@ -522,7 +511,7 @@ class Table(Database):
         data_file = filename + b".dat"
         ll_cols = [c.get_ll_object() for c in self.__columns]
         t = _wormtable.Table(filename, data_file, ll_cols, 
-                self.get_cache_size(), self.__page_size)
+                self.get_cache_size())
         return t
 
     def get_fixed_region_size(self):

### 2013-07-28 test/utilities.py
Fixed the stdin handling and tests.

@@ -29,6 +29,8 @@ import tempfile
 import shutil
 import os.path
 import gzip
+import sys
+import io
 from xml.etree import ElementTree
 
 EXAMPLE_VCF ="test/data/example.vcf"
@@ -52,12 +54,26 @@ class UtilityTest(unittest.TestCase):
     def tearDown(self):
         shutil.rmtree(self._homedir)
 
-    def run_command(self, options=[]):
+    def run_command(self, options=[], stdin=None):
         """
         Runs the command and stores the output and exit status.
         """
-        self.get_program()(options)
-      
+        if stdin is not None:
+            sys.stdin = stdin
+        try:
+            # Ugly workaround for Python2/3 behaviour
+            if sys.version_info[0] == 2:
+                import StringIO
+                sys.stdout = StringIO.StringIO()
+            else:
+                sys.stdout = io.StringIO()
+            self.get_program()(options)
+            ret = sys.stdout.getvalue()
+        finally: 
+            sys.stdin = sys.__stdin__ 
+            sys.stdout = sys.__stdout__
+        return ret
+
     def assert_tables_equal(self, t1, t2):
         """
         Compare the specified tables row-by-row and verify that they are 
@@ -149,12 +165,26 @@ class TestInputMethods(Vcf2wtTest):
         shutil.rmtree(self._homedir)
         os.mkdir(self._homedir)
 
+    def __test_stdin_input(self, input_file):
+        to = os.path.join(self._homedir, "original")
+        ts = os.path.join(self._homedir, "stdin")
+        self.run_command([input_file, to, "-q"]) 
+        with open(input_file, "rb") as f:
+            s = self.run_command(["-", ts], stdin=f) 
+        with wt.open_table(to) as t1:
+            with wt.open_table(ts) as t2:
+                self.assert_tables_equal(t1, t2)
+        shutil.rmtree(self._homedir)
+        os.mkdir(self._homedir)
+
+
     def test_gzip(self):
         self.__test_gzipped_input(EXAMPLE_VCF)
         self.__test_gzipped_input(SAMPLE_VCF)
     
     def test_stdin(self):
-        self.assertTrue(False, "not implemented yet")
+        self.__test_stdin_input(EXAMPLE_VCF) 
+        self.__test_stdin_input(SAMPLE_VCF) 
 
 
 class TestSchemaGeneration(Vcf2wtTest):

### 2013-07-28 _wormtablemodule.c
Tidied high level code and improved test coverage.

@@ -1955,6 +1955,7 @@ out:
 
 static PyMemberDef Table_members[] = {
     {"db_filename", T_OBJECT_EX, offsetof(Table, db_filename), READONLY, "db_filename"},
+    {"data_filename", T_OBJECT_EX, offsetof(Table, data_filename), READONLY, "data_filename"},
     {"cache_size", T_ULONGLONG, offsetof(Table, cache_size), READONLY, "cache_size"},
     {"num_rows", T_ULONGLONG, offsetof(Table, num_rows), READONLY, "num_rows"},
     {"total_row_size", T_ULONGLONG, offsetof(Table, total_row_size), READONLY, "total_row_size"},

### 2013-07-28 test/lowlevel.py

@@ -1143,15 +1143,15 @@ class TestTable(unittest.TestCase):
     def setUp(self):
         fd, self._db_file = tempfile.mkstemp("-test.db", prefix=TEMPFILE_PREFIX) 
         os.close(fd)
+        fd, self._data_file = tempfile.mkstemp("-dat.db", prefix=TEMPFILE_PREFIX) 
+        os.close(fd)
     
     def tearDown(self):
         os.unlink(self._db_file)
+        os.unlink(self._data_file)
 
 
-#class TestTableInitialisation(TestTable):
-# TODO FIXME
-# This test was disabled as part of the data table hack. Update and fix!
-class DisabledTest(object):
+class TestTableInitialisation(TestTable):
     """ 
     Tests the initialisation code for the Table class to make sure everything 
     is checked correctly.
@@ -1160,25 +1160,26 @@ class DisabledTest(object):
         t = _wormtable.Table
         c0 = get_uint_column(1, 1)
         c1 = get_uint_column(1, 1)
-        f = b"file.db"
+        f1 = b"file.db"
+        f2 = b"file.dat"
         # check number of columns.
-        self.assertRaises(ValueError, t, f, [], 0) 
-        self.assertRaises(ValueError, t, f, [c0], 0) 
+        self.assertRaises(ValueError, t, f1, f2, [], 0) 
+        self.assertRaises(ValueError, t, f1, f2, [c0], 0) 
         # Make sure the row_id column is of the right type.
         for j in range(2, 5):
-            self.assertRaises(ValueError, t, f, [get_uint_column(1, j), c0], 0) 
+            self.assertRaises(ValueError, t, f1, f2, [get_uint_column(1, j), c0], 0) 
         for j in range(5):
-            self.assertRaises(ValueError, t, f, [get_int_column(1, j), c0], 0) 
-            self.assertRaises(ValueError, t, f, [get_float_column(4, j), c0], 0) 
-            self.assertRaises(ValueError, t, f, [get_char_column(j), c0], 0) 
+            self.assertRaises(ValueError, t, f1, f2, [get_int_column(1, j), c0], 0) 
+            self.assertRaises(ValueError, t, f1, f2, [get_float_column(4, j), c0], 0) 
+            self.assertRaises(ValueError, t, f1, f2, [get_char_column(j), c0], 0) 
         # check for duplicate columns
         n = 10
[…]

### 2013-07-28 test/utilities.py

@@ -209,27 +209,102 @@ class TestSchemaGeneration(Vcf2wtTest):
         self.__test_schema_generator(EXAMPLE_VCF)
         self.__test_schema_generator(SAMPLE_VCF)
 
-
-class TestSchemaBuild(VcfBuildTest):
+class WtadminTest(UtilityTest):
     """
-    Test the processing of schema files.
+    Class for testing wtadmin 
     """
 
     def setUp(self):
-        super(TestSchemaBuild, self).setUp()
-        # TODO FINISH
-        # Now, alter the schema
-        r = open(schema, "r")
-        w = open(alt_schema, "w") 
-        for l in r:
-            # remove CHROM and POS
-            if CHROM not in l and POS not in l:
-                w.write(l)
-        r.close()
-        w.close()
-        self.run_command([input_file, table , "-qs" + alt_schema]) 
-        with wt.open_table(table) as t:
-            self.assertRaises(KeyError, t.get_column, CHROM)
-            self.assertRaises(KeyError, t.get_column, POS)
+        super(WtadminTest, self).setUp()
+        vcf2wt.main([EXAMPLE_VCF, self._homedir, "-fq"])
+        self._table = wt.open_table(self._homedir)
+
+    def tearDown(self):
+        self._table.close()
+        super(WtadminTest, self).tearDown()
+
+    def run_dump(self, args=[]):
+        return self.run_command(["dump", self._homedir] + args) 
+
+    def run_add(self, args=[]):
+        return self.run_command(["add", self._homedir] + args) 
+    
+    def run_hist(self, args=[]):
+        return self.run_command(["hist", self._homedir] + args) 
+        
+    def get_program(self):
+        return wtadmin.main
+
+    def test_dump_all(self):
+        s = self.run_dump()         
+        n = 0
+        for line, r in zip(s.splitlines(), self._table):
+            n += 1
+            l = line.split()
+            # check the first few columns
+            self.assertEqual(int(l[0]), r[0])
+            self.assertEqual(l[1].encode(), r[1])
+            self.assertEqual(int(l[2]), r[2])
+        self.assertEqual(n, len(self._table))
+    
+    def test_dump_cols(self):
+        cols = ["CHROM", "REF", "ALT"]
+        s = self.run_dump(cols)         
+        c = self._table.cursor(cols)
+        n = 0
+        for line, r in zip(s.splitlines(), c): 
+            n += 1
+            l = line.split()
+            for u, v in zip(l, r):
+                self.assertEqual(u.encode(), v)
[…]

### 2013-07-28 wormtable.py

@@ -262,7 +262,7 @@ class Database(object):
         """
         return self.__ll_object
 
-    def _create_ll_object(self, path):
+    def _create_ll_object(self, build):
         """
         Returns a newly created instance of the low-level object that this 
         Database is a facade for.
@@ -306,7 +306,8 @@ class Database(object):
         """
         Returns the path of the file used to build the database.
         """
-        s = "_build_{0}_{1}.db".format(os.getpid(), self.get_db_name()) 
+        s = "_build_{0}_{1}{2}".format(os.getpid(), self.get_db_name(),
+                self.DB_SUFFIX) 
         return os.path.join(self.get_homedir(), s) 
 
     def get_db_file_size(self):
@@ -409,12 +410,12 @@ class Database(object):
         m = modes[mode]
         self.__open_mode = None
         self.__ll_object = None
-        path = self.get_db_path()
+        build = False 
         if m == WT_WRITE:
-            path = self.get_db_build_path()
+            build = True 
         else:
             self.read_metadata()
-        llo = self._create_ll_object(path)
+        llo = self._create_ll_object(build)
         llo.open(m)
         self.__ll_object = llo
         self.__open_mode = m 
@@ -466,6 +467,7 @@ class Table(Database):
     The main storage table class. 
     """
     DB_NAME = "table"
+    DATA_SUFFIX = ".dat" 
     PRIMARY_KEY_NAME = "row_id"
 
     def __init__(self, homedir):
@@ -476,13 +478,45 @@ class Table(Database):
         self.__total_row_size = 0
         self.__min_row_size = 0 
         self.__max_row_size = 0
-    
+   
+    def get_data_path(self):
+        """
+        Returns the path of the permanent data file. 
+        """
+        return os.path.join(self.get_homedir(), self.get_db_name() + 
+                self.DATA_SUFFIX)
+
+    def get_data_build_path(self):
+        """
+        Returns the path of the file used to build the database.
+        """
+        s = "_build_{0}_{1}{2}".format(os.getpid(), self.get_db_name(),
+                self.DATA_SUFFIX) 
+        return os.path.join(self.get_homedir(), s) 
+
+    def get_data_file_size(self):
+        """
+        Returns the size of the data file in bytes.    
+        """
+        statinfo = os.stat(self.get_data_path())
+        return statinfo.st_size 
 
     def finalise_build(self):
+        """
+        Finalise the build by moving the data and db files to their 
+        permanent values.
+        """
         super(Table, self).finalise_build()
-        new = self.get_db_path()
[…]

### 2013-07-28 docs/paper/paper.tex
Initial conversion to BMC Bioinformatics style.

@@
-\documentclass{bioinfo}
+%% BioMed_Central_Tex_Template_v1.05
+%%                                      %
+%  bmc_article.tex            ver: 1.05 %
+%                                       %
 
-% Bioinformatics definitions.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                     %%
+%%  LaTeX template for BioMed Central  %%
+%%     journal article submissions     %%
+%%                                     %%
+%%         <27 January 2006>           %%
+%%                                     %%
+%%                                     %%
+%% Uses:                               %%
+%% cite.sty, url.sty, bmc_article.cls  %%
+%% ifthen.sty. multicol.sty		       %%
+%%									   %%
+%%                                     %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\copyrightyear{2013}
-\pubyear{2013}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                                                 %%	
+%% For instructions on how to fill out this Tex template           %%
+%% document please refer to Readme.pdf and the instructions for    %%
+%% authors page on the biomed central website                      %%
+%% http://www.biomedcentral.com/info/authors/                      %%
+%%                                                                 %%
+%% Please do not use \input{...} to include other tex files.       %%
+%% Submit your LaTeX manuscript as one .tex document.              %%
+%%                                                                 %%
+%% All additional figures and files should be attached             %%
+%% separately and not embedded in the \TeX\ document itself.       %%
+%%                                                                 %%
+%% BioMed Central currently use the MikTex distribution of         %%
+%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
+%% http://www.miktex.org                                           %%
+%%                                                                 %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+\NeedsTeXFormat{LaTeX2e}[1995/12/01]
+\documentclass[10pt]{bmc_article}
+
+% Load packages
+\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
+\usepackage{url}  % Formatting web addresses  
+\usepackage{ifthen}  % Conditional 
+\usepackage{multicol}   %Columns
+\usepackage[utf8]{inputenc} %unicode support
+%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
+%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
+\urlstyle{rm}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%	
+%%                                             %%
+%%  If you wish to display your graphics for   %%
+%%  your own use using includegraphic or       %%
+%%  includegraphics, then comment out the      %%
+%%  following two lines of code.               %%   
+%%  NB: These line *must* be included when     %%
+%%  submitting to BMC.                         %% 
+%%  All figure files must be submitted as      %%
+%%  separate graphics through the BMC          %%
+%%  submission process, not included in the    %% 
+%%  submitted article.                         %% 
+%%                                             %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\def\includegraphic{}
+\def\includegraphics{}
+
+\setlength{\topmargin}{0.0cm}
+\setlength{\textheight}{21.5cm}
+\setlength{\oddsidemargin}{0cm} 
+\setlength{\textwidth}{16.5cm}
+\setlength{\columnsep}{0.6cm}
+
+\newboolean{publ}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                              %%
+%% You may change the following style settings  %%
+%% Should you wish to format your article       %%
+%% in a publication style for printing out and  %%
+%% sharing with colleagues, but ensure that     %%
+%% before submitting to BMC that the style is   %%
+%% returned to the Review style setting.        %%
+%%                                              %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%Review style settings
+\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
+
+%Publication style settings
+%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
+
+% Begin ...
 \begin{document}
-\firstpage{1}
+\begin{bmcformat}
 
-\title[Wormtable]{Wormtable: A data structure for genome scale data}
-\author[Kelleher \textit{et~al}]{Jerome Kelleher, Rob W. Ness 
-and Daniel L. Halligan}
-\address{
-University of Edinburgh,
-King's Buildings,
-West Mains Road,
-EH9 3JT,
-UK
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the title of your article here     %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\title{Processing genome scale tabular data with Wormtable}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the authors here                   %%
+%%                                          %%
+%% Ensure \and is entered between all but   %%
+%% the last two authors. This will be       %%
+%% replaced by a comma in the final article %%
+%%                                          %%
+%% Ensure there are no trailing spaces at   %% 
+%% the ends of the lines                    %%     	
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\author{Jerome Kelleher\correspondingauthor%
+       \email{Jerome Kelleher\correspondingauthor - jerome.kelleher@ed.ac.uk}%
+      \and
+         Rob W Ness
+         \email{Rob W Ness - rob.ness@ed.ac.uk}%
+       and 
+         Daniel L Halligan%
+         \email{Daniel L Halligan - dan.halligan@gmail.com}%
       }
 
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the authors' addresses here        %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\editor{Associate Editor: XXXXXXX}
+\address{%
+    University of Edinburgh, King's Buildings, West Mains Road, %
+        EH9 3JT, UK}%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The Abstract begins here                 %%
+%%                                          %%
+%% The Section headings here are those for  %%
+%% a Research article submitted to a        %%
+%% BMC-Series journal.                      %%  
+%%                                          %%
+%% If your article is not of this type,     %%
+%% then refer to the Instructions for       %%
+%% authors on http://www.biomedcentral.com  %%
+%% and change the section headings          %%
+%% accordingly.                             %%   
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{abstract}
-\section{Summary:}
+        % Do not use inserted blank lines (ie \\) until main body of text.
+        \paragraph*{Background:} 
       Biological research generates large amounts of tabular data.
@@
 Furthermore, there is no simple means of indexing these files so that 
-rows containing particular values can be quickly found. We introduce a new 
+rows containing particular values can be quickly found. 
+        \paragraph*{Results:} 
+We introduce a new 
 data format and software library called wormtable, which provides efficient 
@@
 indexing on columns within these tables.
-\section{Availability:}
-Package available at 
-\href{http://pypi.python.org/pypi/wormtable}{http://pypi.python.org/pypi/wormtable}
 
-\section{Contact:} \href{jerome.kelleher@ed.ac.uk}{jerome.kelleher@ed.ac.uk}
+        \paragraph*{Conclusions:} Text for this section of the abstract \ldots
 \end{abstract}
 
-\section{Introduction}
+\ifthenelse{\boolean{publ}}{\begin{multicols}{2}}{}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The Main Body begins here                %%
+%%                                          %%
+%% The Section headings here are those for  %%
+%% a Research article submitted to a        %%
+%% BMC-Series journal.                      %%  
+%%                                          %%
+%% If your article is not of this type,     %%
+%% then refer to the instructions for       %%
+%% authors on:                              %%
+%% http://www.biomedcentral.com/info/authors%%
+%% and change the section headings          %%
+%% accordingly.                             %% 
+%%                                          %%
+%% See the Results and Discussion section   %%
+%% for details on how to create sub-sections%%
+%%                                          %%
+%% use \cite{...} to cite references        %%
+%%  \cite{koon} and                         %%
+%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
+%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%
+%% Background %%
+%%
+\section*{Background}
 Despite the ever increasing volumes of data being processed in bioinformatics,
@@
 any platform with the minimum of library dependencies. Using compression,
-text files can be quite compact, and Tabix~\citep{li11} provides  
+text files can be quite compact, and Tabix~\cite{li11} provides  
 fast retrieval of rows from arbitrary locations in the table.
@@
 
-\section{Wormtable}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Results and Discussion %%
+%%
+\section*{Results and Discussion}
+\subsection*{Wormtable}
 % what is wormtable?
@@
 technologies without the overheads and disadvantages of relational databases,
-wormtable uses Berkeley DB~\citep{obs99} for row storage and indexing.
+wormtable uses Berkeley DB~\cite{obs99} for row storage and indexing.
 Berkeley DB is a free and open-source embedded database toolkit that provides a
@@
 and is currently the most widely deployed database toolkit in the 
-world~\citep{sb12}. Berkeley DB is a highly modular library, providing the 
+world~\cite{sb12}. Berkeley DB is a highly modular library, providing the 
 means to define a very lightweight read-only database, without the 
@@
 
-\section{Examples}
+\subsection*{Examples}
 The Variant Call Format (VCF) encodes information about variant sites in a genome 
-as tab-delimited rows in a text file~\citep{da+11}.  
+as tab-delimited rows in a text file~\cite{da+11}.  
 To illustrate the advantages of wormtable, we
@@
 publicly available VCF file, produced as part of the
-Drosophila genetic reference panel~\citep{mr+12}, to wormtable format.
-This VCF\footnote{\href%
-{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}
-{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}}
+Drosophila genetic reference panel~\cite{mr+12}, to wormtable format.
+This VCF%\footnote{\href%
+%{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}
+%{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}}
 consists of 15GB of uncompressed text and contains data from the 
@@
 wrote a script to count the number of transitions and transversions
-in the dataset using wormtable and PyVCF\footnote{%
-\href{http://pyvcf.rtfd.org/}{http://pyvcf.rtfd.org/}}, 
+in the dataset using wormtable and PyVCF%\footnote{%
+%\href{http://pyvcf.rtfd.org/}{http://pyvcf.rtfd.org/}}, 
 a Python VCF parser.
@@
 
-\section{Conclusions}
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Conclusions}
 It is well known that parsing text files is a very inefficient process, 
@@
 
-\section*{Acknowledgement}
+%%%%%%%%%%%%%%%%%%
+\section*{Methods}
+  \subsection*{Methods sub-heading for this section}
+    Text for this sub-section \ldots
+
+  \subsection*{Another methods sub-heading for this section}
+    Text for this sub-section \ldots
+
+  \subsection*{Yet another sub-heading for this section}
+    Text for this sub-section \ldots
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Authors contributions}
+    Text for this section \ldots
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgements}
+  \ifthenelse{\boolean{publ}}{\small}{}
 We would like to thank the Ashworth Codemonkeys for valuable feedback on earlier
-versions of wormtable.
-\paragraph{Funding\textcolon}  
-JK is supported by EPSRC grant EP/I013091/1.
+versions of wormtable.  JK is supported by EPSRC grant EP/I013091/1.
 
-\bibliographystyle{plainnat}
-\bibliography{paper}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                  The Bibliography                       %%
+%%                                                         %%              
+%%  Bmc_article.bst  will be used to                       %%
+%%  create a .BBL file for submission, which includes      %%
+%%  XML structured for BMC.                                %%
+%%                                                         %%
+%%                                                         %%
+%%  Note that the displayed Bibliography will not          %% 
+%%  necessarily be rendered by Latex exactly as specified  %%
+%%  in the online Instructions for Authors.                %% 
+%%                                                         %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
+ \bibliographystyle{bmc_article}  % Style BST file
+  \bibliography{paper.bib} }     % Bibliography file (usually '*.bib' )
+
+%%%%%%%%%%%
+
+\ifthenelse{\boolean{publ}}{\end{multicols}}{}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Figures                       %%
+%%                               %%
+%% NB: this is for captions and  %%
+%% Titles. All graphics must be  %%
+%% submitted separately and NOT  %%
+%% included in the Tex document  %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%
+%% Do not use \listoffigures as most will included as separate files
+
+\end{bmcformat}
 \end{document}

### 2013-07-28 test/utilities.py

@@ -203,7 +203,6 @@ class TestSchemaGeneration(Vcf2wtTest):
         self.assertEqual(root.tag, "schema")
         shutil.rmtree(self._homedir)
         os.mkdir(self._homedir)
-
     
     def test_generator(self):
         self.__test_schema_generator(EXAMPLE_VCF)

### 2013-07-28 docs/paper/paper.tex
Updated references and various style aspects.

@@
 %Review style settings
-\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
+%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
 
 %Publication style settings
-%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
+\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
 
@@
 
-\title{Processing genome scale tabular data with Wormtable}
+\title{Processing genome scale tabular data with wormtable}
 
@@
 
-        \paragraph*{Conclusions:} Text for this section of the abstract \ldots
+        \paragraph*{Conclusions:} \emph{ Conclusions, brief summary and potential implications. }
 \end{abstract}
@@
 \section*{Background}
+
+\emph{The Background section should be written in a
+way that is accessible to researchers without specialist knowledge in
+that area and must clearly state - and, if helpful, illustrate - the
+background to the research and its aims. It should clearly described the
+relevant context and the specific issue which the software described is
+intended to address.}
+
 Despite the ever increasing volumes of data being processed in bioinformatics,
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Results and Discussion %%
+%% Implementation 
 %%
-\section*{Results and Discussion}
-\subsection*{Wormtable}
+\section*{Implementation}
+\emph{This should include a description of the overall architecture of the software implementation, 
+along with details of any critical issues and how they were addressed.
+}
 % what is wormtable?
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Results and Discussion %%
+%%
+\section*{Results and Discussion}
+
+\emph{The Results and Discussion may be combined into a single section or
+presented separately. They may also be broken into subsections with
+short, informative headings. In any case what should be described is the
+functionality of the software together with data on how its performance
+and functionality compare with and improve on functionally similar
+existing software. There should then be a discussion of the intended
+use of the software, and the benefits that are envisioned together,
+if possible, with an outline for the planned future development of
+new features.}
+
 \subsection*{Examples}
@@
 Drosophila genetic reference panel~\cite{mr+12}, to wormtable format.
-This VCF%\footnote{\href%
+This VCF\cite{freeze.vcf}
+%\footnote{\href%
 %{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}
@@
 wrote a script to count the number of transitions and transversions
-in the dataset using wormtable and PyVCF%\footnote{%
-%\href{http://pyvcf.rtfd.org/}{http://pyvcf.rtfd.org/}}, 
+in the dataset using wormtable and PyVCF~\cite{pyvcf}
 a Python VCF parser.
@@
 \section*{Conclusions}
+
+\emph{This should state clearly the main conclusions of the article and give a clear 
+explanation of the importance and relevance of the software.}
+
 It is well known that parsing text files is a very inefficient process, 
@@
 %%%%%%%%%%%%%%%%%%
-\section*{Methods}
-  \subsection*{Methods sub-heading for this section}
-    Text for this sub-section \ldots
-
-  \subsection*{Another methods sub-heading for this section}
-    Text for this sub-section \ldots
-
-  \subsection*{Yet another sub-heading for this section}
-    Text for this sub-section \ldots
+\section*{Availability and requirements}
+\paragraph{Project name:} Wormtable.
+\paragraph{Project home page:}\url{https://pypi.python.org/pypi/wormtable}
+\paragraph{Operating system(s):} POSIX.
+\paragraph{Programming language:} C, Python.
+\paragraph{Other requirements:} Berkeley DB 4.8 or later.
+\paragraph{License:} GNU GPLv3+.
+\paragraph{Any restrictions to use by non-academics:} None.
 

### 2013-07-29 docs/paper/paper.tex
sorted out sectioning a little.

@@
 }
+
 % what is wormtable?
@@
 Wormtable currently supports signed and unsigned integers, 
-floating point and character data. To ensure that values are 
-stored as compactly as possible, integer sizes can range from  
-$1$ to $8$ bytes. The standard single and double precision 
-floating point values are also supported, along with fixed and 
-variable length strings.
+floating point and character data. To allow data to be stored as 
+efficiently as possible, wormtable supports a wide range of 
+type sizes: integers range from one to eight bytes long,
+and IEEE half, single and double precision floating point 
+are supported. Fixed and variable length character strings 
+are also supported.
+
+% Tables
 
@@
 
+% Indexes
+
 In wormtable, each table corresponds to a directory in the file system
@@
 
+% Layout & implementation.
+Each table corresponds to a single directory in the file system.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%
-\section*{Results and Discussion}
+\section*{Results}
 
@@
 
+\subsection{API}
+
 \subsection*{Examples}
@@
 This VCF\cite{freeze.vcf}
-%\footnote{\href%
-%{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}
-%{ftp://ftp.hgsc.bcm.edu/DGRP/freeze2\_Feb\_2013/vcf\_files/freeze2.vcf.gz}}
 consists of 15GB of uncompressed text and contains data from the 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-\section*{Conclusions}
-
-\emph{This should state clearly the main conclusions of the article and give a clear 
-explanation of the importance and relevance of the software.}
-
+\section{Discussion}
 It is well known that parsing text files is a very inefficient process, 
@@
 
-% If we get some help, it could be pretty awesome...
-Wormtable is an open and collaborative project seeking feedback and 
-contributors. The project is in its early stages, currently 
-providing a conversion tool for the VCF format only. With community
-participation it may grow to support many other formats and become
-a powerful common platform for bioinformatics data analysis.
+The library supports efficient access to any data stored in 
+wormtable format, but there is currently a conversion tool for
+the VCF format only. The most important aspect of future development 
+is to develop tools to convert other tabular formats such as 
+PSL, GFF, and SAM to wormtable format [Dan, Rob: does this make sense?
+are there other formats we should chuck in here?]. Since the conversion 
+tool can be written entirely in Python this is not a difficult task.
+
+Other probable directions for future development of wormtable would
+be to incorporate BGZF compression~\cite{li11}, allowing for smaller
+files with efficient random access; to develop a 
+C library and bindings for other popular languages such as 
+Perl and R; and to generally enrich the API with features useful 
+to the community. 
+Wormtable is an open and collaborative project, which would welcome 
+feedback and contributions from the community
+
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Conclusions}
+
+\emph{This should state clearly the main conclusions of the article and give a clear 
+explanation of the importance and relevance of the software.}
 

### 2013-07-30 docs/paper/paper.tex
First full draft.

@@
 %Review style settings
-%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
+\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
 
 %Publication style settings
-\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
+%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
 
@@
 
-        \paragraph*{Conclusions:} \emph{ Conclusions, brief summary and potential implications. }
+\paragraph*{Conclusions:} Using wormtable's simple API users 
+can process tables in a very high performance manner, allowing them
+much greater freedom to explore their data. 
+Since it is no longer necessary to parse complex files,
+data analysis code is also much simpler and less error prone. 
 \end{abstract}
@@
 to index the information within particular columns. 
-This means that most operations on the table require a complete scan 
+This means that most operations on a table require a complete scan 
 through the file. As a result, the only viable means of 
@@
 in the table, facilitating efficient and convenient access to 
-data.
+data. The core data manipulation and storage facilities 
+are written in C for efficiency.
 
 % Wormtable's binary row format 
-Data is stored using a  
+Data is stored in rows using a  
 compact binary format designed specifically for biological applications.
@@
 Wormtable currently supports signed and unsigned integers, 
-floating point and character data. To allow data to be stored as 
-efficiently as possible, wormtable supports a wide range of 
-type sizes: integers range from one to eight bytes long,
-and IEEE half, single and double precision floating point 
-are supported. Fixed and variable length character strings 
+floating point and character data. To store data as 
+compactly as possible, integer sizes range from one to eight bytes,
+and real numbers are stored as IEEE half, single, and double 
+precision floating point values. 
+Fixed and variable length character strings 
 are also supported.
@@
 % Tables
-
-% why do we use Berkeley DB?
-To take advantage of the data processing capabilities of database
-technologies without the overheads and disadvantages of relational databases,
-wormtable uses Berkeley DB~\cite{obs99} for row storage and indexing.
+Rows are stored sequentially in a data file, and the offset and length 
+of each row is stored in a Berkeley DB~\cite{obs99} database. 
+Thus, any given row can be found 
+by first looking up the database for its offset and length, and by then reading
+the required set of bytes from the data file. 
 Berkeley DB is a free and open-source embedded database toolkit that provides a
@@
 and is currently the most widely deployed database toolkit in the 
-world~\cite{sb12}. Berkeley DB is a highly modular library, providing the 
-means to define a very lightweight read-only database, without the 
-overheads and complexity of a relational database system.
+world~\cite{sb12}.
 
 % Indexes
+Besides random access to rows, wormtable also provides indexes
+over arbitrary combinations of columns. An index is a database in 
+which the keys are the values from the columns in question concatenated
+together, and the values are pointers to the original row. Indexes make
+many operations much more efficient, since we can directly seek to the 
+rows we are interested in without performing a full table scan. They also 
+provide a very efficient means of calculating a histogram for a given 
+set of columns.
 
-In wormtable, each table corresponds to a directory in the file system
-which is used to store the table, indexes and metadata.
-Indexes over arbitrary combinations of columns are supported, 
-and these are managed using the \texttt{wtadmin} program, which 
-provides several other administration utilities.
-Wormtable files are portable across operating systems 
-and processor architectures.
-
-% Layout & implementation.
-Each table corresponds to a single directory in the file system.
+% Layout 
+Each table corresponds to a directory in the file system
+which is used to store the data files, indexes and metadata.
+Files within a table's home directory are not intended to be 
+manipulated by users, and the \texttt{wtadmin} program is provided 
+to perform administration tasks such as adding and removing 
+indexes. This approach is 
+very flexible, since no centralised storage is required and 
+tables can be moved around and between systems at will.
+Tables are portable across operating systems and hardware
+architectures.
 
@@
 
-\subsection{API}
+\subsection*{API}
+The principle goal of wormtable is to provide efficient access to data 
+using a simple interface. The \texttt{Table} class provides the main interface,
+and implements the standard Python sequence protocols. Efficient 
+iteration over rows is provided by the \texttt{cursor} method, 
+which takes as an argument the list of columns to read (similar to 
+a select list in SQL). Only the values for the columns 
+of interest are then retrieved, leading to considerable
+time savings.
+Providing an index as an argument to the \texttt{cursor} function
+results in the rows being visited in the order defined by the 
+index. The arguments \texttt{start} and \texttt{stop} 
+then stipulate the range of values we are interested in.
+There are several other useful classes and methods, but 
+the \texttt{cursor} method encapsulates the core functionality
+of wormtable: efficient iteration over rows.
+
+The interface is kept as
+simple and similar to the Python standard library as possible,
+to ensure that there is little difficulty in learning the API.
+Similarly, although one can readily imagine defining a 
+query language using these 
+primitive operations, we have deliberately avoided doing so. Adding 
+an SQL-like query language would greatly increase the complexity 
+of wormtable, and would only serve to raise the barrier to entry.
+Less experienced programmers 
+do not need to learn a query language to process data 
+in a very high performance manner.
 
@@
 on a standard workstation with a single hard disk. The
-conversion required approximately 68 minutes, and the size of the resulting  
-wormtable file was 10.1GB (using the smallest type required
+conversion required approximately 69 minutes, and the size of the resulting  
+wormtable was 10GB (using the smallest type required
 to represent the data in each column).
@@
 the wormtable version required
-55 seconds. To compare against methods that are known to be extremely
+57 seconds. To compare against methods that are known to be extremely
 efficient, we repeated the same 
 example using the Unix tools \texttt{cut}, \texttt{grep} 
-and \texttt{awk},  which required 77 seconds. Both the
+and \texttt{awk},  which required 80 seconds. Both the
 Unix pipeline and wormtable were limited by the sequential read
@@
 faster using modern solid state storage.
-It should be noted, however, that this simple task represents 
-something of a best case scenario for the Unix tools pipeline.
+It should be noted, however, that although Unix 
+pipelines are efficient, it is a difficult and error prone method 
+of processing data when more complex calculations are required.
 
-Another major advantage of wormtable is the ability to create 
+Another advantage of wormtable is the ability to create 
 an index over arbitrary combinations of columns within a table.
@@
 on the REF and ALT columns.
-The index required 3m42s to build and 
+The index required 3m40s to build and 
 consumed 66MB of storage space. Counting the number of 
@@
 
-\section{Discussion}
+\section*{Discussion}
 It is well known that parsing text files is a very inefficient process, 
-and binary file formats are preferable for this reason. The BCF format,
+and binary file formats are preferable for this reason. BCF,
 for example, is a binary version of the VCF format discussed above which 
@@
 Wormtable alleviates the need for these different formats and libraries,
-as it is flexible enough to store many different types of data. Python 
-is currently the only programming language supported, but a C library
-and bindings for other languages are planned. Wormtable is portable,
+as it is flexible enough to store many different types of data. Wormtable is portable,
 and has been tested on big- and little-endian platforms with 32 and 64
@@
 task than designing and supporting a custom binary file format.
+
+The library supports efficient access to any data stored in 
+wormtable format, but there is currently a conversion tool for
+the VCF format only. The most important aspect of future development 
+is to develop tools to convert other tabular formats such as 
+PSL, GFF, and SAM to wormtable format [Dan, Rob: does this make sense?
+are there other formats we should chuck in here?].
+
+Other probable directions for future development of wormtable 
+are to incorporate data file compression; to develop a 
+C library and bindings for other popular languages such as 
+Perl and R; and to generally enrich the API with features useful 
+to the community. It is difficult to predict the 
+precise direction of future development since this depends 
+on community involvement.  Wormtable is an open and 
+collaborative project, actively seeking feedback and 
+contributors.
+
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Conclusions}
+\emph{This should state clearly the main conclusions of the article and give a clear 
+explanation of the importance and relevance of the software.}
+
+The volumes of data being produced and processed in bioinformatics are constantly 
+increasing, but the tools available to end users to 
+process this data are still mostly based on parsing plain text. This 
+approach is very inefficient, and leads to several undesirable outcomes.
+Firstly, and most obviously, a researcher's productivity is inevitably
+reduced by having to wait for several hours to see the result of a 
+simple calculation. There are also some less obvious consequences:
+code quality is reduced, since testing over the entire dataset is 
+infeasible, and it is much less likely that the effects of 
+changing arbitrary quality thresholds (for example) will be 
+systematically examined.
 
@@
 productivity and ability to explore their data.
-
-The library supports efficient access to any data stored in 
-wormtable format, but there is currently a conversion tool for
-the VCF format only. The most important aspect of future development 
-is to develop tools to convert other tabular formats such as 
-PSL, GFF, and SAM to wormtable format [Dan, Rob: does this make sense?
-are there other formats we should chuck in here?]. Since the conversion 
-tool can be written entirely in Python this is not a difficult task.
-
-Other probable directions for future development of wormtable would
-be to incorporate BGZF compression~\cite{li11}, allowing for smaller
-files with efficient random access; to develop a 
-C library and bindings for other popular languages such as 
-Perl and R; and to generally enrich the API with features useful 
-to the community. 
-Wormtable is an open and collaborative project, which would welcome 
-feedback and contributions from the community
-
-%%%%%%%%%%%%%%%%%%%%%%
-\section*{Conclusions}
-
-\emph{This should state clearly the main conclusions of the article and give a clear 
-explanation of the importance and relevance of the software.}
 

### 2013-08-01 _wormtablemodule.c
Fixed compound index bug by inserting separator.

@@ -2618,6 +2618,8 @@ Index_init(Index *self, PyObject *args, PyObject *kwds)
             self->key_buffer_size += col->num_elements * col->element_size;
         }
     }
+    /* Add in space for the column separators */
+    self->key_buffer_size += self->num_columns;
     self->key_buffer = PyMem_Malloc(self->key_buffer_size);
     if (self->key_buffer == NULL) {
         PyErr_NoMemory();
@@ -2733,7 +2735,7 @@ Index_fill_key(Index *self, void *row, DBT *skey)
     Column *col;
     uint32_t j;
     int len;
-    void *v = skey->data;
+    unsigned char *v = skey->data;
     skey->size = 0;
     for (j = 0; j < self->num_columns; j++) {
         col = self->table->columns[self->columns[j]]; 
@@ -2748,8 +2750,14 @@ Index_fill_key(Index *self, void *row, DBT *skey)
         if (len < 0) {
             goto out;
         }
-        skey->size += len;
         v += len;
+        skey->size += len;
+        /* insert the separator between columns */
+        if (j < self->num_columns - 1) {
+            *v = 0;
+            v++;
+            skey->size++;
+        }
     }
     ret = 0;
 out: 
@@ -2770,7 +2778,7 @@ Index_set_key(Index *self, PyObject *args, void *buffer)
     Column *col = NULL;
     PyObject *elements = NULL;
     PyObject *v = NULL;
-    void *dest = buffer; 
+    unsigned char *dest = buffer; 
     if (!PyArg_ParseTuple(args, "O!", &PyTuple_Type, &elements)) { 
         goto out;
     }
@@ -2802,6 +2810,12 @@ Index_set_key(Index *self, PyObject *args, void *buffer)
         col->pack_elements(col, dest);
         dest += m;
         size += m;
+        /* insert the separator between columns */
+        if (j < n - 1) {
+            *dest = 0;
+            dest++;
+            size++; 
+        }
     }
     ret = size;
 out:

### 2013-08-01 test/highlevel.py

@@ -462,6 +462,41 @@ class BinnedIndexIntegrityTest(WormtableTest):
                 self.assertEqual(d[k], v)
             i.close()
 
+class StringIndexIntegrityTest(WormtableTest):
+    """
+    Tests the integrity of indexes over variable length length 
+    columns by checking some simple cases over strings.
+    """
+    
+    def test_unique_keys(self):
+        """
+        Test the simplest possible case where we have keys that 
+        cannot be distinguished if we concatentate the values
+        together.
+        """
+        t = wt.Table(self._homedir) 
+        t.add_id_column(1)
+        t.add_char_column("s1")
+        t.add_char_column("s2")
+        t.open("w")
+        t.append([None, b"A", b"AA"])
+        t.append([None, b"AA", b"A"])
+        t.close()
+        t.open("r")
+        i = wt.Index(t, "test")
+        i.add_key_column(t.get_column("s1"))
+        i.add_key_column(t.get_column("s2"))
+        i.open("w")
+        i.build()
+        i.close()
+        i.open("r")
+        c = i.counter()
+        self.assertEqual(c[(b"A", b"AA")], 1)
+        self.assertEqual(c[(b"AA", b"A")], 1)
+        t.close()
+ 
+
+
 class TableCursorTest(WormtableTest):
     """
     Tests the cursor over a table.

### 2013-08-01 test/lowlevel.py

@@ -905,22 +905,17 @@ class TestIndexIntegrity(object):
         for j, k in pairs:
             c1 = self._columns[j]
             c2 = self._columns[k]
-            # TODO: variable length columns don't sort the same way
-            # this might be a bug or it might not - this should be 
-            # verified and the correct sorting procedure used here.
-            if c1.num_elements != _wormtable.WT_VAR_1 and \
-                    c2.num_elements != _wormtable.WT_VAR_1:
-                fd, index_file = tempfile.mkstemp("-index-test.db", prefix=TEMPFILE_PREFIX)  
-                index = _wormtable.Index(self._database, index_file.encode(), [j, k], 
-                        cache_size)
-                os.close(fd)
-                index.open(WT_WRITE)
-                index.build()
-                index.close()
-                index_files.append(index_file)
-                indexes.append(index)
-                col_positions.append([j, k])
-                original_values.append([(row[j], row[k]) for row in self.rows])
+            fd, index_file = tempfile.mkstemp("-index-test.db", prefix=TEMPFILE_PREFIX)  
+            index = _wormtable.Index(self._database, index_file.encode(), [j, k], 
+                    cache_size)
+            os.close(fd)
+            index.open(WT_WRITE)
+            index.build()
+            index.close()
+            index_files.append(index_file)
+            indexes.append(index)
+            col_positions.append([j, k])
+            original_values.append([(row[j], row[k]) for row in self.rows])
 
         for index, original, cols in zip(indexes, original_values, col_positions):
             index.open(WT_READ)
@@ -1018,12 +1013,9 @@ class TestMultiColumnIndex(object):
         self._index_files = []
         # generate n indexes each with a random choice of 1,...,k columns.
         v = []
-        # Variable length columns sort differently so we leave them out.
-        # TODO fix this!
         for j in range(1, len(self._columns)):
             c = self._columns[j]
-            if c.num_elements != _wormtable.WT_VAR_1:
-                v.append(j)
+            v.append(j)
         max_columns = 5
         max_indexes = 10 
         for j in range(2, max_columns):

### 2013-08-01 wormtable.py

@@ -39,7 +39,7 @@ import _wormtable
 
 __version__ = '0.1.0b2'
 TABLE_METADATA_VERSION = "0.1"
-INDEX_METADATA_VERSION = "0.1"
+INDEX_METADATA_VERSION = "0.1.1"
 
 DEFAULT_CACHE_SIZE = 16 * 2**20 # 16M 
 DEFAULT_CACHE_SIZE_STR = "16M" 

### 2013-08-01 wormtable.py
Added initial version of the gtf2wt converter.

@@ -1257,4 +1257,111 @@ class ProgressMonitor(object):
         """
         print()
 
+BROKEN_GZIP_MESSAGE = """
+An error occurred reading the input gzip file. This is probably due to a
+bug in recent versions of Python, resulting in an error when trying 
+to read BGZF files. See http://bugs.python.org/issue17666 
+To convert this file to wormtable, you must decompress it first using gunzip.
+"""
 
+class FileReader(object):
+    """
+    A class for reading data files from a variety of sources and 
+    with progress updating.
+    """
+    def __init__(self, in_file):
+        if in_file == '-':
+            self.__input_file = sys.stdin
+            if sys.version_info[:2] >= (3, 1):
+                try:
+                    self.__input_file = sys.stdin.buffer
+                except AttributeError:
+                    # When we're testing, we replace stdin with an ordinary
+                    # file. This does not support buffer, but is also not 
+                    # needed so we can skip this step
+                    pass
+            self.__progress_monitor = None
+            self.__input_file_size = None
+            self.__progress_file = None
+        else:
+            if in_file.endswith(".gz"):
+                # Detect broken GZIP handling in 2.7/3.2 and others and abort
+                # TODO this has been fixed upstream and can be removed at 
+                # some point.
+                f = gzip.open(in_file, "rb")
+                try:
+                    s = f.readline()
+                except Exception as e:
+                    print(BROKEN_GZIP_MESSAGE)
+                    sys.exit(1)
+                f.close() 
+                # Carry on as before
+                self.__input_file = gzip.open(in_file, "rb")
+                self.__progress_file = self.__input_file.fileobj
+            else:
+                self.__input_file = open(in_file, "rb")
+                self.__progress_file = self.__input_file 
+            statinfo = os.stat(in_file)
+            self.__input_file_size = statinfo.st_size 
+        self.__progress_update_rows = 2**32 
+        self.__progress_monitor = None
+
+    def get_progress_update_rows(self):
+        """
+        Returns the number of rows after which we should update the progress 
+        monitor.
+        """
+        return self.__progress_update_rows
+
+    def set_progress_update_rows(self, update_rows):
+        """
+        Sets the number of rows after which we should update progress 
+        to the specified value.
+        """
+        self.__progress_update_rows = update_rows
+
+    def get_input_file(self):
+        """
+        Returns the File object that is the source of the information 
+        in this reader.
+        """
+        return self.__input_file
+
[…]

### 2013-08-03 test/utilities.py
Fixed Python2/3 bug in gtf2wt and tests.

@@ -288,7 +288,7 @@ class WtadminTest(UtilityTest):
             i.close()
             i.open("r")
             s = self.run_dump(["0", "--index=" + i.get_name()])         
-            c = self._table.cursor([0], i)
+            c = i.cursor([0])
             n = 0
             for line, r in zip(s.splitlines(), c): 
                 n += 1
@@ -371,7 +371,13 @@ class GtfBuildTest(object):
                    d[spl[0]] = spl[1].strip("\"")
             row[GENE_ID] = d[GENE_ID]
             row[TRANSCRIPT_ID] = d[TRANSCRIPT_ID]
-            self._rows.append(row)
+            # if the type is str, encode it.
+            r = {}
+            for k, v in row.items():
+                r[k] = v
+                if isinstance(v, str):
+                    r[k] = v.encode()
+            self._rows.append(r)
             self._num_rows += 1
   
     def tearDown(self):

### 2013-08-03 _wormtablemodule.c
Fixed get_max bug on index and test cases.
This required a change in the index file format where we now append a
zero after every column, not just the internal ones.

@@ -2753,11 +2753,9 @@ Index_fill_key(Index *self, void *row, DBT *skey)
         v += len;
         skey->size += len;
         /* insert the separator between columns */
-        if (j < self->num_columns - 1) {
-            *v = 0;
-            v++;
-            skey->size++;
-        }
+        *v = 0;
+        v++;
+        skey->size++;
     }
     ret = 0;
 out: 
@@ -2766,19 +2764,20 @@ out:
 
 /* 
  * Reads the arguments and sets a key in the specified buffer, returning 
- * its length.
+ * its length. If increment is equal to 1, increment the key so that it
+ * is equal to the next key larger than the specified value.
  */
 static int 
-Index_set_key(Index *self, PyObject *args, void *buffer)
+Index_set_key(Index *self, PyObject *args, void *buffer, int increment)
 {
     int ret = -1;
     int j, m;
-    int size = 0;
+    int key_size = 0;
     Py_ssize_t n;
     Column *col = NULL;
     PyObject *elements = NULL;
     PyObject *v = NULL;
-    unsigned char *dest = buffer; 
+    unsigned char *key_buffer = buffer; 
     if (!PyArg_ParseTuple(args, "O!", &PyTuple_Type, &elements)) { 
         goto out;
     }
@@ -2803,24 +2802,42 @@ Index_set_key(Index *self, PyObject *args, void *buffer)
             goto out;
         }
         m = col->num_buffered_elements * col->element_size;
-        if (size + m > self->key_buffer_size) {
-            PyErr_Format(PyExc_SystemError, "Max key size exceeded."); 
+        if (key_size + m + 1 > self->key_buffer_size) {
+            PyErr_Format(PyExc_SystemError, "Max key key_size exceeded."); 
             goto out;
         }
-        col->pack_elements(col, dest);
-        dest += m;
-        size += m;
+        col->pack_elements(col, key_buffer);
+        key_buffer += m;
+        key_size += m;
         /* insert the separator between columns */
-        if (j < n - 1) {
-            *dest = 0;
-            dest++;
-            size++; 
-        }
+        *key_buffer = 0;
+        key_buffer++;
+        key_size++; 
     }
-    ret = size;
+    ret = key_size;
 out:
     return ret;
 }
+
+
+static void
+Index_increment_key(Index *self, void *buffer, u_int32_t key_size)
+{
+    int j;
+    unsigned char *key_buffer = (unsigned char *) buffer;
[…]

### 2013-08-03 test/highlevel.py

@@ -500,6 +500,35 @@ class StringIndexIntegrityTest(WormtableTest):
         self.assertEqual(c[(b"AA", b"A")], 1)
         t.close()
  
+    def test_max_prefix(self):
+        """
+        Test the simplest possible case where we must get the maximum 
+        key for a given prefix.
+        """
+        t = wt.Table(self._homedir) 
+        t.add_id_column(1)
+        t.add_char_column("s1")
+        t.add_char_column("s2")
+        t.open("w")
+        t.append([None, b"A", b"A"])
+        t.append([None, b"A", b"AA"])
+        t.append([None, b"A", b"B"])
+        t.append([None, b"AA", b""])
+        t.append([None, b"B", b"A"])
+        t.close()
+        t.open("r")
+        i = wt.Index(t, "test")
+        i.add_key_column(t.get_column("s1"))
+        i.add_key_column(t.get_column("s2"))
+        i.open("w")
+        i.build()
+        i.close()
+        i.open("r")
+        self.assertEqual(i.max_key(), (b"B", b"A"))
+        self.assertEqual(i.max_key("A"), (b"A", b"B"))
+        self.assertEqual(i.max_key("AA"), (b"AA", b""))
+        t.close()
+ 
 
 
 class TableCursorTest(WormtableTest):

### 2013-08-03 wormtable.py

@@ -40,7 +40,7 @@ import _wormtable
 
 __version__ = '0.1.0b2'
 TABLE_METADATA_VERSION = "0.1"
-INDEX_METADATA_VERSION = "0.1.1"
+INDEX_METADATA_VERSION = "0.1.2"
 
 DEFAULT_CACHE_SIZE = 16 * 2**20 # 16M 
 DEFAULT_CACHE_SIZE_STR = "16M" 
@@ -641,11 +641,9 @@ class Table(Database):
         ret = None
         if isinstance(col_id, int):
             ret = self.__columns[col_id]
-        elif isinstance(col_id, str):
+        else:
             k = self.__column_name_map[col_id]
             ret = self.__columns[k]
-        else:
-            raise TypeError("column ids must be strings or integers")
         return ret
     
     def translate_columns(self, columns):
@@ -1220,66 +1218,6 @@ class IndexCounter(collections.Mapping):
         for v in dvi:
             n += 1
         return n
-
-class Cursor(object):
-    """
-    Superclass of Cursor objects. Subclasses are responsible for allocating 
-    an iterator.
-    """
-    def __iter__(self):
-        for r in self._row_iterator:
-            yield r
-
-    def set_min(self, *v):
-        """
-        Sets the minimum row key to retrieve to the specified value.
-        """
-        raise NotImplementedError
-
-    def set_max(self, *v):
-        """
-        Sets the maximum row key to retrieve to the specified value.
-        """
-        raise NotImplementedError
-
-class DELTableCursor(Cursor):
-    """
-    A cursor over the rows of the table in the order defined by an index. 
-    """
-    def __init__(self, table, columns):
-        self._columns = columns
-        col_positions = [c.get_position() for c in columns]
-        self._row_iterator = _wormtable.TableRowIterator(table.get_ll_object(), 
-                col_positions)
-
-    def set_start(self, v):
-        if v < 0:
-            raise ValueError("negative row_ids not supported")
-        self._row_iterator.set_min(v)
-    def set_stop(self, v):
-        if v < 0:
-            raise ValueError("negative row_ids not supported")
-        self._row_iterator.set_max(v)
-
-class DELIndexCursor(Cursor):
-    """
-    A cursor over the rows of the table in the order defined by an index. 
-    """
-    def __init__(self, index, columns):
-        self.__index = index
-        self._columns = columns
-        col_positions = [c.get_position() for c in columns]
[…]

### 2013-08-03 _wormtablemodule.c
Removed extra parameter.

@@ -2764,11 +2764,10 @@ out:
 
 /* 
  * Reads the arguments and sets a key in the specified buffer, returning 
- * its length. If increment is equal to 1, increment the key so that it
- * is equal to the next key larger than the specified value.
+ * its length. 
  */
 static int 
-Index_set_key(Index *self, PyObject *args, void *buffer, int increment)
+Index_set_key(Index *self, PyObject *args, void *buffer)
 {
     int ret = -1;
     int j, m;
@@ -2891,7 +2890,7 @@ Index_get_num_rows(Index *self, PyObject *args)
     DB *db = NULL;
     DBC *cursor = NULL;
     DBT key, data;
-    key_size = Index_set_key(self, args, self->key_buffer, 0);
+    key_size = Index_set_key(self, args, self->key_buffer);
     if (key_size < 0) {
         goto out;
     }
@@ -2939,7 +2938,7 @@ Index_get_min(Index* self, PyObject *args)
     int db_ret;
     DBC *cursor = NULL;
     DBT primary_key, primary_data, secondary_key;
-    int key_size = Index_set_key(self, args, self->key_buffer, 0);
+    int key_size = Index_set_key(self, args, self->key_buffer);
     if (key_size < 0) {
         goto out;
     }
@@ -2988,7 +2987,7 @@ Index_get_max(Index* self, PyObject *args)
     DBC *cursor = NULL;
     DBT primary_key, primary_data, secondary_key;
     unsigned char *search_key = NULL;
-    int key_size = Index_set_key(self, args, self->key_buffer, 0);
+    int key_size = Index_set_key(self, args, self->key_buffer);
     if (key_size < 0) {
         goto out;
     }
@@ -3885,7 +3884,7 @@ static PyObject *
 IndexRowIterator_set_min(IndexRowIterator *self, PyObject *args)
 {
     PyObject *ret = NULL;
-    int size = Index_set_key(self->index, args, self->min_key, 0);
+    int size = Index_set_key(self->index, args, self->min_key);
     if (size < 0) {
         goto out;
     }
@@ -3899,7 +3898,7 @@ static PyObject *
 IndexRowIterator_set_max(IndexRowIterator *self, PyObject *args)
 {
     PyObject *ret = NULL;
-    int size = Index_set_key(self->index, args, self->max_key, 0);
+    int size = Index_set_key(self->index, args, self->max_key);
     if (size < 0) {
         goto out;
     }

### 2013-08-04 test/utilities.py
Minor updates to tests.

@@ -26,6 +26,7 @@ import scripts.gtf2wt as gtf2wt
 import scripts.wtadmin as wtadmin
 
 import unittest
+import random
 import tempfile
 import shutil
 import os.path
@@ -277,6 +278,19 @@ class WtadminTest(UtilityTest):
             for u, v in zip(l, r):
                 self.assertEqual(u.encode(), v)
         self.assertEqual(n, len(self._table))
+        # Now check it over a range
+        for j in range(10):
+            start = random.randint(0, len(self._table))
+            stop = random.randint(start, len(self._table))
+            s = self.run_dump(cols + [ "--start=" + str(start), 
+                    "--stop=" + str(stop)])         
+            c = self._table.cursor(cols)
+            for line, r in zip(s.splitlines(), c): 
+                l = line.split()
+                for u, v in zip(l, r):
+                    self.assertEqual(u.encode(), v)
+            
+
     
     def test_dump_index(self):
         cols = ["CHROM", "REF", "ALT"]

### 2013-08-04 docs/paper/paper.tex
Added Tabix discussion and seek performance section.

@@
 binary format, provides random access to rows, and enables sophisticated
-indexing on columns within these tables.
+indexing on columns within these tables. Files written in existing 
+formats can be easily converted to wormtable format, and we 
+provide conversion utilities for the VCF and GTF format.
 
@@
 provide a very efficient means of calculating a histogram for a given 
-set of columns.
+set of columns. Wormtable also supports binned indexes, in which a range of 
+values is mapped to a single index key. This is very useful for floating 
+point columns, where it may not be necessary to distinguish between
+values that differ by small amounts.
 
@@
 The principle goal of wormtable is to provide efficient access to data 
-using a simple interface. The \texttt{Table} class provides the main interface,
+using an easy to learn interface. 
+The \texttt{Table} class provides the main interface,
 and implements the standard Python sequence protocols. Efficient 
 iteration over rows is provided by the \texttt{cursor} method, 
-which takes as an argument the list of columns to read (similar to 
-a select list in SQL). Only the values for the columns 
+which takes as an argument the list of columns to read.
+Only the values for the columns 
 of interest are then retrieved, leading to considerable
-time savings.
-Providing an index as an argument to the \texttt{cursor} function
-results in the rows being visited in the order defined by the 
-index. The arguments \texttt{start} and \texttt{stop} 
-then stipulate the range of values we are interested in.
-There are several other useful classes and methods, but 
-the \texttt{cursor} method encapsulates the core functionality
-of wormtable: efficient iteration over rows.
+time savings. The \texttt{cursor} function takes two additional 
+arguments, \texttt{start} and \texttt{stop}, which dictate 
+rows of interest. This allows us to efficiently seek to an 
+arbitary location in the table and to read a given number 
+of rows sequentially from this point.
 
-The interface is kept as
-simple and similar to the Python standard library as possible,
-to ensure that there is little difficulty in learning the API.
-Similarly, although one can readily imagine defining a 
-query language using these 
-primitive operations, we have deliberately avoided doing so. Adding 
-an SQL-like query language would greatly increase the complexity 
-of wormtable, and would only serve to raise the barrier to entry.
-Less experienced programmers 
-do not need to learn a query language to process data 
-in a very high performance manner.
+The \texttt{Index} class also has a \texttt{cursor} function
+with the same signature, but in this case, rows 
+are returned in the order defined by the index. The  
+\texttt{start} and \texttt{stop} arguments are now defined 
+in terms of index keys: all rows in which the index key
+is greater than or equal to \texttt{start} and less 
+than \texttt{stop} are returned. Partial keys may also be 
+provided for multi-column indexes. The \texttt{Index} 
+class also provides an iterator over all keys, as well 
+as a means of counting the number of rows with a given 
+key.
 
-\subsection*{Examples}
+The API is straightforward, but it is flexible, powerful and 
+extensible. This simplicity ensures that 
+programmer of all experience levels  
+can take full advantage of the powerful data 
+processing that wormtable provides. In the following subsections 
+we illustrate the performance advantages of wormtable via some 
+examples. These are not intended to be definitive benchmarks
+and are simply concrete examples to demonstrate the type of improvements 
+that can be expected by using wormtable over existing methods.
+
+\subsection*{Scan performance}
 The Variant Call Format (VCF) encodes information about variant sites in a genome 
 as tab-delimited rows in a text file~\cite{da+11}.  
-To illustrate the advantages of wormtable, we
-converted a large 
-publicly available VCF file, produced as part of the
-Drosophila genetic reference panel~\cite{mr+12}, to wormtable format.
+To illustrate the advantages of wormtable when performing calculations over 
+a whole table, we converted a large 
+publicly available VCF file (produced as part of the
+Drosophila genetic reference panel~\cite{mr+12}) to wormtable format.
 This VCF\cite{freeze.vcf}
@@
 the VCF was converted to wormtable format 
-on a standard workstation with a single hard disk. The
+on a workstation with an Intel Xeon processor, 12GB of RAM 
+and a single hard disk. The
 conversion required approximately 69 minutes, and the size of the resulting  
@@
 in the dataset using wormtable and PyVCF~\cite{pyvcf}
-a Python VCF parser.
-In this example we proceed 
+a Python VCF parser.  In this example we proceed 
 row-by-row, examining the REF and ALT columns and
@@
 
-Another advantage of wormtable is the ability to create 
-an index over arbitrary combinations of columns within a table.
-Indexes make many operations much more efficient, as they provide 
-a way to quickly find particular rows and avoid the 
-heavy cost of a full table scan.
-To illustrate this advantage, we repeated the example 
+Full table scans are often uneccessary in wormtable. For many tasks,
+creating an appropriate index allows us to seek directly to the 
+rows of interest. To illustrate this, we repeated the example 
 of counting transitions and transversions using an index 
@@
 
+\subsection*{Seek performance}
+One of the most serious problems with tabular data stored in text files
+is that accessing a particular row efficiently without some auxiliary 
+index. Without some information on where a particular row is located
+in a file, there is nothing that can be done except to read the file 
+line-by-line until the required row is found. Tabix~\cite{li11} solves
+this problem by compressing a tab-delimited text file into blocks, and 
+then storing an index mapping genome position to the location 
+of the compressed block, and the position of the row within the block.
+Tabix is specifically designed for range queries, allowing us to 
+efficiently retrieve all of the rows within a given genomic range.
+
+To compare the seek performance of wormtable with Tabix, we compressed
+the VCF file mentioned in the previous subsection using \texttt{bgzip}
+and indexed it using \texttt{tabix}. Since the VCF file was not 
+strictly in order of genome position (in violation of the VCF format),
+we first needed to sort the VCF file, which required 5m13s. Afterwards,
+compression required 11m8s, and indexing with \texttt{tabix} required
+1m12s. The size of the resulting compressed file was 2.9GB and 
+the size of the index file was 106KB.
+To duplicate the functionality of Tabix in wormtable, we simply need to 
+create an index on the CHROM and POS columns using \texttt{wtadmin add}. 
+This required approximately
+2 minutes and the resulting index consumed 115.4MB of space.
+
+We compared the seek performance of Tabix and wormtable by generating 
+a genomic location randomly and retrieving all rows within 1Kb of
+this location. This 
+was repeated $10^4$ times, and we measured both the elapsed and 
+processor time. The Tabix Python module was used for the 
+comparison, ensuring that no overheads associated with process 
+forking were incurred. The same set of random locations were used 
+for wormtable and Tabix, ensuring a fair comparison.
+
+When using a cold cache (i.e., no pages of the files in question are 
+present in the operating system's cache) the elapsed
+time for Tabix was 103 seconds with a processor time of 65
+seconds. For wormtable, the elapsed time was 181 seconds 
+with a processor time of 5 seconds. Thus, the time required 
+to perform this test is dominated by waiting for I/O in both 
+cases. Since the wormtable file is considerably larger than the 
+compressed file used by Tabix, more and larger seeks were required 
+to bring the required pages into memory. However, once the 
+pages were in memory, Tabix needed to do much more work to 
+decompress then and make them usable, as shown by the 
+difference in processor times.
+
+This difference is well illustrated by immediately repeating 
+the same experiment, so that all the relevent pages are 
+in cache. In this case, the elapsed time was 56 seconds for Tabix and around 
+2 seconds for wormtable. Thus, compression is a double-edged sword: 
+it does lead to much smaller files, which reduces the number of random seeks 
+required.  On the other hand, though, decompression is expensive and must 
+be repeated each time a block is accessed, unless some sophisticated
+caching mechanism is devised.
+
+The penalty for random seeks is exceedingly high in this 
+example, however, since we have used a hard disk for storage.
+Repeating the cold cache example on a laptop equipped with 
+solid state storage, the elapsed time for Tabix was 76 seconds
+with a CPU time of 72 seconds, whereas the elapsed time for 
+wormtable was 31 seconds with a CPU time of 6.5 seconds. Thus, using 
+solid state storage negates the advantage of smaller files, since 
+seeks are much less expensive.  
+ With the rapidly reducing cost and 
+increasing capacity of solid state storage, it seems likely 
+that it will soon be routinely used for all but the largest 
+of bulk storage. Wormtable takes much better advantage of
+modern hardware and large cache sizes, allowing the full bandwidth 
+and random seek performance to be utilised. Compression based 
+techniques will always be limited by the CPU time required to 
+decompress data.
+
 \section*{Discussion}
-It is well known that parsing text files is a very inefficient process, 
-and binary file formats are preferable for this reason. BCF,
-for example, is a binary version of the VCF format discussed above which 
-is designed to avoid just this problem. There are significant difficulties,
+The problems of enabling efficient random access to rows and 
+avoiding the large overhead of parsing text are well understood, 
+and efforts to address them are proceeding in parallel for 
+different file formats. 
+BCF, for example, is the binary version of the VCF format discussed 
+above, in which values within rows are stored in a packed binary 
+format.  Similiarly, BigBed and 
+BigWig~\cite{kzbhk10} are compressed binary versions of the BED 
+and WIG file formats, which offer efficient random access along 
+with the ability to operate over a network.
+
+There are significant difficulties,
 however, with having many different binary file formats for bioinformatics 
@@
 is a complex software engineering task. Maintaining this ecosystem 
-separately for several different file formats is surely 
+separately for many different file formats is surely 
 unsustainable.
@@
 The library supports efficient access to any data stored in 
-wormtable format, but there is currently a conversion tool for
-the VCF format only. The most important aspect of future development 
+wormtable format, and currently provides conversion utilitites 
+for the VCF and GTF formats.
+The most important aspect of future development 
 is to develop tools to convert other tabular formats such as 
-PSL, GFF, and SAM to wormtable format [Dan, Rob: does this make sense?
-are there other formats we should chuck in here?].
+PSL, GFF and BED to wormtable format. Such tools are 
+not difficult to develop, since all that is required is a parser
+for the format in question written in Python.
 
-Other probable directions for future development of wormtable 
-are to incorporate data file compression; to develop a 
-C library and bindings for other popular languages such as 
-Perl and R; and to generally enrich the API with features useful 
-to the community. It is difficult to predict the 
-precise direction of future development since this depends 
-on community involvement.  Wormtable is an open and 
-collaborative project, actively seeking feedback and 
+Wormtable is currently limited to supporting Python only, and 
+another important aspect of future development is to 
+develop a C library along with bindings for other popular
+languages such as Perl, R and PHP. Compression of data can
+result in poor performance, but it is often necessary when 
+volumes of data are very large. Thus, we plan on introducing 
+optional compression of the data file in wormtable in 
+a future release. Beyond these additions, it is difficult
+predict the precise direction of future development since this depends 
+on feedback from the community.  Wormtable is an open and 
+collaborative project actively seeking feedback and 
 contributors.
@@
 
+We have concentrated on the VCF format here since it is a widely 
+used and popular format. After converting the VCF format to 
+wormtable using \texttt{vcf2wt} a user can process the 
+data very efficiently using Python. Accessing data from 
+rows in wormtable is many times faster than is possible 
+by parsing rows encoded as text.
+This is also a very convenient 
+way to access VCF data, since individual columns are already 
+parsed and all that is needed is the name of the columns of 
+interest. For accessing regions of the genome efficiently
+the user simply needs to create an index on the CHROM and POS columns
+using \texttt{wtadmin add} command, and can then efficiently
+seek to arbitrary locations. In the worst case, when using a 
+slow hard disk this is somewhat slower than Tabix. In the best
+case, when frequently accessing the same regions or using 
+fast solid state storage, wormtable is many times faster than 
+Tabix. Moreover, wormtable is much more general than Tabix, 
+since it can index any column or combination of columns.
+Similarly, wormtable is not limited to VCF, but can store
+any form of fixed tabular data. We also provide a conversion 
+tool for the GTF format, and several others are planned.
+
 Wormtable is not intended to replace text files as the universal 
@@
 database systems can take full advantage of sophisticated 
-data management techniques, and write simple code to process 
-data in a high performance manner. 
+data management techniques, and write straightforward code to process 
+data efficiently. 
 Different file formats can be handled consistently 
@@
 \paragraph{Other requirements:} Berkeley DB 4.8 or later.
-\paragraph{License:} GNU GPLv3+.
+\paragraph{License:} GNU LGPLv3+.
 \paragraph{Any restrictions to use by non-academics:} None.

### 2013-08-04 docs/paper/paper.tex
Added discussion of interval search + minor edits

@@
 
-Another problem with tables stored text files is that it is difficult 
+Another problem with tables stored as text files is that it is difficult 
 to index the information within particular columns. 
-This means that most operations on a table require a complete scan 
+This means that many operations on a table require a complete scan 
 through the file. As a result, the only viable means of 
@@
 particular data set, particularly not if portability across different 
-databases is required. Similarly, accessing the data requires a knowledge
+databases is required. Similarly, accessing data requires a knowledge
 of SQL. In the most common case, a database server must also be maintained
@@
 time savings. The \texttt{cursor} function takes two additional 
-arguments, \texttt{start} and \texttt{stop}, which dictate 
+arguments, \texttt{start} and \texttt{stop}, which specify 
 rows of interest. This allows us to efficiently seek to an 
@@
 counting the transitions and transversions we encounter. 
-Using PyVCF this required approximately 126 minutes;
+Using PyVCF this required approximately 126 minutes, whereas 
 the wormtable version required
@@
 One of the most serious problems with tabular data stored in text files
-is that accessing a particular row efficiently without some auxiliary 
-index. Without some information on where a particular row is located
-in a file, there is nothing that can be done except to read the file 
+is that it is not possible to access a particular row 
+efficiently without some auxiliary 
+index. Unless we have some information on where a particular row is located
+in a file, there is little that can be done except to read the file 
 line-by-line until the required row is found. Tabix~\cite{li11} solves
@@
 the VCF file mentioned in the previous subsection using \texttt{bgzip}
-and indexed it using \texttt{tabix}. Since the VCF file was not 
+and indexed it with \texttt{tabix}. Since the VCF file was not 
 strictly in order of genome position (in violation of the VCF format),
@@
 the size of the index file was 106KB.
-To duplicate the functionality of Tabix in wormtable, we simply need to 
+To duplicate the functionality of Tabix on VCF data, we simply need to 
 create an index on the CHROM and POS columns using \texttt{wtadmin add}. 
 This required approximately
-2 minutes and the resulting index consumed 115.4MB of space.
+2 minutes and the resulting index consumed 115MB of space.
 
@@
 it does lead to much smaller files, which reduces the number of random seeks 
-required.  On the other hand, though, decompression is expensive and must 
-be repeated each time a block is accessed, unless some sophisticated
+required.  On the other hand, however, decompression is expensive and must 
+be repeated each time a block is accessed unless some sophisticated
 caching mechanism is devised.
@@
 develop a C library along with bindings for other popular
-languages such as Perl, R and PHP. Compression of data can
+languages such as Perl, R and PHP. 
+Wormtable does not support interval search, and so it is not
+straightforward to find, for example, all rows overlapping 
+a given genomic region in GTF files. This problem has been 
+solved several times, however, and we aim to adapt existing 
+techniques~\cite{li11,kzbhk10} and incorporate them into 
+wormtable.  Compression of data can
 result in poor performance, but it is often necessary when 
@@
 the user simply needs to create an index on the CHROM and POS columns
-using \texttt{wtadmin add} command, and can then efficiently
+using the  \texttt{wtadmin add} command, and can then efficiently
 seek to arbitrary locations. In the worst case, when using a 

### 2013-08-06 test/highlevel.py
Made start/stop systematic accross tools and tests.

@@ -359,9 +359,9 @@ class IndexIntegrityTest(WormtableTest):
             self._indexes.append(i)
 
 
-    def test_accessors(self):
+    def test_keys(self):
         """
-        Test if the accessor operations are working properly.
+        Test if the key operations are working properly.
         """
         for i in self._indexes:
             i.open("r")
@@ -390,15 +390,49 @@ class IndexIntegrityTest(WormtableTest):
                 else:
                     key = tuple(kk.get_value() for kk in k)
                 self.assertEqual(v, c2[key])
-            # now test the cursor
-            read_cols = self._table.columns()
-            j = 0
-            for r1 in i.cursor(read_cols):
-                j = int(r1[0])
-                row = self._table[j]
-                r2 = tuple(row[col.get_position()] for col in read_cols)
-                self.assertEqual(r1, r2)
             i.close()
+
+    def test_cursors(self):
+        read_cols = self._table.columns()
+        for i in self._indexes:
+            i.open("r")
+            cols = i.key_columns()
+            t = [[tuple(ColumnValue(c, r) for c in cols), r] for r in self._table] 
+            t.sort(key=lambda x: x[0])
+            for (k, r1), r2  in zip(t, i.cursor(read_cols)):
+                self.assertEqual(r1, r2)
+            if len(cols) == 1:
+                keys = [(k,) for k in i.keys()]
+            else:
+                keys = [k for k in i.keys()]
+            key_rows = [tuple(v.get_value() for v in k) for k, r in t]
+            # Now generate some slices
+            for j in range(10):
+                k = random.randint(0, len(keys) - 1)
+                start_key = keys[k]
+                start_index = key_rows.index(start_key)
+                k = random.randint(k, len(keys) - 1)
+                stop_key = keys[k]
+                stop_index = key_rows.index(stop_key)
+                l = [r for k, r in t[start_index:stop_index]]
+                c = 0
+                for r1, r2 in zip(i.cursor(read_cols, start_key, stop_key), l):
+                    self.assertEqual(r1, r2)
+                    c += 1
+                self.assertEqual(c, stop_index - start_index)
+                # If we have a single column index we also support passing the 
+                # values directly
+                if len(cols) == 1:
[…]

### 2013-08-06 test/utilities.py

@@ -284,14 +284,34 @@ class WtadminTest(UtilityTest):
             stop = random.randint(start, len(self._table))
             s = self.run_dump(cols + [ "--start=" + str(start), 
                     "--stop=" + str(stop)])         
-            c = self._table.cursor(cols)
+            c = self._table.cursor(cols, start, stop)
             for line, r in zip(s.splitlines(), c): 
                 l = line.split()
                 for u, v in zip(l, r):
                     self.assertEqual(u.encode(), v)
-            
+   
+    def verify_dump(self, index, start=None, stop=None):
+        cols = [col.get_name() for col in index.key_columns()]
+        args = cols + ["--index=" + index.get_name()]
+        if start is not None:
+            args += ["--start=" + start.decode()]
+        if stop is not None:
+            args += ["--stop=" + stop.decode()]
+        s = self.run_dump(args)
+        n1 = 0
+        for r in index.cursor(cols, start, stop):
+            n1 += 1
+        n2 = 0
+        for line in s.splitlines():
+            n2 += 1
+        c = index.cursor(cols, start, stop)
+        for line, r in zip(s.splitlines(), c): 
+            l = line.split()
+            for u, v in zip(l, r):
+                self.assertEqual(u.encode(), v)
+        self.assertEqual(n1, n2)
 
-    
+
     def test_dump_index(self):
         cols = ["CHROM", "REF", "ALT"]
         indexes = [wt.Index(self._table, col) for col in cols]
@@ -308,9 +328,35 @@ class WtadminTest(UtilityTest):
                 n += 1
                 self.assertEqual(int(line), r[0])
             self.assertEqual(n, len(self._table))
+            keys = [k for k in i.keys()]
+            for j in range(10):
+                k = random.randint(0, len(keys) - 1)
+                start = keys[k]
+                stop = keys[random.randint(k, len(keys) - 1)]
+                self.verify_dump(i, start, stop)
+                self.verify_dump(i, stop=stop)
+                self.verify_dump(i, start=start)
             i.close()
             i.delete() 
-    
+   
+    def test_dump_multi_index(self):
+        i = wt.Index(self._table, "REF+ALT") 
+        i.add_key_column(self._table.get_column("REF"))
+        i.add_key_column(self._table.get_column("ALT"))
+        i.open("w")
+        i.build()
+        i.close()
+        i.open("r")
+        s = self.run_dump(["0", "--index=" + i.get_name()])         
[…]

### 2013-08-06 wormtable.py

@@ -1163,16 +1163,19 @@ class Index(Database):
         return iri
 
 
-
     def key_to_ll(self, v):
         """
         Translates the specified tuple as a key to a tuple ready to 
         for use in the low-level API.
         """
-        n = len(v)
+        cols = self.__key_columns
+        u = v
+        if len(cols) == 1 and not isinstance(v, tuple):
+            u = v,
+        n = len(u)
         l = [None for j in range(n)]
         for j in range(n):
-            l[j] = v[j]
+            l[j] = u[j]
             if isinstance(l[j], str):
                 l[j] = l[j].encode()
         return tuple(l)

### 2013-08-12 test/lowlevel.py
Mitigated memory leak in tests.

@@ -131,6 +131,7 @@ class TestDatabase(unittest.TestCase):
         self._row_buffer = self._database 
         self.num_random_test_rows = num_random_test_rows 
         self.num_columns = len(self._columns) 
+        self.rows = []
 
     def open_reading(self):
         """
@@ -143,7 +144,7 @@ class TestDatabase(unittest.TestCase):
         self._database.close()
         os.unlink(self._db_file)
         os.unlink(self._data_file)
-
+        self.rows = None
 
     def assertRowListsEqual(self, l1, l2):
         """
@@ -1312,7 +1313,8 @@ class TestIndex(unittest.TestCase):
         os.unlink(self._table_db_file)
         os.unlink(self._table_data_file)
         os.unlink(self._index_db_file)
-
+        self._table = None
+        self._columns = None
 
 class TestIndexInitialisation(TestIndex):
     """ 

### 2013-08-17 docs/paper/paper.tex
Added some comments to paper.tex
Incorporated Dan's comments.

@@
 line-by-line, parsing values encoded as text into native types so that 
-calculations may be done. This is a time consuming process.
+calculations may be done. This is an error prone and time consuming process.
 Furthermore, there is no simple means of indexing these files so that 
@@
 
-\emph{The Background section should be written in a
-way that is accessible to researchers without specialist knowledge in
-that area and must clearly state - and, if helpful, illustrate - the
-background to the research and its aims. It should clearly described the
-relevant context and the specific issue which the software described is
-intended to address.}
-
+%\emph{The Background section should be written in a
+%way that is accessible to researchers without specialist knowledge in
+%that area and must clearly state - and, if helpful, illustrate - the
+%background to the research and its aims. It should clearly described the
+%relevant context and the specific issue which the software described is
+%intended to address.}
+%
 Despite the ever increasing volumes of data being processed in bioinformatics,
@@
 any platform with the minimum of library dependencies. Using compression,
-text files can be quite compact, and Tabix~\cite{li11} provides  
-fast retrieval of rows from arbitrary locations in the table.
+text files can be quite compact, and specialised indexing methods 
+are available to quickly retrieve rows intersecting with a 
+given genomic interval~\cite{li11}.
 
@@
 machine values. This is an expensive process, and 
-compression (if it is used) adds another substantial overhead. The result 
-of this is that simple calculations over a large dataset may take 
+compression (if it is used) adds another substantial overhead. 
+As a result,  simple calculations over a large dataset may take 
 many hours to complete.
@@
 This means that many operations on a table require a complete scan 
-through the file. As a result, the only viable means of 
-working with a subset of the data
+through the file. The only viable means of 
+working with a subset of the data, therefore, 
 is to create another file consisting of the subset of interest. 
@@
 \section*{Implementation}
-\emph{This should include a description of the overall architecture of the software implementation, 
-along with details of any critical issues and how they were addressed.
-}
+%\emph{This should include a description of the overall architecture of the software implementation, 
+%along with details of any critical issues and how they were addressed.
+%}
 
@@
 data of essentially unlimited size. Tables can
-be either written directly via the library, or converted
+be either written directly using the library, or converted
 from existing formats via command line tools. 
@@
 % Wormtable's binary row format 
-Data is stored in rows using a  
-compact binary format designed specifically for biological applications.
+Data is stored in rows using a compact binary format.
 Each row consists of a fixed number of columns, and each column 
 holds either a fixed or variable number of elements of a fixed type and size.
+Each column is allocated a fixed number of bytes within the `fixed region' of 
+a row. For fixed
+length columns, the encoded elements are stored entirely within the
+fixed region; for variable length columns, the encoded elements are 
+stored in the `variable region' of the row, and the location and number 
+of elements stored in the fixed region. The type, size and number 
+of elements of 
+each column are specified in the table schema, which is defined 
+in XML as part of the table's metadata.
+
 Wormtable currently supports signed and unsigned integers, 
-floating point and character data. To store data as 
-compactly as possible, integer sizes range from one to eight bytes,
-and real numbers are stored as IEEE half, single, and double 
-precision floating point values. 
+floating point and character data. Integer sizes range 
+from one to eight bytes, and real numbers are stored as 
+IEEE half, single, and double precision~\cite{ieee08}
+floating point values.  
 Fixed and variable length character strings 
-are also supported.
+are also supported. Using this flexible type system, columns can 
+be assigned the most appropriate type and size for the range of
+values to be represented.
 
@@
 Besides random access to rows, wormtable also provides indexes
-over arbitrary combinations of columns. An index is a database in 
-which the keys are the values from the columns in question concatenated
+over arbitrary combinations of columns. An index is a 
+Berkeley DB database in 
+which the keys are the elements from the columns in question concatenated
 together, and the values are pointers to the original row. Indexes make
@@
 
-\emph{The Results and Discussion may be combined into a single section or
-presented separately. They may also be broken into subsections with
-short, informative headings. In any case what should be described is the
-functionality of the software together with data on how its performance
-and functionality compare with and improve on functionally similar
-existing software. There should then be a discussion of the intended
-use of the software, and the benefits that are envisioned together,
-if possible, with an outline for the planned future development of
-new features.}
-
+%\emph{The Results and Discussion may be combined into a single section or
+%presented separately. They may also be broken into subsections with
+%short, informative headings. In any case what should be described is the
+%functionality of the software together with data on how its performance
+%and functionality compare with and improve on functionally similar
+%existing software. There should then be a discussion of the intended
+%use of the software, and the benefits that are envisioned together,
+%if possible, with an outline for the planned future development of
+%new features.}
+%
 \subsection*{API}
@@
 extensible. This simplicity ensures that 
-programmer of all experience levels  
+programmers of all experience levels  
 can take full advantage of the powerful data 
-processing that wormtable provides. In the following subsections 
+processing facilities that wormtable provides. In the following subsections 
 we illustrate the performance advantages of wormtable via some 
@@
 wrote a script to count the number of transitions and transversions
-in the dataset using wormtable and PyVCF~\cite{pyvcf}
+in the dataset using wormtable and PyVCF~\cite{pyvcf},
 a Python VCF parser.  In this example we proceed 
@@
 the VCF file mentioned in the previous subsection using \texttt{bgzip}
-and indexed it with \texttt{tabix}. Since the VCF file was not 
-strictly in order of genome position (in violation of the VCF format),
-we first needed to sort the VCF file, which required 5m13s. Afterwards,
-compression required 11m8s, and indexing with \texttt{tabix} required
-1m12s. The size of the resulting compressed file was 2.9GB and 
+(which required 11m8s) and indexed it with \texttt{tabix} (1m12s). 
+The size of the resulting compressed file was 2.9GB and 
 the size of the index file was 106KB.
@@
 cases. Since the wormtable file is considerably larger than the 
-compressed file used by Tabix, more and larger seeks were required 
-to bring the required pages into memory. However, once the 
-pages were in memory, Tabix needed to do much more work to 
+compressed file used by Tabix, more and larger hard drive
+seeks were required 
+to bring the required pages into memory (solid state 
+storage does not suffer from this problem). Once the 
+pages were in memory, however, Tabix needed to do much more work to 
 decompress then and make them usable, as shown by the 
@@
 
-The penalty for random seeks is exceedingly high in this 
-example, however, since we have used a hard disk for storage.
-Repeating the cold cache example on a laptop equipped with 
-solid state storage, the elapsed time for Tabix was 76 seconds
-with a CPU time of 72 seconds, whereas the elapsed time for 
-wormtable was 31 seconds with a CPU time of 6.5 seconds. Thus, using 
-solid state storage negates the advantage of smaller files, since 
-seeks are much less expensive.  
- With the rapidly reducing cost and 
-increasing capacity of solid state storage, it seems likely 
-that it will soon be routinely used for all but the largest 
-of bulk storage. Wormtable takes much better advantage of
-modern hardware and large cache sizes, allowing the full bandwidth 
-and random seek performance to be utilised. Compression based 
-techniques will always be limited by the CPU time required to 
-decompress data.
+The test used here is also highly synthetic, and unlikely to be 
+indicative of most real-world applications. The majority of 
+workloads have strong locality of reference~\cite{d05}, 
+and such large and extreme seeks across genomic regions 
+are unlikely to occur. In this case, wormtable 
+is  much faster than Tabix, since there
+is no CPU overhead of decompression. Furthermore, since 
+Tabix returns rows of text, the problem of parsing rows
+must still be solved. This is expensive (as illustrated 
+in the previous subsection), and creates extra code 
+complexity. Wormtable, by providing a simple API to 
+access both rows and columns, gives a 
+unified interface for accessing data that is 
+both straightforward to use and highly efficient.
 
@@
 Wormtable alleviates the need for these different formats and libraries,
-as it is flexible enough to store many different types of data. Wormtable is portable,
+as it is flexible enough to store many different types of data. 
+Wormtable is portable,
 and has been tested on big- and little-endian platforms with 32 and 64
@@
 is to develop tools to convert other tabular formats such as 
-PSL, GFF and BED to wormtable format. Such tools are 
-not difficult to develop, since all that is required is a parser
+PSL, GFF, SAM and BED to wormtable format. 
+Such tools are not difficult to develop, since all that is required is a parser
 for the format in question written in Python.
@@
 straightforward to find, for example, all rows overlapping 
-a given genomic region in GTF files. This problem has been 
+a given genomic region in GTF files. 
+This problem has been 
 solved several times, however, and we aim to adapt existing 
@@
 \section*{Conclusions}
-\emph{This should state clearly the main conclusions of the article and give a clear 
-explanation of the importance and relevance of the software.}
-
-The volumes of data being produced and processed in bioinformatics are constantly 
+%\emph{This should state clearly the main conclusions of the article and give a clear 
+%explanation of the importance and relevance of the software.}
+%
+The volume of data being produced in bioinformatics is constantly 
 increasing, but the tools available to end users to 
-process this data are still mostly based on parsing plain text. This 
+process data are still mostly based on parsing plain text. This 
 approach is very inefficient, and leads to several undesirable outcomes.
 Firstly, and most obviously, a researcher's productivity is inevitably
-reduced by having to wait for several hours to see the result of a 
-simple calculation. There are also some less obvious consequences:
+reduced by being forced to wait for several hours for the result of a 
+simple calculation. Without flexible indexing, working with 
+a subset of a data file usually requires the creation of 
+another file consisting of the subset in question, requiring 
+extra storage and maintenance.
+There are also some less obvious consequences:
 code quality is reduced, since testing over the entire dataset is 
@@
 
-We have concentrated on the VCF format here since it is a widely 
-used and popular format. After converting the VCF format to 
+The classical approach to solving problems of this type 
+is to use a relational database, which provide sophisticated 
+data management techniques. Relational databases are unsuitable 
+for storing static datasets, however, as they are complex to 
+use and incur many unnecessary overheads.
+Wormtable provides the most important features of database
+technologies (packed binary storage of values; random access 
+to rows; general purpose indexing) without their complexities 
+and overheads. Wormtable is also more flexible than 
+relational databases, supporting, for example, columns 
+containing a variable number of integers. Finally, wormtable 
+is far more adaptable than a relational database.
+Any widely used database system today is a hugely complex 
+piece of software, and adding necessary features would be 
+very difficult. Adding new features to wormtable, on the 
+other hand, is straightforward since it is far simpler and 
+does not need to be compatible with the relational model
+and decades worth of existing software.
+
+The most important aspect of wormtable is its efficiency
+and ease of use for end users, and we illustrated these
+points using some examples of VCF data.
+After converting a file in VCF format to 
 wormtable using \texttt{vcf2wt} a user can process the 
@@
 way to access VCF data, since individual columns are already 
-parsed and all that is needed is the name of the columns of 
-interest. For accessing regions of the genome efficiently
+parsed and all that is needed is the name of the column of 
+interest. To access regions of the genome efficiently
 the user simply needs to create an index on the CHROM and POS columns
 using the  \texttt{wtadmin add} command, and can then efficiently
-seek to arbitrary locations. In the worst case, when using a 
-slow hard disk this is somewhat slower than Tabix. In the best
-case, when frequently accessing the same regions or using 
-fast solid state storage, wormtable is many times faster than 
-Tabix. Moreover, wormtable is much more general than Tabix, 
-since it can index any column or combination of columns.
-Similarly, wormtable is not limited to VCF, but can store
+seek to arbitrary locations. Wormtable is not 
+limited to VCF, but can store
 any form of fixed tabular data. We also provide a conversion 
@@
 \section*{Authors contributions}
-    Text for this section \ldots
+JK conceived the project, developed wormtable source code and associated
+scripts and drafted the manuscript. RN wrote the tutorial and contributed to
+the example scripts and the manuscript. DH wrote the example scripts and vcf-utils,
+and contributed to vcf2wt, the tutorial and the manuscript.
+All authors read and approved the final manuscript.
 

### 2013-08-17 docs/paper/paper.tex
Spell checked.

@@
 rows of interest. This allows us to efficiently seek to an 
-arbitary location in the table and to read a given number 
+arbitrary location in the table and to read a given number 
 of rows sequentially from this point.
@@
 
-Full table scans are often uneccessary in wormtable. For many tasks,
+Full table scans are often unnecessary in wormtable. For many tasks,
 creating an appropriate index allows us to seek directly to the 
@@
 This difference is well illustrated by immediately repeating 
-the same experiment, so that all the relevent pages are 
+the same experiment, so that all the relevant pages are 
 in cache. In this case, the elapsed time was 56 seconds for Tabix and around 
@@
 above, in which values within rows are stored in a packed binary 
-format.  Similiarly, BigBed and 
+format.  Similarly, BigBed and 
 BigWig~\cite{kzbhk10} are compressed binary versions of the BED 
@@
 The library supports efficient access to any data stored in 
-wormtable format, and currently provides conversion utilitites 
+wormtable format, and currently provides conversion utilities 
 for the VCF and GTF formats.

### 2013-08-18 _wormtablemodule.c
Changed generic tests to Column_is_variable.

@@ -431,6 +431,15 @@ missing_float(u_int32_t k)
  *==========================================================
  */
 
+/* 
+ * Returns true if this Column is variable length.
+ */
+static int 
+Column_is_variable(Column *self)
+{
+    return self->num_elements == WT_VAR_1;
+}
+
 /**************************************
  *
  * Native values to Python conversion. 
@@ -499,7 +508,7 @@ Column_native_to_python_char(Column *self, int index)
     PyObject *ret = NULL;
     int j = self->num_buffered_elements - 1;
     char *str = (char *) self->element_buffer;
-    if (self->num_elements != WT_VAR_1) {
+    if (!Column_is_variable(self)) {
         /* check for shortened fixed-length strings, which will be padded 
          * with NULLs */
         while (str[j] == '\0' && j >= 0) {
@@ -890,7 +899,7 @@ Column_parse_python_sequence(Column *self, PyObject *elements)
             goto out;
         }
         num_elements = PySequence_Fast_GET_SIZE(seq);
-        if (self->num_elements == WT_VAR_1) {
+        if (Column_is_variable(self)) {
             if (num_elements > MAX_NUM_ELEMENTS) {
                 PyErr_Format(PyExc_ValueError, 
                         "too many elements for column '%s'",
@@ -910,7 +919,7 @@ Column_parse_python_sequence(Column *self, PyObject *elements)
             self->input_elements[j] = v; 
         }
     }
-    if (self->num_elements != WT_VAR_1) {
+    if (!Column_is_variable(self)) {
         if (num_elements != self->num_elements) {
             PyErr_Format(PyExc_ValueError, 
                     "incorrect number of elements for column '%s'",
@@ -1095,7 +1104,7 @@ Column_python_to_native_char(Column *self, PyObject *elements)
     }
     memcpy(self->element_buffer, s, length);
     self->num_buffered_elements = length;
-    if (self->num_elements != WT_VAR_1) {
+    if (!Column_is_variable(self)) {
         /* Pad out the remaining space with zeros */
         while (self->num_buffered_elements != self->num_elements) {
             ((char *) self->element_buffer)[self->num_buffered_elements] = '\0';
@@ -1160,7 +1169,7 @@ Column_parse_string_sequence(Column *self, char *s)
             j++;
         }
     }
-    if (self->num_elements != WT_VAR_1) {
+    if (!Column_is_variable(self)) {
         if (num_elements != self->num_elements) {
             Column_encoded_elements_parse_error(self, 
[…]

### 2013-08-18 _wormtablemodule.c
Initial working version of var(2).

@@ -35,7 +35,9 @@
 #define WT_CHAR 3
 
 #define WT_VAR_1 0 
-#define MAX_NUM_ELEMENTS 255
+#define WT_VAR_2 (-1) 
+#define WT_VAR_1_MAX_ELEMENTS 255
+#define WT_VAR_2_MAX_ELEMENTS 65535 
 #define MAX_ROW_SIZE 65536
 
 #define OFFSET_LEN_RECORD_SIZE 10 
@@ -56,7 +58,7 @@ typedef struct Column_t {
     int element_size;
     int num_elements;
     int fixed_region_offset;
-    void **input_elements; /* pointer to each elements in input format */
+    void **input_elements; /* pointer to each element in input format */
     void *element_buffer; /* parsed input elements in native CPU format */
     int num_buffered_elements;
     int (*string_to_native)(struct Column_t*, char *);  
@@ -437,7 +439,24 @@ missing_float(u_int32_t k)
 static int 
 Column_is_variable(Column *self)
 {
-    return self->num_elements == WT_VAR_1;
+    return self->num_elements == WT_VAR_1 || self->num_elements == WT_VAR_2;
+}
+
+/* 
+ * Returns the maximum number of elements that can be stored in this Column. 
+ */
+static uint32_t 
+Column_get_max_num_elements(Column *self)
+{
+    uint32_t ret = self->num_elements;
+    if (Column_is_variable(self)) {
+        if (self->num_elements == WT_VAR_1) {
+            ret = WT_VAR_1_MAX_ELEMENTS;
+        } else {
+            ret = WT_VAR_2_MAX_ELEMENTS;
+        }
+    }
+    return ret;
 }
 
 /**************************************
@@ -889,6 +908,7 @@ Column_parse_python_sequence(Column *self, PyObject *elements)
     int j, num_elements;
     PyObject *seq = NULL;
     PyObject *v;
+    int overflow;
     self->num_buffered_elements = 0;
     if (self->num_elements == 1) {
         self->input_elements[0] = elements;
@@ -900,7 +920,12 @@ Column_parse_python_sequence(Column *self, PyObject *elements)
         }
         num_elements = PySequence_Fast_GET_SIZE(seq);
         if (Column_is_variable(self)) {
-            if (num_elements > MAX_NUM_ELEMENTS) {
+            overflow = 
[…]

### 2013-08-18 test/highlevel.py

@@ -226,10 +226,14 @@ class TableBuildTest(WormtableTest):
             t.add_int_column("i_" + str(j),  num_elements=j) 
             t.add_float_column("f_" + str(j), num_elements=j) 
             t.add_char_column("c_" + str(j), num_elements=j) 
-        t.add_uint_column("u_v", num_elements=wt.WT_VAR_1)  
-        t.add_int_column("i_v", num_elements=wt.WT_VAR_1)  
-        t.add_float_column("f_v", num_elements=wt.WT_VAR_1)  
-        t.add_char_column("c_v", num_elements=wt.WT_VAR_1) 
+        t.add_uint_column("u_v1", num_elements=wt.WT_VAR_1)  
+        t.add_int_column("i_v1", num_elements=wt.WT_VAR_1)  
+        t.add_float_column("f_v1", num_elements=wt.WT_VAR_1)  
+        t.add_char_column("c_v1", num_elements=wt.WT_VAR_1) 
+        t.add_uint_column("u_v2", num_elements=wt.WT_VAR_2)  
+        t.add_int_column("i_v2", num_elements=wt.WT_VAR_2)  
+        t.add_float_column("f_v2", num_elements=wt.WT_VAR_2)  
+        t.add_char_column("c_v2", num_elements=wt.WT_VAR_2) 
         t.open("w")
         t.append([])
         t.close()
@@ -859,5 +863,69 @@ class FloatTest(WormtableTest):
             da = np.array([d], dtype=np.float64)
             results.append((float(ha[0]), float(sa[0]), float(da[0])))
         self.assert_tables_equal(values, results)
+
+
+class LargeStringTest(WormtableTest):
+    """
+    Tests the behaviour of large strings in both fixed and variable 
+    sizes.
+    """
+    def test_single_column(self):
+        t = wt.Table(self._homedir)
+        t.add_id_column(1)
+        n = 2**16 - 1 
+        strings = [b"", b"x" * (n // 4), b"x" * (n // 2), b"x" * n]
+        t.add_char_column("s", num_elements=n)
+        t.open("w")
+        for s in strings:
+            t.append([None, s])
+        for s in strings:
+            t.append_encoded([None, s])
+        strings = strings + strings 
+        # try to add longer strings
+        self.assertRaises(ValueError, t.append, [None, b"y" * (n + 1)])
+        self.assertRaises(ValueError, t.append_encoded, [None, b"y" * (n + 1)])
[…]

### 2013-08-18 test/lowlevel.py

@@ -32,7 +32,6 @@ import _wormtable
 from _wormtable import WT_READ 
 from _wormtable import WT_WRITE
 from _wormtable import MAX_ROW_SIZE 
-from _wormtable import WT_VAR_1 as WT_VARIABLE 
 from _wormtable import WormtableError 
 
 __column_id = 0
@@ -112,6 +111,19 @@ def random_string(n):
     """
     s = ''.join(random.choice(string.ascii_letters) for x in range(n)) 
     return s
+
+def get_random_num_elements(col):
+    """
+    Returns a random number of elements of the specified column. If the 
+    column is fixed, return the number of elements. If not, return a 
+    random number.
+    """
+    n = col.num_elements
+    if n == _wormtable.WT_VAR_1:
+        n = random.randint(1, _wormtable.WT_VAR_1_MAX_ELEMENTS)
+    elif n == _wormtable.WT_VAR_2:
+        n = random.randint(1, 2 * _wormtable.WT_VAR_1_MAX_ELEMENTS)
+    return n 
 
 class TestDatabase(unittest.TestCase):
     """
@@ -230,8 +242,6 @@ class TestElementParsers(TestDatabase):
                 self.assertRaises(TypeError, rb.insert_elements, c.position, v)
         
 
-   
-
 class TestListParsers(TestDatabase):
     """
     Test the list parsers to make sure that malformed lists are correctly
@@ -272,7 +282,7 @@ class TestDatabaseLimits(TestDatabase):
     def get_columns(self):
         var_1_overhead = 3
         n = (_wormtable.MAX_ROW_SIZE - self._key_size) // (
-                _wormtable.MAX_NUM_ELEMENTS * 8 + var_1_overhead)
+                _wormtable.WT_VAR_1_MAX_ELEMENTS * 8 + var_1_overhead)
                 
         columns = [get_int_column(8) for j in range(n)]
         return columns
@@ -283,7 +293,7 @@ class TestDatabaseLimits(TestDatabase):
         it should overflow whatever we do.
         """
         rb = self._row_buffer
-        v = [j for j in range(_wormtable.MAX_NUM_ELEMENTS)]
+        v = [j for j in range(_wormtable.WT_VAR_1_MAX_ELEMENTS)]
         n = len(self._columns)
         for k in range(1, n):
             rb.insert_elements(k, v) 
@@ -301,7 +311,7 @@ class TestDatabaseLimits(TestDatabase):
[…]

### 2013-08-18 wormtable.py

@@ -53,6 +53,7 @@ WT_CHAR = _wormtable.WT_CHAR
 WT_READ = _wormtable.WT_READ
 WT_WRITE = _wormtable.WT_WRITE
 WT_VAR_1  = _wormtable.WT_VAR_1
+WT_VAR_2  = _wormtable.WT_VAR_2
 
 def open_table(homedir, db_cache_size=DEFAULT_CACHE_SIZE_STR):
     """
@@ -145,12 +146,25 @@ class Column(object):
     def get_num_elements(self):
         """
         Returns the number of elements in this column. This is either a 
-        positive integer >= 1 or WT_VAR1. If the number of elements 
-        is WT_VAR1, the number of elements in the column is variable,
-        from 0 to 255.
+        positive integer >= 1, or the symbolic constants WT_VAR_1
+        or WT_VAR_2.
         """
         return self.__ll_object.num_elements
   
+    def get_max_num_elements(self):
+        """
+        Returns the maximum number of elements in this column. If the 
+        column is fixed length, this is the number of elements. If 
+        it is variable, this is the maximum number that can be inserted.
+        """
+        return self.__ll_object.get_max_num_elements()
+
+    def is_variable(self):
+        """
+        Returns True if this column supports a variable number of elements. 
+        """ 
+        return self.__ll_object.is_variable()
+
     def get_missing_value(self):
         """
         Returns the missing value for this column.
@@ -163,7 +177,7 @@ class Column(object):
         if t == WT_CHAR:
             ret = b''
         elif n != 1: 
-            if n == WT_VAR_1:
+            if n == self.is_variable(): 
                 ret = tuple()
             else:
                 ret = tuple(None for j in range(n))
@@ -191,6 +205,8 @@ class Column(object):
         n = self.get_num_elements()
         if n == WT_VAR_1:
             num_elements = "var(1)"
+        elif n == WT_VAR_2:
+            num_elements = "var(2)"
         else:
             num_elements = str(self.get_num_elements())
         d = {
@@ -220,6 +236,8 @@ class Column(object):
         s = xmlcol.get("num_elements")
         if s == "var(1)":
             num_elements = WT_VAR_1
+        elif s == "var(2)":
+            num_elements = WT_VAR_2
         else:
             num_elements = int(s)
         element_type = reverse[xmlcol.get("element_type")]

### 2013-08-20 _wormtablemodule.c
Implemented different encoding.

@@ -36,8 +36,8 @@
 
 #define WT_VAR_1 0 
 #define WT_VAR_2 (-1) 
-#define WT_VAR_1_MAX_ELEMENTS 255
-#define WT_VAR_2_MAX_ELEMENTS 65535 
+#define WT_VAR_1_MAX_ELEMENTS 254
+#define WT_VAR_2_MAX_ELEMENTS 65534 
 #define MAX_ROW_SIZE 65536
 
 #define OFFSET_LEN_RECORD_SIZE 10 
@@ -1350,6 +1350,34 @@ Column_pack_variable_elements_address(Column *self, void *dest,
         uint32_t offset, uint32_t num_elements)
 {
     int ret = -1;
+    void *v = dest;
+
+    //printf("packed %d %d\n", offset, num_elements);
+    if (offset >= MAX_ROW_SIZE) {
+        PyErr_SetString(PyExc_SystemError, "Row overflow");
+        goto out;
+    }
+    pack_uint((u_int64_t) offset, v, 2);
+    v += 2;
+    if (self->num_elements == WT_VAR_1) {
+        if (num_elements > WT_VAR_1_MAX_ELEMENTS) {
+            PyErr_SetString(PyExc_SystemError, "too many elements");
+            goto out;
+        }
+        pack_uint((u_int64_t) num_elements, v, 1);
+    } else {
+        if (num_elements > WT_VAR_2_MAX_ELEMENTS) {
+            PyErr_SetString(PyExc_SystemError, "too many elements");
+            goto out;
+        }
+        pack_uint((u_int64_t) num_elements, v, 2);
+    }
+    ret = 0;
+out:
+    return ret;
+    
+#if 0
+    int ret = -1;
     uint16_t off = (uint16_t) offset;
     uint8_t n_1 = (uint8_t) num_elements;
     uint16_t n_2 = (uint16_t) num_elements;
@@ -1387,6 +1415,7 @@ Column_pack_variable_elements_address(Column *self, void *dest,
     ret = 0;
 out:
     return ret;
+#endif
 }   
 
 /*
@@ -1397,6 +1426,41 @@ static int
 Column_unpack_variable_elements_address(Column *self, void *src, 
         uint32_t *offset, uint32_t *num_elements)
 {
+    int ret = -1;
+    u_int64_t off = 0;
+    u_int64_t n = 0;
+    void *v = src;
+    off = unpack_uint(v, 2);
+    if (off != missing_uint(2)) {
+        if (off > MAX_ROW_SIZE) {
[…]

### 2013-08-20 _wormtablemodule.c
Changed variable size columns storage format.
The variable size column format is now consistent with the encoding
used elsewhere. This makes the process of adding large rows and
string much simpler.

Bumped version numbers for metadata formats and module version as
this is a backwards incompatible change.

@@ -35,8 +35,8 @@
 #define WT_CHAR 3
 
 #define WT_VAR_1 0 
-#define MAX_NUM_ELEMENTS 255
-#define MAX_ROW_SIZE 65536
+#define MAX_NUM_ELEMENTS 254
+#define MAX_ROW_SIZE 65535
 
 #define OFFSET_LEN_RECORD_SIZE 10 
 
@@ -1314,9 +1314,11 @@ Column_pack_variable_elements_address(Column *self, void *dest,
         uint32_t offset, uint32_t num_elements)
 {
     int ret = -1;
-    uint16_t off = (uint16_t) offset;
-    uint8_t n = (uint8_t) num_elements;
     void *v = dest;
+    /* these are currently hard coded to 2 and 1 - future versions
+     * will support general address_size and var_size */
+    unsigned int address_size = 2;
+    unsigned int var_size = 1;
     if (offset >= MAX_ROW_SIZE) {
         PyErr_SetString(PyExc_SystemError, "Row overflow");
         goto out;
@@ -1325,13 +1327,9 @@ Column_pack_variable_elements_address(Column *self, void *dest,
         PyErr_SetString(PyExc_SystemError, "too many elements");
         goto out;
     }
-#if WORDS_BIGENDIAN
-    memcpy(v, &off, sizeof(off)); 
-    memcpy(v + sizeof(off), &n, sizeof(n)); 
-#else
-    byteswap_copy(v, &off, sizeof(off)); 
-    byteswap_copy(v + sizeof(off), &n, sizeof(n)); 
-#endif
+    pack_uint((uint64_t) offset, v, address_size);
+    v += address_size;
+    pack_uint((uint64_t) num_elements, v, var_size);
     ret = 0;
 out:
     return ret;
@@ -1347,23 +1345,26 @@ Column_unpack_variable_elements_address(Column *self, void *src,
 {
     int ret = -1;
     void *v = src;
-    uint16_t off = 0;
-    uint8_t n = 0;
-#if WORDS_BIGENDIAN
-    memcpy(&off, v, sizeof(off)); 
-    memcpy(&n, v + sizeof(off), sizeof(n)); 
-#else
-    byteswap_copy(&off, v, sizeof(off)); 
-    byteswap_copy(&n, v + sizeof(off), sizeof(n)); 
-#endif
-    /* this is actually always false now, but will be important in future */
-    if (((u_int32_t) off) >= MAX_ROW_SIZE) {
[…]

### 2013-08-20 test/lowlevel.py

@@ -1213,7 +1213,7 @@ class TestTableInitialisation(TestTable):
         cols = [get_uint_column(8, 1) for k in range(MAX_ROW_SIZE // 8)]
         # This should be fine
         t = _wormtable.Table(f1, f2, cols, 0)
-        cols += [get_uint_column(1, 1)]
+        cols += [get_uint_column(8, 1)]
         self.assertRaises(WormtableError, _wormtable.Table, f1, f2, cols, 0)
 
     def test_open(self):

### 2013-08-20 wormtable.py

@@ -38,9 +38,9 @@ from xml.etree import ElementTree
 
 import _wormtable
 
-__version__ = '0.1.0b2'
-TABLE_METADATA_VERSION = "0.1"
-INDEX_METADATA_VERSION = "0.1.2"
+__version__ = '0.1.0b3'
+TABLE_METADATA_VERSION = "0.2"
+INDEX_METADATA_VERSION = "0.2"
 
 DEFAULT_CACHE_SIZE = 16 * 2**20 # 16M 
 DEFAULT_CACHE_SIZE_STR = "16M" 

### 2013-08-20 _wormtablemodule.c
Fixed long list overflow crash.

@@ -1135,6 +1135,8 @@ Column_parse_string_sequence(Column *self, char *s)
 {
     int ret = -1;
     int j, num_elements, delimiter;
+    int max_num_elements = self->num_elements == WT_VAR_1 ? MAX_NUM_ELEMENTS
+            : self->num_elements;
     self->num_buffered_elements = 0;
     if (self->num_elements == 1) {
         self->input_elements[0] = s;
@@ -1147,12 +1149,16 @@ Column_parse_string_sequence(Column *self, char *s)
             Column_encoded_elements_parse_error(self, "Empty value", s); 
             goto out;
         }
-        /* TODO this needs lots of error checking! */
         while (s[j] != '\0') {
             if (s[j] == ',' || s[j] == ';') {
                 delimiter = j; 
             }
             if (j == delimiter + 1) {
+                if (num_elements >= max_num_elements) {
+                    Column_encoded_elements_parse_error(self, 
+                            "incorrect number of elements", s); 
+                    goto out;
+                }
                 /* this is the start of a new element */
                 self->input_elements[num_elements] = &s[j];
                 num_elements++;

### 2013-08-20 test/lowlevel.py

@@ -230,8 +230,6 @@ class TestElementParsers(TestDatabase):
                 self.assertRaises(TypeError, rb.insert_elements, c.position, v)
         
 
-   
-
 class TestListParsers(TestDatabase):
     """
     Test the list parsers to make sure that malformed lists are correctly
@@ -253,6 +251,11 @@ class TestListParsers(TestDatabase):
             self._float_columns[j] = k
             cols.append(get_float_column(4, j))
             k += 1
+        self._variable_cols = [k]
+        cols.append(get_int_column(1, _wormtable.WT_VAR_1))
+        k += 1
+        self._variable_cols.append(k)
+        cols.append(get_float_column(4, _wormtable.WT_VAR_1))
         return cols
      
     def test_malformed_python_lists(self):
@@ -262,6 +265,26 @@ class TestListParsers(TestDatabase):
         for s in [[1], [1, 2, 3], (1, 2, 3), range(40)]:
             self.assertRaises(ValueError, rb.insert_elements, f2, s)
             self.assertRaises(ValueError, rb.insert_elements, i2, s)
+    
+    def test_long_lists(self):
+        rb = self._row_buffer
+        i2 = self._int_columns[2]
+        f2 = self._int_columns[2]
+        for j in [0, 1, 3, 4, 50]:
+            s = [0 for k in range(j)]
+            self.assertRaises(ValueError, rb.insert_elements, f2, s)
+            self.assertRaises(ValueError, rb.insert_elements, i2, s)
+            ss = ",".join([str(u) for u in s])
+            sse = ss.encode()
+            self.assertRaises(ValueError, rb.insert_encoded_elements, f2, sse)
+            self.assertRaises(ValueError, rb.insert_encoded_elements, i2, sse)
+        for k in self._variable_cols:
+            for l in range(1, 50):
+                s = [0 for j in range(_wormtable.MAX_NUM_ELEMENTS + l)]
+                self.assertRaises(ValueError, rb.insert_elements, k, s)
+                ss = ",".join([str(u) for u in s])
+                sse = ss.encode()
+                self.assertRaises(ValueError, rb.insert_encoded_elements, k, sse)
 
 
 class TestDatabaseLimits(TestDatabase):

### 2013-08-21 docs/paper/paper.tex
modified slightly the background and abstract sections

@@
 \paragraph*{Background:} 
-Biological research generates large amounts of tabular data.
-These tables are produced by 
-programs and exported to plain text files, which are often very large. 
+Biological research, including genomics generates huge amounts of data.
+The data produced by 
+programs are commonly stored as tables in plain text files and are often very large. 
 Researchers must process these files 
@@
 can process tables in a very high performance manner, allowing them
-much greater freedom to explore their data. 
-Since it is no longer necessary to parse complex files,
+much greater freedom to explore their data. Our comparisons with some 
+alternative software demonstrates a 100-fold increase in speed. Additionally, 
+because it is no longer necessary to parse complex files,
 data analysis code is also much simpler and less error prone. 
@@
 text files can be quite compact, and specialised indexing methods 
-are available to quickly retrieve rows intersecting with a 
+are available to retrieve specific rows, for example rows which intersect with a 
 given genomic interval~\cite{li11}.
@@
 must first parse the encoded information into native 
-machine values. This is an expensive process, and 
-compression (if it is used) adds another substantial overhead. 
+machine values. This is a computationally expensive process, and 
+compression (if it is used) adds substantial overhead. 
 As a result,  simple calculations over a large dataset may take 
@@
 Another problem with tables stored as text files is that it is difficult 
-to index the information within particular columns. 
+to index the information in a particular columns. 
 This means that many operations on a table require a complete scan 
@@
 relational database. Databases store values in binary form so that parsing is 
-not required, and support efficient retrieval and indexing. There 
-are many problems with this approach, however. Relational databases are 
-extremely complex systems, each supporting different features and SQL 
+not required, and support efficient retrieval and indexing. However, there 
+are many problems with this approach. Relational databases are 
+complex systems, each supporting different features and SQL 
 dialects. It is not a straightforward task to design a schema for a 
@@
 and so storage and user permissions must be carefully managed. All of 
-these aspects require significant expertise.
+these aspects require significant expertise which is why so many researchers 
+and programs continue to use tabular data formats.
 
-A relational database server, however, is far more than we require in the 
+A relational database server is far more than we require in the 
 majority of cases. Data files are usually written once and 
@@
 from existing formats via command line tools. 
-Wormtable provides a simple Python API to access rows
+Wormtable provides a simple, user friendly Python API to access rows
 in the table, facilitating efficient and convenient access to 
@@
 together, and the values are pointers to the original row. Indexes make
-many operations much more efficient, since we can directly seek to the 
+many operations much more efficient, because we can go directly to the 
 rows we are interested in without performing a full table scan. They also 
 provide a very efficient means of calculating a histogram for a given 
-set of columns. Wormtable also supports binned indexes, in which a range of 
+set of columns. Moreover, wormtable supports binned indexes, in which a range of 
 values is mapped to a single index key. This is very useful for floating 
 point columns, where it may not be necessary to distinguish between
-values that differ by small amounts.
+very similar values.
 
@@
 Files within a table's home directory are not intended to be 
-manipulated by users, and the \texttt{wtadmin} program is provided 
+manipulated directly by users, and the \texttt{wtadmin} program is provided 
 to perform administration tasks such as adding and removing 

### 2013-08-21 docs/paper/paper.tex
modified up to seek performance section

@@
 using an easy to learn interface. 
-The \texttt{Table} class provides the main interface,
+In Python, the \texttt{Table} class provides the main interface,
 and implements the standard Python sequence protocols. Efficient 
@@
 of interest are then retrieved, leading to considerable
-time savings. The \texttt{cursor} function takes two additional 
+time savings. The \texttt{cursor} function can take two additional 
 arguments, \texttt{start} and \texttt{stop}, which specify 
-rows of interest. This allows us to efficiently seek to an 
+the rows of interest. This allows us to efficiently seek to an 
 arbitrary location in the table and to read a given number 
@@
 examples. These are not intended to be definitive benchmarks
-and are simply concrete examples to demonstrate the type of improvements 
+but are simple examples to demonstrate the type of improvements 
 that can be expected by using wormtable over existing methods.
@@
 The Variant Call Format (VCF) encodes information about variant sites in a genome 
-as tab-delimited rows in a text file~\cite{da+11}.  
+as tab-delimited rows in a text file~\cite{da+11}. The VCF format is one of 
+the most commonly used formats to store genomic data from next generation sequencing.
 To illustrate the advantages of wormtable when performing calculations over 

### 2013-08-21 docs/paper/paper.tex
modified up to discussion

@@
 in cache. In this case, the elapsed time was 56 seconds for Tabix and around 
-2 seconds for wormtable. Thus, compression is a double-edged sword: 
-it does lead to much smaller files, which reduces the number of random seeks 
-required.  On the other hand, however, decompression is expensive and must 
-be repeated each time a block is accessed unless some sophisticated
-caching mechanism is devised.
+2 seconds for wormtable. Thus, the use of compression is a trade-off: 
+it reduces file size, which reduces the number of random seeks 
+required but decompression is expensive and must 
+be repeated each time a block is accessed.
 

### 2013-08-21 docs/paper/paper.tex
RN is done his edits

@@
 
-Wormtable is currently limited to supporting Python only, and 
+Wormtable is currently limited to supporting Python, and 
 another important aspect of future development is to 
@@
 %
-The volume of data being produced in bioinformatics is constantly 
-increasing, but the tools available to end users to 
+The volume of data being produced in biological research is growing 
+rapidly, but the tools available to end users to 
 process data are still mostly based on parsing plain text. This 
@@
 Firstly, and most obviously, a researcher's productivity is inevitably
-reduced by being forced to wait for several hours for the result of a 
+constrained while waiting several hours for the result of a 
 simple calculation. Without flexible indexing, working with 
@@
 extra storage and maintenance.
-There are also some less obvious consequences:
-code quality is reduced, since testing over the entire dataset is 
-infeasible, and it is much less likely that the effects of 
-changing arbitrary quality thresholds (for example) will be 
+Additionally, code quality is reduced, since testing over 
+the entire dataset is 
+infeasible and it is less likely that the effects of 
+changing arbitrary analysis parameters will be 
 systematically examined.
@@
 is to use a relational database, which provide sophisticated 
-data management techniques. Relational databases are unsuitable 
-for storing static datasets, however, as they are complex to 
+data management techniques. However, relational databases are unsuitable 
+for storing static datasets as they are complex to 
 use and incur many unnecessary overheads.
@@
 technologies (packed binary storage of values; random access 
-to rows; general purpose indexing) without their complexities 
+to rows; general purpose indexing) without additional complexities 
 and overheads. Wormtable is also more flexible than 
@@
 is far more adaptable than a relational database.
-Any widely used database system today is a hugely complex 
-piece of software, and adding necessary features would be 
+All widely used database systems are complex 
+and adding the necessary features would be 
 very difficult. Adding new features to wormtable, on the 
-other hand, is straightforward since it is far simpler and 
+other hand, is straightforward because it is far simpler and 
 does not need to be compatible with the relational model
@@
 interest. To access regions of the genome efficiently
-the user simply needs to create an index on the CHROM and POS columns
-using the  \texttt{wtadmin add} command, and can then efficiently
-seek to arbitrary locations. Wormtable is not 
+the user simply needs to create an index on the chromosome 
+(CHROM) and position (POS) columns
+using the  \texttt{wtadmin add} command. Wormtable is not 
 limited to VCF, but can store 
-any form of fixed tabular data. We also provide a conversion 
-tool for the GTF format, and several others are planned.
+any form of fixed tabular data. We provide a conversion 
+tool for the GTF format, and several others are planned or 
+could be contributed by users.
 

### 2013-08-21 docs/paper/paper.tex
Updated abstract and made another editing pass.

@@
 \paragraph*{Background:} 
-Biological research, including genomics generates huge amounts of data.
-The data produced by 
-programs are commonly stored as tables in plain text files and are often very large. 
+Genomics generates huge amounts of tabular data which is 
+usually stored in plain text files.
+%Biological research, including genomics generates huge amounts of data.
+%The data produced by 
+%programs are commonly stored as tables in plain text files and are often very large. 
 Researchers must process these files 
@@
 formats can be easily converted to wormtable format, and we 
-provide conversion utilities for the VCF and GTF format.
+provide conversion utilities for the VCF and GTF formats.
 
-\paragraph*{Conclusions:} Using wormtable's simple API users 
-can process tables in a very high performance manner, allowing them
-much greater freedom to explore their data. Our comparisons with some 
-alternative software demonstrates a 100-fold increase in speed. Additionally, 
-because it is no longer necessary to parse complex files,
-data analysis code is also much simpler and less error prone. 
+\paragraph*{Conclusions:} 
+Wormtable's simple API allows users to process large 
+tables orders of magnitude more quickly than is possible 
+when parsing text. Furthermore, the indexing facilities 
+provide efficient access to subsets of the data along with 
+providing useful methods of summarising columns. Since 
+third-party libraries or custom code are no longer needed 
+to parse complex plain text formats, analysis code 
+can also be substantially simpler as well as being 
+uniform across different data formats. These benefits of 
+reduced code complexity and greatly increased performance
+allow users much greater freedom to explore their data.
 \end{abstract}
@@
 it must be \emph{processed}. This is the major flaw in 
-using text files as a data format: before we can perform calculations, we 
+using text files as a data format; before we can perform calculations, we 
 must first parse the encoded information into native 
@@
 Another problem with tables stored as text files is that it is difficult 
-to index the information in a particular columns. 
+to index the information in particular columns. 
 This means that many operations on a table require a complete scan 
@@
 dialects. It is not a straightforward task to design a schema for a 
-particular data set, particularly not if portability across different 
+particular dataset, particularly not if portability across different 
 databases is required. Similarly, accessing data requires a knowledge
-of SQL. In the most common case, a database server must also be maintained
+of SQL. In the most common case, a database server must be maintained
 and so storage and user permissions must be carefully managed. All of 
 these aspects require significant expertise which is why so many researchers 
-and programs continue to use tabular data formats.
+and programs continue to use textual data formats.
 
@@
 in a relational database with its sophisticated concurrency control
-is entirely unnecessary. Centralised storage of data sets
+is entirely unnecessary. Centralised storage of datasets
 creates an unnecessary administration burden, as does an extra
@@
 provide powerful data management technologies, they are complex
-to use and maintain and are therefore rarely used for 
+to use and maintain and are rarely used for 
 static datasets.
@@
 Data is stored in rows using a compact binary format.
-Each row consists of a fixed number of columns, and each column 
-holds either a fixed or variable number of elements of a fixed type and size.
+Each row consists of a number of columns, and each column 
+holds elements of a specific type and size.
 Each column is allocated a fixed number of bytes within the `fixed region' of 
 a row. For fixed
-length columns, the encoded elements are stored entirely within the
+length columns, the encoded elements are stored entirely within this
 fixed region; for variable length columns, the encoded elements are 
@@
 which the keys are the elements from the columns in question concatenated
-together, and the values are pointers to the original row. Indexes make
+together, and the values are pointers to the original rows. Indexes make
 many operations much more efficient, because we can go directly to the 
@@
 of interest are then retrieved, leading to considerable
-time savings. The \texttt{cursor} function can take two additional 
+time savings. The \texttt{cursor} method can take two additional 
 arguments, \texttt{start} and \texttt{stop}, which specify 
@@
 
-The \texttt{Index} class also has a \texttt{cursor} function
+The \texttt{Index} class also has a \texttt{cursor} method 
 with the same signature, but in this case, rows 
@@
 whole genome over 6,146,611 rows.
-Using the included \texttt{vcf2wt} program
+Using the included \texttt{vcf2wt} program,
 the VCF was converted to wormtable format 
@@
 seeks were required 
-to bring the required pages into memory (solid state 
-storage does not suffer from this problem). Once the 
+to bring the required pages into memory. Once the 
 pages were in memory, however, Tabix needed to do much more work to 
@@
 it reduces file size, which reduces the number of random seeks 
-required but decompression is expensive and must 
+required, but decompression is expensive and must 
 be repeated each time a block is accessed.
+The advantages of a smaller file in terms of reducing the
+number of seeks incurred would also be largely negated 
+by using solid state storage, where random seeks do not
+incur such a heavy penalty.
 
@@
 workloads have strong locality of reference~\cite{d05}, 
-and such large and extreme seeks across genomic regions 
+and such large and extreme jumps across genomic regions 
 are unlikely to occur. In this case, wormtable 
@@
 another important aspect of future development is to 
-develop a C library along with bindings for other popular
+create a C library along with bindings for other popular
 languages such as Perl, R and PHP. 

### 2013-08-22 docs/paper/paper.tex
Replaced binary row format paragraph

@@
 Data is stored in rows using a compact binary format.
-Each row consists of a number of columns, and each column 
-holds elements of a specific type and size.
+Each row consists of a defined number of columns, and each column 
+holds elements of a given type and size.
 Each column is allocated a fixed number of bytes within the `fixed region' of 
-a row. For fixed
-length columns, the encoded elements are stored entirely within this
-fixed region; for variable length columns, the encoded elements are 
+a row. For columns with fixed length, the encoded elements are stored entirely within the
+fixed region; for columns of variable length, the encoded elements are 
 stored in the `variable region' of the row, and the location and number 

### 2013-08-25 _wormtablemodule.c
Changed index separator to Missing Value sentinel.
This change adds the missing value to the end of the contribution
of each column when generating index keys. This allows us to
reconstruct index keys without reading the source rows.

@@ -138,7 +138,7 @@ typedef struct {
     PyObject_HEAD
     Index *index;
     DBC *cursor;
-} DistinctValueIterator;
+} IndexKeyIterator;
 
 
 static void 
@@ -1415,6 +1415,61 @@ out:
     return ret;
 }
 
+
+/*
+ * Extract values from the specified key buffer starting at the specified 
+ * offset. Return the number of elements read, or < 0 in case of an error. 
+ */
+static int 
+Column_extract_key(Column *self, void *key_buffer, uint32_t offset, 
+        uint32_t key_size)
+{
+    int ret = -1;
+    unsigned char *v;
+    uint32_t j, k, s;
+    uint32_t element_size = self->element_size;
+    uint32_t num_elements = self->num_elements;
+    int not_done;
+    if (num_elements == WT_VAR_1) {
+        /* count the number of elements until we hit the sentinel */
+        num_elements = 0;
+        not_done = 1;
+        j = offset;
+        v = key_buffer;
+        while (not_done) {
+            s = 0;
+            k = j + element_size;
+            while (j < k) {
+                if (j >= key_size) {
+                    PyErr_SetString(PyExc_SystemError, "Key buffer overflow");
+                    goto out; 
+                }
+                s += v[j];
+                j++;
+            }
+            if (s == 0) {
+                not_done = 0;
+            } else {
+                num_elements++;
+            }
+        }
+        v = key_buffer;
+    } 
+    if (offset + num_elements * element_size >= key_size) {
+        PyErr_SetString(PyExc_SystemError, "Key offset too long");
+        goto out; 
+    }
+    v = key_buffer + offset;
+    self->num_buffered_elements = num_elements;
+    ret = self->unpack_elements(self, v);
+    if (ret < 0) {
+        goto out;
+    }
+    ret = num_elements;
+out:
+    return ret;
+}
+
 /*
  * Extracts elements from the specified row and inserts them into the 
  * element buffer. 
@@ -2568,6 +2623,8 @@ Index_init(Index *self, PyObject *args, PyObject *kwds)
     PyObject *db_filename = NULL;
     PyObject *columns = NULL;
     Table *table = NULL;
+    uint32_t n;
+
     self->db = NULL;
     self->table = NULL;
     self->db_filename = NULL;
[…]

### 2013-08-25 test/highlevel.py

@@ -477,6 +477,7 @@ class BinnedIndexIntegrityTest(WormtableTest):
             w = i.bin_widths()[0]
             t = [r[c.get_position()] for r in self._table] 
             d = histogram(t, w)
+            d2 = dict(i.counter().items())
             for k, v in i.counter().items():
                 self.assertTrue(k in d)
                 self.assertEqual(d[k], v)

### 2013-08-25 test/lowlevel.py

@@ -972,7 +972,7 @@ class TestIndexIntegrity(object):
             index = self._indexes[k]
             index.open(WT_READ)
             u = sorted(distinct_values.keys())
-            dvi = _wormtable.DistinctValueIterator(index)
+            dvi = _wormtable.IndexKeyIterator(index)
             v = list(t[0] for t in dvi)
             self.assertEqual(u, v)
             for key, count in distinct_values.items():
@@ -1481,7 +1481,7 @@ class TestIndexInitialisation(TestIndex):
         self._table.open(WT_READ)
         for j in range(n):
             r = self._table.get_row(j)
-        g = _wormtable.DistinctValueIterator
+        g = _wormtable.IndexKeyIterator
         self.assertRaises(TypeError, g) 
         self.assertRaises(TypeError, g, None) 
         self.assertRaises(TypeError, g, self._table) 
@@ -1494,7 +1494,7 @@ class TestIndexInitialisation(TestIndex):
         index.close()
         self.assertRaises(WormtableError, g, index) 
         index.open(WT_READ)
-        dvi = _wormtable.DistinctValueIterator(index)
+        dvi = _wormtable.IndexKeyIterator(index)
         index.close()
         def f():
             for x in dvi:

### 2013-08-25 wormtable.py

@@ -1094,7 +1094,7 @@ class Index(Database):
         order.
         """
         self.verify_open(WT_READ)
-        dvi = _wormtable.DistinctValueIterator(self.get_ll_object())
+        dvi = _wormtable.IndexKeyIterator(self.get_ll_object())
         for k in dvi:
             yield self.ll_to_key(k)
 
@@ -1213,13 +1213,13 @@ class IndexCounter(collections.Mapping):
         return self.__index.get_ll_object().get_num_rows(k) 
    
     def __iter__(self):
-        dvi = _wormtable.DistinctValueIterator(self.__index.get_ll_object())
+        dvi = _wormtable.IndexKeyIterator(self.__index.get_ll_object())
         for v in dvi:
             yield self.__index.ll_to_key(v)
 
     def __len__(self):
         n = 0
-        dvi = _wormtable.DistinctValueIterator(self.__index.get_ll_object())
+        dvi = _wormtable.IndexKeyIterator(self.__index.get_ll_object())
         for v in dvi:
             n += 1
         return n

### 2013-08-25 _wormtablemodule.c
Tidied code and bumped index metadata version.

@@ -2929,6 +2929,7 @@ Index_key_to_python(Index *self, void *key_buffer, uint32_t key_size)
             Py_DECREF(t);
             goto out;
         }
+        /* skip the sentinel */
         offset += (n + 1) * col->element_size;
         value = Column_get_python_elements(col); 
         if (value == NULL) {
@@ -4115,7 +4116,6 @@ IndexKeyIterator_next(IndexKeyIterator *self)
     if (Index_check_read_mode(self->index) != 0) {
         goto out;
     }
-    //Index_init_dbts(self->index, &primary_key, &primary_data);
     memset(&key, 0, sizeof(DBT));
     memset(&data, 0, sizeof(DBT));
     if (self->cursor == NULL) {
@@ -4129,15 +4129,6 @@ IndexKeyIterator_next(IndexKeyIterator *self)
     }
     db_ret = self->cursor->get(self->cursor, &key, &data, DB_NEXT_NODUP);
     if (db_ret == 0) {
-        /*
-        if (Table_retrieve_row(self->index->table) != 0) {
-            goto out;
-        }
-        ret = Index_row_to_python(self->index, self->index->table->row_buffer);
-        if (ret == NULL) {
-            goto out;
-        }
-        */
         ret = Index_key_to_python(self->index, key.data, key.size);
         if (ret == NULL) {
             goto out;

### 2013-08-25 wormtable.py

@@ -40,7 +40,7 @@ import _wormtable
 
 __version__ = '0.1.0b3'
 TABLE_METADATA_VERSION = "0.2"
-INDEX_METADATA_VERSION = "0.2"
+INDEX_METADATA_VERSION = "0.3"
 
 DEFAULT_CACHE_SIZE = 16 * 2**20 # 16M 
 DEFAULT_CACHE_SIZE_STR = "16M" 

### 2013-08-25 docs/paper/paper.tex

@@
 \paragraph*{Background:} 
-Genomics generates huge amounts of tabular data which is 
-usually stored in plain text files.
-%Biological research, including genomics generates huge amounts of data.
-%The data produced by 
-%programs are commonly stored as tables in plain text files and are often very large. 
-Researchers must process these files 
-line-by-line, parsing values encoded as text into native types so that 
-calculations may be done. This is an error prone and time consuming process.
+Modern biological science generates a vast amount of data, the analysis 
+of which presents a major challenge to researchers. 
+Data are commonly represented in tables stored as plain text files and 
+require line-by-line parsing for analysis, which is time consuming 
+and error prone.
 Furthermore, there is no simple means of indexing these files so that 

### 2013-08-26 test/highlevel.py
Fixed key translation for multivalue columns.

@@ -39,9 +39,13 @@ def histogram(l, width):
     """
     d = {}
     for u in l:
-        v = None
-        if u is not None:
-            v = u - math.fmod(u, width)
+        if isinstance(u, tuple):
+            v = [uu - math.fmod(uu, width) for uu in u]
+            v = tuple(v)
+        else:
+            v = None
+            if u is not None:
+                v = u - math.fmod(u, width)
         if v in d:
             d[v] += 1
         else:
@@ -72,6 +76,7 @@ class WormtableTest(unittest.TestCase):
         t.add_int_column("int")
         t.add_float_column("float", size=4)
         t.add_char_column("char", num_elements=3)
+        t.add_uint_column("uintv", num_elements=wt.WT_VAR_1)
         t.open("w")
         def g():
             return random.random() < 0.25
@@ -80,7 +85,9 @@ class WormtableTest(unittest.TestCase):
             i = None if g() else random.randint(0, max_value)
             f = None if g() else random.uniform(0, max_value)
             c = None if g() else str(random.randint(0, max_value)).encode()
-            t.append([None, u, i, f, c]) 
+            n = random.randint(1, 5)
+            v = [0 for j in range(n)]
+            t.append([None, u, i, f, c, v]) 
             if random.random() < 0.33:
                 t.append([])
         t.close()
@@ -416,20 +423,13 @@ class IndexIntegrityTest(WormtableTest):
                 stop_index = key_rows.index(stop_key)
                 l = [r for k, r in t[start_index:stop_index]]
                 c = 0
+                if len(cols) == 1:
+                    start_key = start_key[0]
+                    stop_key = stop_key[0]
                 for r1, r2 in zip(i.cursor(read_cols, start_key, stop_key), l):
                     self.assertEqual(r1, r2)
                     c += 1
                 self.assertEqual(c, stop_index - start_index)
-                # If we have a single column index we also support passing the 
-                # values directly
-                if len(cols) == 1:
-                    l = [r for k, r in t[start_index:stop_index]]
-                    c = 0
-                    for r1, r2 in zip(i.cursor(read_cols, start_key[0], stop_key[0]), l):
-                        self.assertEqual(r1, r2)
-                        c += 1
-                    self.assertEqual(c, stop_index - start_index)
-                  
 
             i.close()
             
@@ -457,7 +457,7 @@ class BinnedIndexIntegrityTest(WormtableTest):
                     self._indexes.append(i)
[…]

### 2013-08-26 wormtable.py

@@ -1104,7 +1104,10 @@ class Index(Database):
         Returns the smallest key greater than or equal to the specified 
         prefix.
         """
-        key = self.key_to_ll(k)
+        if len(k) == 0:
+            key = k
+        else:
+            key = self.key_to_ll(k)
         v = self.get_ll_object().get_min(key)
         return self.ll_to_key(v) 
 
@@ -1112,7 +1115,10 @@ class Index(Database):
         """
         Returns the largest index key less than the specified prefix. 
         """
-        key = self.key_to_ll(k)
+        if len(k) == 0:
+            key = k
+        else:
+            key = self.key_to_ll(k)
         v = self.get_ll_object().get_max(key)
         return self.ll_to_key(v) 
 
@@ -1171,15 +1177,15 @@ class Index(Database):
         for use in the low-level API.
         """
         cols = self.__key_columns
-        u = v
-        if not isinstance(v, tuple):
-            u = v,
-        n = len(u)
-        l = [None for j in range(n)]
-        for j in range(n):
-            l[j] = u[j]
-            if isinstance(l[j], str):
-                l[j] = l[j].encode()
+        if len(cols) == 1:
+            l = [v]
+        else:
+            n = len(v)
+            l = [None for j in range(n)]
+            for j in range(n):
+                l[j] = v[j]
+                if isinstance(l[j], str):
+                    l[j] = l[j].encode()
         return tuple(l)
     
     def ll_to_key(self, v):
@@ -1190,9 +1196,7 @@ class Index(Database):
         ret = v
         cols = self.__key_columns
         if len(cols) == 1:
-            c = cols[0]
-            if c.get_num_elements() == 1 or c.get_type() == WT_CHAR:
-                ret = v[0]
+            ret = v[0]
         return ret
 
 
@@ -1206,10 +1210,7 @@ class IndexCounter(collections.Mapping):
         self.__index = index
     
     def __getitem__(self, key):
-        if isinstance(key, tuple):
-            k = self.__index.key_to_ll(key)
-        else:
-            k = self.__index.key_to_ll((key,))
+        k = self.__index.key_to_ll(key)
         return self.__index.get_ll_object().get_num_rows(k) 
    
     def __iter__(self):

### 2013-08-26 test/highlevel.py
Added extra tests for multivalue columns.

@@ -484,6 +484,65 @@ class BinnedIndexIntegrityTest(WormtableTest):
             i.close()
 
 
+class MultivalueColumnTest(IndexIntegrityTest):
+    """
+    Tests specific to the properties of multivalue columns.
+    """
+
+    def setUp(self):
+        super(MultivalueColumnTest, self).setUp()
+        # Add some indexes with specific properties.
+        index_cols = [["var_uint"], ["var_uint", "var_int"], 
+                ["var_uint", "var_int", "var_char"]]
+        index_cols = [[self._table.get_column(c) for c in cols]
+                for cols in index_cols]
+        for cols in index_cols: 
+            name = "multi" + "+".join(c.get_name() for c in cols) 
+            i = wt.Index(self._table, name) 
+            for c in cols:
+                i.add_key_column(c)
+            i.open("w")
+            i.build()
+            i.close()
+            self._indexes.append(i)
+
+    def make_random_table(self):
+        """
+        Make a small random table with small random values.
+        """
+        n = 200
+        max_value = 2 
+        max_len = 5
+        self._table = wt.Table(self._homedir)
+        t = self._table
+        t.add_id_column(2)
+        t.add_uint_column("fixed_uint", num_elements=max_len)
+        t.add_uint_column("var_uint", num_elements=wt.WT_VAR_1)
+        t.add_int_column("fixed_int", num_elements=max_len)
+        t.add_int_column("var_int", num_elements=wt.WT_VAR_1)
+        t.add_char_column("fixed_char", num_elements=max_len)
+        t.add_char_column("var_char", num_elements=wt.WT_VAR_1)
+        t.open("w")
+        def fixed():
+            return [random.randint(0, max_value) for j in range(max_len)]
+        def var():
+            n = random.randint(0, max_len)
+            return [random.randint(0, max_value) for j in range(n)]
+        for j in range(n):
+            fu = fixed()
+            vu = var()
+            fi = fixed() 
+            vi = var()
+            fc = "".join(str(u) for u in fixed())
+            vc = "".join(str(u) for u in var())
+            t.append([None, fu, vu, fi, vi, fc.encode(), vc.encode()])
+            if random.random() < 0.33:
+                t.append([])
+        t.close()
+        t.open("r")
+
+
+
 class StringIndexIntegrityTest(WormtableTest):
     """
     Tests the integrity of indexes over variable length length 
@@ -860,5 +919,6 @@ class FloatTest(WormtableTest):
[…]

### 2013-08-26 _wormtablemodule.c
First pass implemenation of index key sentinel.

@@ -1455,7 +1455,7 @@ Column_extract_key(Column *self, void *key_buffer, uint32_t offset,
         }
         v = key_buffer;
     } 
-    if (offset + num_elements * element_size >= key_size) {
+    if (offset + num_elements * element_size > key_size) {
         PyErr_SetString(PyExc_SystemError, "Key offset too long");
         goto out; 
     }
@@ -2814,11 +2814,13 @@ Index_fill_key(Index *self, void *row, DBT *skey)
         }
         v += len;
         skey->size += len;
-        /* insert the separator between columns */
-        for (k = 0; k < col->element_size; k++) {
-            *v = 0;
-            v++;
-            skey->size++;
+        if (col->num_elements == WT_VAR_1) {
+            /* insert the separator between columns */
+            for (k = 0; k < col->element_size; k++) {
+                *v = 0;
+                v++;
+                skey->size++;
+            }
         }
     }
     ret = 0;
@@ -2872,11 +2874,13 @@ Index_set_key(Index *self, PyObject *args, void *buffer)
         col->pack_elements(col, key_buffer);
         key_buffer += m;
         key_size += m;
-        /* insert the separator between columns */
-        for (k = 0; k < col->element_size; k++) {
-            *key_buffer = 0;
-            key_buffer++;
-            key_size++;
+        if (col->num_elements == WT_VAR_1) {
+            /* insert the separator between columns */
+            for (k = 0; k < col->element_size; k++) {
+                *key_buffer = 0;
+                key_buffer++;
+                key_size++;
+            }
         }
     }
     ret = key_size;
@@ -2929,8 +2933,11 @@ Index_key_to_python(Index *self, void *key_buffer, uint32_t key_size)
             Py_DECREF(t);
             goto out;
         }
-        /* skip the sentinel */
-        offset += (n + 1) * col->element_size;
+        if (col->num_elements == WT_VAR_1) {
+            /* skip the sentinel */
+            n++;
+        }
+        offset += n * col->element_size;
         value = Column_get_python_elements(col); 
         if (value == NULL) {
             Py_DECREF(t);

### 2013-08-26 test/highlevel.py
Added num_rows control to high level tests.

@@ -33,6 +33,9 @@ import itertools
 
 from xml.etree import ElementTree
 
+# module variables used to control the number of tests that we do.
+num_random_test_rows = 10
+
 def histogram(l, width):
     """
     Returns a histogram over the data in v with bin widths width.
@@ -67,7 +70,7 @@ class WormtableTest(unittest.TestCase):
         """
         Make a small random table with small random values.
         """
-        n = 100
+        num_rows = num_random_test_rows 
         max_value = 10
         self._table = wt.Table(self._homedir)
         t = self._table
@@ -80,7 +83,7 @@ class WormtableTest(unittest.TestCase):
         t.open("w")
         def g():
             return random.random() < 0.25
-        for j in range(n):
+        for j in range(num_rows):
             u = None if g() else random.randint(0, max_value)
             i = None if g() else random.randint(0, max_value)
             f = None if g() else random.uniform(0, max_value)
@@ -510,7 +513,7 @@ class MultivalueColumnTest(IndexIntegrityTest):
         """
         Make a small random table with small random values.
         """
-        n = 200
+        num_rows = num_random_test_rows 
         max_value = 2 
         max_len = 5
         self._table = wt.Table(self._homedir)
@@ -528,7 +531,7 @@ class MultivalueColumnTest(IndexIntegrityTest):
         def var():
             n = random.randint(0, max_len)
             return [random.randint(0, max_value) for j in range(n)]
-        for j in range(n):
+        for j in range(num_rows):
             fu = fixed()
             vu = var()
             fi = fixed() 

### 2013-08-26 tests.py

@@ -58,6 +58,7 @@ def main():
     random.seed(int(options.random_seed))
     testloader = unittest.TestLoader()
     test.lowlevel.num_random_test_rows = num_rows 
+    test.highlevel.num_random_test_rows = num_rows 
     if options.name is not None:
         suite = testloader.loadTestsFromName(options.name)
     else:

### 2013-08-26 _wormtablemodule.c
Tidied the implementation of sentinels.

@@ -2677,10 +2677,10 @@ Index_init(Index *self, PyObject *args, PyObject *kwds)
         col = self->table->columns[k];
         n = col->num_elements;
         if (col->num_elements == WT_VAR_1) {
-            n = MAX_NUM_ELEMENTS;
+            /* allow space for the sentinel */
+            n = MAX_NUM_ELEMENTS + 1;
         }
-        /* allow space for separator */
-        self->key_buffer_size += (n + 1) * col->element_size;
+        self->key_buffer_size += n * col->element_size;
     }
     self->key_buffer = PyMem_Malloc(self->key_buffer_size);
     if (self->key_buffer == NULL) {
@@ -2815,7 +2815,7 @@ Index_fill_key(Index *self, void *row, DBT *skey)
         v += len;
         skey->size += len;
         if (col->num_elements == WT_VAR_1) {
-            /* insert the separator between columns */
+            /* insert the sentinel */
             for (k = 0; k < col->element_size; k++) {
                 *v = 0;
                 v++;
@@ -2867,15 +2867,15 @@ Index_set_key(Index *self, PyObject *args, void *buffer)
             goto out;
         }
         m = col->num_buffered_elements * col->element_size;
-        if (key_size + m + 1 > self->key_buffer_size) {
-            PyErr_Format(PyExc_SystemError, "Max key key_size exceeded."); 
+        if (key_size + m > self->key_buffer_size) {
+            PyErr_Format(PyExc_SystemError, "Max key_size exceeded."); 
             goto out;
         }
         col->pack_elements(col, key_buffer);
         key_buffer += m;
         key_size += m;
         if (col->num_elements == WT_VAR_1) {
-            /* insert the separator between columns */
+            /* insert the sentinel */
             for (k = 0; k < col->element_size; k++) {
                 *key_buffer = 0;
                 key_buffer++;

### 2013-08-26 wormtable.py
Added address_size to XML and bumped versions.

@@ -38,8 +38,8 @@ from xml.etree import ElementTree
 
 import _wormtable
 
-__version__ = '0.1.0b3'
-TABLE_METADATA_VERSION = "0.2"
+__version__ = '0.1.0b4'
+TABLE_METADATA_VERSION = "0.3"
 INDEX_METADATA_VERSION = "0.3"
 
 DEFAULT_CACHE_SIZE = 16 * 2**20 # 16M 
@@ -679,6 +679,11 @@ class Table(Database):
         supported_versions = [TABLE_METADATA_VERSION]
         if version not in supported_versions: 
             raise ValueError("Unsupported schema version.")
+        address_size = root.get("address_size")
+        if address_size is None:
+            raise ValueError("invalid xml: schema address_size missing")
+        if address_size != "2":
+            raise ValueError("Unsupported address size.")
         self._parse_schema_xml(root)
     
     def write_schema(self, filename):
@@ -701,6 +706,7 @@ class Table(Database):
         Generates the XML representing the schema for this table.
         """
         schema = ElementTree.Element("schema")
+        schema.set("address_size", "2")
         columns = ElementTree.Element("columns")
         schema.append(columns)
         for c in self.__columns:

### 2013-11-18 test/highlevel.py
Corrected Solaris platform and large tests.

@@ -74,7 +74,7 @@ class WormtableTest(unittest.TestCase):
         max_value = 10
         self._table = wt.Table(self._homedir)
         t = self._table
-        t.add_id_column(2)
+        t.add_id_column(4)
         t.add_uint_column("uint")
         t.add_int_column("int")
         t.add_float_column("float", size=4)
@@ -518,7 +518,7 @@ class MultivalueColumnTest(IndexIntegrityTest):
         max_len = 5
         self._table = wt.Table(self._homedir)
         t = self._table
-        t.add_id_column(2)
+        t.add_id_column(4)
         t.add_uint_column("fixed_uint", num_elements=max_len)
         t.add_uint_column("var_uint", num_elements=wt.WT_VAR_1)
         t.add_int_column("fixed_int", num_elements=max_len)

### 2013-11-18 _wormtablemodule.c
Initial work making all missing values = None

@@ -38,6 +38,8 @@
 #define MAX_NUM_ELEMENTS 254
 #define MAX_ROW_SIZE 65535
 
+#define WT_MISSING_VALUE 1
+
 #define OFFSET_LEN_RECORD_SIZE 10 
 
 #define MODULE_DOC \
@@ -520,18 +522,25 @@ Column_native_to_python_char(Column *self, int index)
  *
  *************************************/
 
+/*
+ * Unpack values starting at the specified pointer, and return the
+ * number of missing_values encountered.
+ */
 static int 
 Column_unpack_elements_uint(Column *self, void *source)
 {
     int j;
-    int ret = -1;
+    int ret = 0;
     char *v = (char *) source;
     uint64_t *elements = (uint64_t *) self->element_buffer;
+    uint64_t missing_value = missing_uint(self->element_size); 
     int size = self->element_size; 
     for (j = 0; j < self->num_buffered_elements; j++) {
         elements[j] = unpack_uint(v + j * size, size); 
+        if (elements[j] == missing_value) {
+            ret += 1;
+        }
     }
-    ret = 0;
     return ret; 
 }
 
@@ -540,14 +549,17 @@ static int
 Column_unpack_elements_int(Column *self, void *source)
 {
     int j;
-    int ret = -1;
+    int ret = 0;
     char *v = (char *) source;
     int64_t *elements = (int64_t *) self->element_buffer;
+    int64_t missing_value = missing_int(self->element_size); 
     int size = self->element_size; 
     for (j = 0; j < self->num_buffered_elements; j++) {
         elements[j] = unpack_int(v + j * size, size); 
+        if (elements[j] == missing_value) {
+            ret += 1;
+        }
     }
-    ret = 0;
     return ret; 
 }
 
@@ -555,14 +567,19 @@ static int
 Column_unpack_elements_float_2(Column *self, void *source)
 {
     int j;
-    int ret = -1;
+    int ret = 0;
     char *v = (char *) source;
     double *elements = (double *) self->element_buffer;
+    union { double value; uint64_t bits; } conv;
+    uint64_t missing_bits = missing_float(self->element_size);
     for (j = 0; j < self->num_buffered_elements; j++) {
         elements[j] = unpack_half(v); 
         v += self->element_size;
+        conv.value = elements[j];
+        if (conv.bits == missing_bits) {
+            ret += 1;
+        }
     }
-    ret = 0;
     return ret; 
 }
 
[…]

### 2013-11-18 test/lowlevel.py

@@ -1109,31 +1109,16 @@ class TestDatabaseCharMultiColumnIndex(TestDatabaseChar, TestMultiColumnIndex):
 
 class TestMissingValues(object):
     def test_missing_values(self):
-        # Insert an empty row    
-        self._row_buffer.commit_row()
-        # Now writing in empty values
-        for j in range(1, self.num_columns):
-            col = self._columns[j]
-            if col.num_elements == 1:
-                v = None
-            elif col.num_elements > 1:
-                v = tuple(None for k in range(col.num_elements))
-            else:
-                v = tuple()
-            self._row_buffer.insert_elements(j, v)
-        self._row_buffer.commit_row()
+        n = 10
+        # Insert empty rows 
+        for j in range(n):
+            self._row_buffer.commit_row()
         self.open_reading()
-        for j in range(2):
+        for j in range(n):
             r = self._database.get_row(j)
             for k in range(1, len(self._columns)):
-                c = self._columns[k]
-                if c.num_elements == _wormtable.WT_VAR_1:
-                    self.assertEqual(r[k], tuple())
-                elif c.num_elements < 2:
-                    self.assertEqual(r[k], None)
-                else:
-                    v = [None for j in range(c.num_elements)]
-                    self.assertEqual(tuple(v), r[k])
+                self.assertEqual(None, r[k])
+                    
 
 
 class TestIntegerMissingValues(TestMissingValues, TestDatabaseInteger):
@@ -1143,15 +1128,8 @@ class TestFloatMissingValues(TestMissingValues, TestDatabaseFloat):
     pass
     
 class TestCharMissingValues(TestMissingValues, TestDatabaseChar):
+    pass   
    
-   
-    def test_missing_values(self):
-        self._row_buffer.commit_row()
-        self.open_reading()
-        r = self._database.get_row(0)
-        for k in range(1, len(self._columns)):
-            self.assertEqual(b"", r[k])
-
 class TestTable(unittest.TestCase):
     """
     Base class for testing tables.

### 2013-11-18 wormtable.py

@@ -155,8 +155,7 @@ class Column(object):
         """
         Returns the missing value for this column.
 
-        TODO: document
-        """
+        TODO: REMOVE 
         t = self.get_type()
         n = self.get_num_elements()
         ret = None
@@ -168,6 +167,8 @@ class Column(object):
             else:
                 ret = tuple(None for j in range(n))
         return ret
+        """
+        return None
 
     def format_value(self, v):
         """

### 2013-11-19 _wormtablemodule.c
Fixed issues with fixed length strings.

@@ -949,14 +949,6 @@ Column_parse_python_sequence(Column *self, PyObject *elements)
             self->input_elements[j] = v; 
         }
     }
-    if (self->num_elements != WT_VAR_1) {
-        if (num_elements != self->num_elements) {
-            PyErr_Format(PyExc_ValueError, 
-                    "incorrect number of elements for column '%s'",
-                    PyBytes_AsString(self->name));
-            goto out;
-        }
-    }
     self->num_buffered_elements = num_elements;
     ret = 0;
 out:
@@ -1110,8 +1102,6 @@ Column_python_to_native_char(Column *self, PyObject *elements)
 {
     int ret = -1;
     char *s;
-    Py_ssize_t max_length = self->num_elements == WT_VAR_1?
-            MAX_NUM_ELEMENTS: self->num_elements;
     Py_ssize_t length;
     /* Elements must be a single Python bytes object */
     if (!PyBytes_Check(elements)) {
@@ -1126,21 +1116,23 @@ Column_python_to_native_char(Column *self, PyObject *elements)
                 PyBytes_AsString(self->name));
         goto out;
     }
-    if (length > max_length) {
-        PyErr_Format(PyExc_ValueError, 
-                "String too long for column '%s'",
-                PyBytes_AsString(self->name));
-        goto out;
+    if (self->num_elements == WT_VAR_1) {
+        if (length > MAX_NUM_ELEMENTS) {
+            PyErr_Format(PyExc_ValueError, 
+                    "String too long for column '%s'",
+                    PyBytes_AsString(self->name));
+            goto out;
+        }
+    } else {
+        if (length != self->num_elements) {
+            PyErr_Format(PyExc_ValueError, 
+                    "String incorrect length for column '%s'",
+                    PyBytes_AsString(self->name));
+            goto out;
+        }
     }
     memcpy(self->element_buffer, s, length);
     self->num_buffered_elements = length;
-    if (self->num_elements != WT_VAR_1) {
-        /* Pad out the remaining space with zeros */
-        while (self->num_buffered_elements != self->num_elements) {
-            ((char *) self->element_buffer)[self->num_buffered_elements] = '\0';
-            self->num_buffered_elements++;
-        }
-    }
     ret = 0; 
 out:
     return ret;
@@ -1332,17 +1324,24 @@ static int
 Column_string_to_native_char(Column *self, char *string)
 {
     int ret = -1;
-    size_t n = strlen(string);
[…]

### 2013-11-19 test/lowlevel.py

@@ -268,7 +268,7 @@ class TestListParsers(TestDatabase):
     def test_long_lists(self):
         rb = self._row_buffer
         i2 = self._int_columns[2]
-        f2 = self._int_columns[2]
+        f2 = self._float_columns[2]
         for j in [0, 1, 3, 4, 50]:
             s = [0 for k in range(j)]
             self.assertRaises(ValueError, rb.insert_elements, f2, s)
@@ -284,7 +284,7 @@ class TestListParsers(TestDatabase):
                 ss = ",".join([str(u) for u in s])
                 sse = ss.encode()
                 self.assertRaises(ValueError, rb.insert_encoded_elements, k, sse)
-
+    
 
 class TestDatabaseLimits(TestDatabase):
     """
@@ -741,9 +741,9 @@ class TestDatabaseChar(TestDatabase):
 
 class TestDatabaseCharIntegrity(TestDatabaseChar):
     
-    def test_illegal_long_strings(self):
+    def test_illegal_length_strings(self):
         """
-        Test to ensure that long strings are trapped correctly.
+        Test to ensure that long and short strings are trapped correctly.
         """
         rb = self._row_buffer
         for j in range(1, len(self._columns)):
@@ -751,11 +751,16 @@ class TestDatabaseCharIntegrity(TestDatabaseChar):
             n = c.num_elements
             if n == _wormtable.WT_VAR_1:
                 n = _wormtable.MAX_NUM_ELEMENTS
-            for k in [1, 2, 3, 10, 500, 1000]:
-                s = random_string(n + k).encode() 
-                self.assertRaises(ValueError, rb.insert_elements, j, s)
-                self.assertRaises(ValueError, rb.insert_encoded_elements, j, s)
-        
+                for k in [1, 2, 3, 10, 500, 1000]:
+                    s = random_string(n + k).encode() 
+                    self.assertRaises(ValueError, rb.insert_elements, j, s)
+                    self.assertRaises(ValueError, rb.insert_encoded_elements, j, s)
+            else:
+                for k in [0, n - 1, n + 1, n + 2, n + 100]:
+                    s = b"x" * k
+                    self.assertRaises(ValueError, rb.insert_elements, j, s)
+                    self.assertRaises(ValueError, rb.insert_encoded_elements, j, s)
+                
 
     def test_variable_char_retrieval(self):
         rb = self._row_buffer
@@ -784,7 +789,7 @@ class TestDatabaseCharIntegrity(TestDatabaseChar):
             for k in cols: 
[…]

### 2013-11-19 _wormtablemodule.c
Completed changes to index format. Tests passing.

@@ -639,9 +639,6 @@ Column_unpack_elements_char(Column *self, void *source)
     return ret; 
 }
 
-
-
-
 /**************************************
  *
  * Packing native values from the element_buffer to a row.
@@ -966,14 +963,22 @@ Column_python_to_native_uint(Column *self, PyObject *elements)
     uint64_t max_value = max_uint(self->element_size); 
     PyObject *v;
     int j;
-    if (Column_parse_python_sequence(self, elements) < 0) {
-        goto out;
-    }
-    for (j = 0; j < self->num_buffered_elements; j++) {
-        v = (PyObject *) self->input_elements[j];
-        if (v == Py_None) {
-            native[j] = missing_value;
-        } else { 
+    if (elements == Py_None) {
+        if (self->num_elements == WT_VAR_1) {
+            self->num_buffered_elements = 0;
+        } else {
+            for (j = 0; j < self->num_elements; j++) {
+                native[j] = missing_value;
+            }
+            self->num_buffered_elements = self->num_elements;
+        }
+        ret = WT_MISSING_VALUE;
+    } else {
+        if (Column_parse_python_sequence(self, elements) < 0) {
+            goto out;
+        }
+        for (j = 0; j < self->num_buffered_elements; j++) {
+            v = (PyObject *) self->input_elements[j];
             if (!PyNumber_Check(v)) {
                 PyErr_Format(PyExc_TypeError, 
                         "Values for column '%s' must be numeric",
@@ -1010,8 +1015,8 @@ Column_python_to_native_uint(Column *self, PyObject *elements)
                 goto out;
             }
         }
+        ret = 0;
     }
-    ret = 0;
 out:
     return ret;
 }
@@ -1027,14 +1032,22 @@ Column_python_to_native_int(Column *self, PyObject *elements)
     int64_t max_value = max_int(self->element_size); 
     PyObject *v;
     int j;
-    if (Column_parse_python_sequence(self, elements) < 0) {
-        goto out;
-    }
-    for (j = 0; j < self->num_buffered_elements; j++) {
-        v = (PyObject *) self->input_elements[j];
-        if (v == Py_None) {
-            native[j] = missing_value;
-        } else { 
+    if (elements == Py_None) {
+        if (self->num_elements == WT_VAR_1) {
+            self->num_buffered_elements = 0;
+        } else {
+            for (j = 0; j < self->num_elements; j++) {
+                native[j] = missing_value;
+            }
+            self->num_buffered_elements = self->num_elements;
+        }
[…]

### 2013-11-19 test/highlevel.py

@@ -87,8 +87,9 @@ class WormtableTest(unittest.TestCase):
             u = None if g() else random.randint(0, max_value)
             i = None if g() else random.randint(0, max_value)
             f = None if g() else random.uniform(0, max_value)
-            c = None if g() else str(random.randint(0, max_value)).encode()
-            n = random.randint(1, 5)
+            s = str(random.randint(0, max_value)).zfill(3)
+            c = None if g() else s.encode()
+            n = random.randint(0, 5)
             v = [0 for j in range(n)]
             t.append([None, u, i, f, c, v]) 
             if random.random() < 0.33:
@@ -249,7 +250,7 @@ class TableBuildTest(WormtableTest):
         n = len(t.columns())
         for j in range(1, n):
             c = t.get_column(j)
-            self.assertEqual(c.get_missing_value(), r[j])
+            self.assertEqual(None, r[j])
         t.close()
 
  
@@ -311,7 +312,7 @@ class ColumnValue(object):
     def __lt__(self, other):
         v1 = self.__value
         v2 = other.__value
-        missing = self.__column.get_missing_value()
+        missing = None 
         ret = False
         if v1 == missing and v2 != missing:
             ret = True
@@ -380,9 +381,9 @@ class IndexIntegrityTest(WormtableTest):
                 t = [ColumnValue(c, r) for c in cols for r in self._table] 
             else:
                 t = [tuple(ColumnValue(c, r) for c in cols) for r in self._table] 
+            keys = [k for k in i.keys()]
             self.assertEqual(i.min_key(), min(t)) 
             self.assertEqual(i.max_key(), max(t)) 
-            keys = [k for k in i.keys()]
             c1 = {}
             for k in t:
                 if k not in c1:
@@ -415,6 +416,7 @@ class IndexIntegrityTest(WormtableTest):
                 keys = [(k,) for k in i.keys()]
             else:
                 keys = [k for k in i.keys()]
+
             key_rows = [tuple(v.get_value() for v in k) for k, r in t]
             # Now generate some slices
             for j in range(10):

### 2013-11-19 test/lowlevel.py

@@ -730,7 +730,7 @@ class TestDatabaseChar(TestDatabase):
                 c = self._columns[k]
                 n = c.num_elements
                 if n == _wormtable.WT_VAR_1:
-                    n = random.randint(1, _wormtable.MAX_NUM_ELEMENTS)
+                    n = random.randint(0, _wormtable.MAX_NUM_ELEMENTS)
                 row[k] = random_string(n).encode() 
                 if j % 2 == 0:
                     rb.insert_elements(k, row[k]) 
@@ -896,6 +896,7 @@ class TestIndexIntegrity(object):
             max_val = max(s),
             row_iter = _wormtable.IndexRowIterator(index, [j])
             row_iter.set_min(max_val)
+            row_iter.set_max(min_val)
             row_iter.set_max(min_val)
             l = [row[0] for row in row_iter]
             self.assertEqual(len(l), 0)

### 2013-11-19 wormtable.py

@@ -151,38 +151,21 @@ class Column(object):
         """
         return self.__ll_object.num_elements
   
-    def get_missing_value(self):
-        """
-        Returns the missing value for this column.
-
-        TODO: REMOVE 
-        t = self.get_type()
-        n = self.get_num_elements()
-        ret = None
-        if t == WT_CHAR:
-            ret = b''
-        elif n != 1: 
-            if n == WT_VAR_1:
-                ret = tuple()
-            else:
-                ret = tuple(None for j in range(n))
-        return ret
-        """
-        return None
-
     def format_value(self, v):
         """
         Formats the specified value from this column for printing.
         """ 
-        # TODO do something a bit better with missing values
-        n = self.get_num_elements()
-        if self.get_type() == WT_CHAR:
-            s = v.decode()
-        elif n == 1:
-            s = str(v)
+        if v is None:
+            s = "."
         else:
-            s = ",".join(str(u) for u in v) 
-            s = "(" + s + ")"
+            n = self.get_num_elements()
+            if self.get_type() == WT_CHAR:
+                s = v.decode()
+            elif n == 1:
+                s = str(v)
+            else:
+                s = ",".join(str(u) for u in v) 
+                s = "(" + s + ")"
         return s
 
     def get_xml(self):

### 2013-11-19 wormtable.py
Bumped version number + index metadata version.

@@ -38,9 +38,9 @@ from xml.etree import ElementTree
 
 import _wormtable
 
-__version__ = '0.1.0b4'
+__version__ = '0.1.0b5'
 TABLE_METADATA_VERSION = "0.3"
-INDEX_METADATA_VERSION = "0.3"
+INDEX_METADATA_VERSION = "0.4"
 
 DEFAULT_CACHE_SIZE = 16 * 2**20 # 16M 
 DEFAULT_CACHE_SIZE_STR = "16M" 

### 2013-11-19 _wormtablemodule.c
Clarified insertion of missing values + tests.

@@ -2238,9 +2238,9 @@ static PyObject *
 Table_insert_elements(Table* self, PyObject *args)
 {
     PyObject *ret = NULL;
-    Column *column = NULL;
+    Column *col = NULL;
     PyObject *elements = NULL;
-    int m, col_index;
+    int m, col_index, wt_ret;
     if (!PyArg_ParseTuple(args, "iO", &col_index, &elements)) { 
         goto out;
     }
@@ -2248,21 +2248,24 @@ Table_insert_elements(Table* self, PyObject *args)
         goto out;
     }
     if (col_index == 0) {
-        PyErr_Format(WormtableError, "Cannot update ID column."); 
+        PyErr_Format(WormtableError, "Cannot update ID col."); 
         goto out;
     }
     if (Table_check_write_mode(self) != 0) {
         goto out;
     }
-    column = self->columns[col_index];
-    if (column->python_to_native(column, elements) < 0) {
+    col = self->columns[col_index];
+    wt_ret = col->python_to_native(col, elements);
+    if (wt_ret < 0) {
         goto out;   
     }
-    m = Column_update_row(column, self->row_buffer, self->current_row_size); 
-    if (m < 0) {
-        goto out;
+    if (wt_ret != WT_MISSING_VALUE) {
+        m = Column_update_row(col, self->row_buffer, self->current_row_size); 
+        if (m < 0) {
+            goto out;
+        }
+        self->current_row_size += m;
     }
-    self->current_row_size += m;
     Py_INCREF(Py_None);
     ret = Py_None;
 out:

### 2013-11-19 test/lowlevel.py

@@ -897,7 +897,6 @@ class TestIndexIntegrity(object):
             row_iter = _wormtable.IndexRowIterator(index, [j])
             row_iter.set_min(max_val)
             row_iter.set_max(min_val)
-            row_iter.set_max(min_val)
             l = [row[0] for row in row_iter]
             self.assertEqual(len(l), 0)
             # Check if the correct lists are returned. 
@@ -1115,17 +1114,54 @@ class TestDatabaseCharMultiColumnIndex(TestDatabaseChar, TestMultiColumnIndex):
     """
 
 class TestMissingValues(object):
+    """
+    Test that missing and empty values are correctly handled.
+    """
     def test_missing_values(self):
         n = 10
-        # Insert empty rows 
         for j in range(n):
+            # Insert empty row 
+            self._row_buffer.commit_row()
+            # Insert Empty values
+            for k in range(1, len(self._columns)):
+                self._row_buffer.insert_elements(k, None)
+            self._row_buffer.commit_row()
+            # Insert alternating randomly
+            for k in range(1, len(self._columns)):
+                if random.random() < 0.5:
+                    self._row_buffer.insert_elements(k, None)
+            self._row_buffer.commit_row()
+        self.open_reading()
+        for j in range(3 * n):
+            r = self._database.get_row(j)
+            for k in range(1, len(self._columns)):
+                c = self._columns[k]
+                self.assertEqual(None, r[k])
+                    
+    def get_empty_value(self):
+        return tuple() 
+
+    def test_empty_values(self):
+        """
+        Variable length columns support empty values. These are not the same 
+        as missing values!
+        """
+        ev = self.get_empty_value()
+        n = 10
+        for j in range(n):
+            for k, c in enumerate(self._columns):
+                if c.num_elements == _wormtable.WT_VAR_1:
+                    self._row_buffer.insert_elements(k, ev)
             self._row_buffer.commit_row()
         self.open_reading()
         for j in range(n):
             r = self._database.get_row(j)
-            for k in range(1, len(self._columns)):
-                self.assertEqual(None, r[k])
-                    
+            for k, c in enumerate(self._columns):
+                v = None if k > 0 else j 
+                if c.num_elements == _wormtable.WT_VAR_1:
+                    v = ev
+                self.assertEqual(r[k], v)
+
 
 
 class TestIntegerMissingValues(TestMissingValues, TestDatabaseInteger):
[…]

### 2013-11-19 wormtable.py
Changed missing value string to NA.

@@ -156,7 +156,7 @@ class Column(object):
         Formats the specified value from this column for printing.
         """ 
         if v is None:
-            s = "."
+            s = "NA"
         else:
             n = self.get_num_elements()
             if self.get_type() == WT_CHAR:

### 2013-11-19 docs/paper/paper.tex
Addressed comments from editor and reviewers.

@@
 to rows; general purpose indexing) without additional complexities 
-and overheads. Wormtable is also more flexible than 
+and overheads. Wormtable's data model is also less rigid
+than 
 relational databases, supporting, for example, columns 
@@
 All widely used database systems are complex 
-and adding the necessary features would be 
+and adding required features (e.g.\ compression) would be 
 very difficult. Adding new features to wormtable, on the 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Competing interests}
+The authors declare that they have no competing interests.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Authors contributions}
@@
 the example scripts and the manuscript. DH wrote the example scripts and vcf-utils,
-and contributed to vcf2wt, the tutorial and the manuscript.
+and contributed to \texttt{vcf2wt}, the tutorial and the manuscript.
 All authors read and approved the final manuscript.

### 2013-11-20 test/lowlevel.py
Made empty strings much more common to provoke bug

@@ -730,7 +730,9 @@ class TestDatabaseChar(TestDatabase):
                 c = self._columns[k]
                 n = c.num_elements
                 if n == _wormtable.WT_VAR_1:
-                    n = random.randint(0, _wormtable.MAX_NUM_ELEMENTS)
+                    n = 0
+                    if random.random() < 0.75:
+                        n = random.randint(0, _wormtable.MAX_NUM_ELEMENTS)
                 row[k] = random_string(n).encode() 
                 if j % 2 == 0:
                     rb.insert_elements(k, row[k]) 
@@ -1081,7 +1083,6 @@ class TestMultiColumnIndex(object):
             # pick a row at random with a given prefix and find the min 
             # and max with this prefix
             prefix_row = random.choice(rows)
-            #print("row = ", prefix_row)
             ri = _wormtable.IndexRowIterator(index, list(cols))
             l1 = [r for r in ri]
             l2 = sorted(rows)
@@ -1098,6 +1099,7 @@ class TestMultiColumnIndex(object):
                 while l2[j][:k] > prefix:
                     j -= 1
                 self.assertEqual(l2[j], index.get_max(prefix))
+
         self.destroy_indexes()
 
 class TestDatabaseIntegerMultiColumnIndex(TestDatabaseInteger, TestMultiColumnIndex):

### 2013-11-20 test/highlevel.py
Added reliable test case to provoke max bug.

@@ -609,8 +609,40 @@ class StringIndexIntegrityTest(WormtableTest):
         self.assertEqual(i.max_key("A"), (b"A", b"B"))
         self.assertEqual(i.max_key("AA"), (b"AA", b""))
         t.close()
+    
+    def test_empty_prefix(self):
+        """
+        Tests the simplest possible case where we have a 
+        empty string prefix.
+        """
+        t = wt.Table(self._homedir) 
+        t.add_id_column(1)
+        t.add_char_column("s1")
+        t.add_char_column("s2", num_elements=7)
+        t.add_char_column("s3")
+        t.open("w")
+        values = [
+            (b'', b'zGwKwDr', b''),
+            (b'gq', b'wRIFXIp', b'QO'),
+            (b'ugn', b'utgTGMs', b't'),
+            (b'Po', b'pqnlsjI', b'K'),
+            (b'x', b'rwNVozJ', b'')
+        ]
+        for r in values:
+            t.append([None] + list(r))
+        t.close()
+        t.open("r")
+        i = wt.Index(t, "test")
+        i.add_key_column(t.get_column("s1"))
+        i.add_key_column(t.get_column("s2"))
+        i.add_key_column(t.get_column("s3"))
+        i.open("w")
+        i.build()
+        i.close()
+        i.open("r")
+        self.assertEqual(i.max_key(b""), values[0]) 
+        t.close()
  
-
 
 class TableCursorTest(WormtableTest):
     """

### 2013-11-20 test/lowlevel.py

@@ -731,8 +731,11 @@ class TestDatabaseChar(TestDatabase):
                 n = c.num_elements
                 if n == _wormtable.WT_VAR_1:
                     n = 0
-                    if random.random() < 0.75:
+                    u = random.random()
+                    if u < 0.5: 
                         n = random.randint(0, _wormtable.MAX_NUM_ELEMENTS)
+                    elif u < 0.75:
+                        n = _wormtable.MAX_NUM_ELEMENTS
                 row[k] = random_string(n).encode() 
                 if j % 2 == 0:
                     rb.insert_elements(k, row[k]) 

### 2013-11-21 _wormtablemodule.c
Fixed index.get_max and get_min bugs.

@@ -500,16 +500,9 @@ static PyObject *
 Column_native_to_python_char(Column *self, int index)
 {
     PyObject *ret = NULL;
-    int j = self->num_buffered_elements - 1;
+    int j = self->num_buffered_elements;
     char *str = (char *) self->element_buffer;
-    if (self->num_elements != WT_VAR_1) {
-        /* check for shortened fixed-length strings, which will be padded 
-         * with NULLs */
-        while (str[j] == '\0' && j >= 0) {
-            j--;
-        }
-    }
-    ret = PyBytes_FromStringAndSize(str, j + 1); 
+    ret = PyBytes_FromStringAndSize(str, j); 
     if (ret == NULL) {
         PyErr_NoMemory();
     }
@@ -1208,10 +1201,6 @@ Column_parse_string_sequence(Column *self, char *s)
         j = 0;
         num_elements = 0;
         delimiter = -1;
-        if (s[0] == '\0') {
-            Column_encoded_elements_parse_error(self, "Empty value", s); 
-            goto out;
-        }
         while (s[j] != '\0') {
             if (s[j] == ',' || s[j] == ';') {
                 delimiter = j; 
@@ -2857,10 +2846,13 @@ out:
 
 /* 
  * Reads the arguments and sets a key in the specified buffer, returning 
- * its length. 
+ * its length. If prefix_bytes is not null, it is set to 1 at the position
+ * of each variabl length column prefix indicator byte and 0 for all 
+ * other positions.
  */
 static int 
-Index_set_key(Index *self, PyObject *args, void *buffer)
+Index_set_key(Index *self, PyObject *args, void *buffer, 
+        unsigned char *prefix_bytes)
 {
     int ret = -1;
     int j, m, k, wt_ret, overhead;
@@ -2880,6 +2872,9 @@ Index_set_key(Index *self, PyObject *args, void *buffer)
     if (n > self->num_columns) {
         PyErr_Format(PyExc_ValueError, "More key values than columns."); 
         goto out;
+    }
+    if (prefix_bytes != NULL) {
+        memset(prefix_bytes, 0, self->key_buffer_size);
     }
     for (j = 0; j < n; j++) {
         col = self->table->columns[self->columns[j]]; 
@@ -2901,6 +2896,9 @@ Index_set_key(Index *self, PyObject *args, void *buffer)
             /* insert the missing value prefix */
             *key_buffer = wt_ret == WT_MISSING_VALUE? 0 : 1;
             key_buffer++;
[…]

### 2013-11-21 test/highlevel.py

@@ -615,18 +615,25 @@ class StringIndexIntegrityTest(WormtableTest):
         Tests the simplest possible case where we have a 
         empty string prefix.
         """
+        # TODO we need to write a more general test class to exercise this 
+        # code for min/max with prefixes more thoroughly.
         t = wt.Table(self._homedir) 
         t.add_id_column(1)
         t.add_char_column("s1")
-        t.add_char_column("s2", num_elements=7)
+        t.add_char_column("s2")
         t.add_char_column("s3")
         t.open("w")
         values = [
-            (b'', b'zGwKwDr', b''),
-            (b'gq', b'wRIFXIp', b'QO'),
-            (b'ugn', b'utgTGMs', b't'),
-            (b'Po', b'pqnlsjI', b'K'),
-            (b'x', b'rwNVozJ', b'')
+            (None, None, None),
+            (None, None, b""),
+            (None, None, b"z"),
+            (b'', b'', b'a'),
+            (b'', b'', b'b'),
+            (b'', b'', b'c'),
+            (b'', b'a', b'b'),
+            (b'', b'a', b'z'),
+            (b'b', b'b', b'b'),
+            (b'z', b'z', b'z')
         ]
         for r in values:
             t.append([None] + list(r))
@@ -640,7 +647,32 @@ class StringIndexIntegrityTest(WormtableTest):
         i.build()
         i.close()
         i.open("r")
-        self.assertEqual(i.max_key(b""), values[0]) 
+        # Max keys
+        self.assertEqual(i.max_key(None), values[2]) 
+        self.assertEqual(i.max_key(None, None), values[2]) 
+        self.assertEqual(i.max_key(None, None, b""), values[1]) 
+        self.assertEqual(i.max_key(b"", b""), values[5]) 
+        self.assertEqual(i.max_key(b"", b"", b"a"), values[3]) 
+        self.assertEqual(i.max_key(b"", b"", b"b"), values[4]) 
+        self.assertEqual(i.max_key(b"", b"", b"c"), values[5]) 
+        self.assertEqual(i.max_key(b"", b"a"), values[7]) 
+        self.assertEqual(i.max_key(b"", b"a", b"z"), values[7]) 
+        self.assertRaises(KeyError, i.max_key, "x") 
+        self.assertRaises(KeyError, i.max_key, "", "x") 
+        self.assertRaises(KeyError, i.max_key, None, "x") 
[…]

### 2013-11-21 test/lowlevel.py

@@ -410,7 +410,12 @@ class TestDatabaseInteger(TestDatabase):
                 else:
                     n = c.num_elements
                     if n == _wormtable.WT_VAR_1:
-                        n = random.randint(1, _wormtable.MAX_NUM_ELEMENTS)
+                        n = 0
+                        u = random.random()
+                        if u < 0.5: 
+                            n = random.randint(0, _wormtable.MAX_NUM_ELEMENTS)
+                        elif u < 0.75:
+                            n = _wormtable.MAX_NUM_ELEMENTS
                     v = tuple([random.randint(min_v, max_v) for l in range(n)])
                     row[k] = v
                 if j % 2 == 1:
@@ -662,7 +667,12 @@ class TestDatabaseFloat(TestDatabase):
                 else:
                     n = c.num_elements
                     if n == _wormtable.WT_VAR_1:
-                        n = random.randint(1, _wormtable.MAX_NUM_ELEMENTS)
+                        n = 0
+                        u = random.random()
+                        if u < 0.5: 
+                            n = random.randint(0, _wormtable.MAX_NUM_ELEMENTS)
+                        elif u < 0.75:
+                            n = _wormtable.MAX_NUM_ELEMENTS
                     row[k] = tuple([f() for l in range(n)])
                 if j % 2 == 0:
                     rb.insert_elements(k, row[k]) 
@@ -1109,7 +1119,7 @@ class TestDatabaseIntegerMultiColumnIndex(TestDatabaseInteger, TestMultiColumnIn
     """
     Test multicolumn integer indexes 
     """
-class TestDatabaseFloatMultiColumnIndex(TestDatabaseInteger, TestMultiColumnIndex):
+class TestDatabaseFloatMultiColumnIndex(TestDatabaseFloat, TestMultiColumnIndex):
     """
     Test multicolumn integer indexes 
     """

### 2013-11-21 tests.py

@@ -69,7 +69,10 @@ def main():
         suite.addTests(l)
     try:
         for i in range(iterations):
-            unittest.TextTestRunner(verbosity=2).run(suite)
+            r = unittest.TextTestRunner(verbosity=2).run(suite)
+            if len(r.errors) > 0 or len(r.failures) > 0:
+                print("Error detected: exiting!")
+                break
     finally:
         cleanup()
 if __name__ == '__main__':
